### 2017-02-20 abstract.tex
init

\begin{abstract}
  Runtime validation of wireless protocol implementations cannot always employ
  direct instrumentation of the device under test (DUT). The DUT may not
  implement the required instrumentation, or the instrumentation may alter the
  DUT's behavior when enabled. Wireless sniffers can monitor the DUT's
  behavior without instrumentation, but they introduce new validation challenges.
  Losses caused by wireless propagation prevent sniffers
  from perfectly reconstructing the actual DUT packet trace. As a result,
  accurate validation requires distinguishing between
  specification deviations that represent implementation errors and those
  caused by sniffer uncertainty.

  \sloppy{
  We present a new approach enabling sniffer-based validation of wireless
  protocol implementations. Beginning with the original protocol monitor state
  machine, we automatically and completely encode sniffer uncertainty by
  selectively adding non-deterministic transitions. We characterize the
  NP-completeness of the resulting decision problem and provide an exhaustive
  algorithm for searching over all mutated traces. We also present
  practical protocol-oblivious heuristics for searching over the most likely
  mutated traces.  We have implemented our framework and show that it can
  accurately identify implementation errors in the face of uncertainty.
}

  \keywords{
    Runtime Verification \and
    Wireless Protocol \and
    Sniffer \and
  Uncertainty}
\end{abstract}

### 2017-02-20 background.tex
init

\section{Prerequisites and Problem Statement}
\label{sec:background}

\subsection{Packet, Trace and Monitor State Machine}
\label{subsec:basic}

The alphabet of the monitor state machine is the finite set of all valid packets
defined by the protocol, denoted as $\mathbb{P}$.  A packet is a binary string
of a finite number of bits, encoding interesting protocol attributes such as
\texttt{src}, \texttt{dest}, \texttt{type}, \texttt{flags}, and physical layer
information, such as \texttt{channel}, \texttt{modulation}, etc. The input of
the state machine then corresponds to a time-ordered sequence of packets.

\begin{definition}[Packet Trace]
  \sloppy{
  A \textit{packet trace} is a finite sequence of $(timestamp, packet)$ tuple:
  $[(t_1, p_1), (t_2, p_2),\ldots,(t_n, p_n)]$ where $t_i \in \mathbb{Z}^+$ is
  the \textit{discrete} timestamp and $p_i$ is the packet observed at time
  $t_i$. The timestamps are strictly monotonically increasing: $t_i < t_{i+1}$
  for $1 \le i < n$.
}
\end{definition}%

In addition to timestamp monotonicity, we also require that adjacent packets do
not overlap in time, $t_{i+1}-t_i > \text{\texttt{airtime}}(p_i)$ for $1 \le i <
n$, where \texttt{airtime()} calculates the time taken to transmit a packet.
The timestamp represents the observer's local clock ticks, and need not to be
synchronized among devices.

We use \textit{timed automata}~\cite{alur1994theory} to model the expected
behaviors of the DUT. A timed automata is a finite state machine with timing
constraints on the transitions: each transition can optionally start one or more
timers, which can later be used to assert certain events should be seen before
or after the time out event. We refer the readers to~\cite{alur1994theory} for
more details about timed automata.

\begin{definition}[Monitor]
  \sloppy{%
    A protocol monitor state machine $S$ is a 7-tuple
    $\{\Sigma, \mathbb{S}, \mathbb{X}, s_0, C, E, G\}$, where:
  }
  \begin{itemize}
    \item $\Sigma = \mathbb{P}$ is the finite input alphabet.

    \item $\mathbb{S}$ is a non-empty, finite set of states. $s_0 \in
      \mathbb{S}$ is the initial state.

    \item $\mathbb{X}$ is the set of boolean variables. We use $v = \{x
      \leftarrow true/false\ |\ x \in \mathbb{X}\}$ to denote an assignment of
      the variables. Let $\mathbb{V}$ be the set of such values $v$.

    \item $C$ is the set of clock variables. A \textit{clock variable} can be
      reset along any state transitions. At any instant, reading a clock
      variable returns the time elapsed since last time it was reset.

    \item $G$ is the set of guard conditions defined inductively by
      \begin{align*}
        g := true\ |\ c \le T\ |\ c \ge T\ |\ x\ |\ \neg g\ |\ g_1 \land g_2
      \end{align*}%
      where $c \in C$ is a clock variable, $T$ is a constant, and $x$ is a
      variable in $\mathbb{X}$.  A transition can choose not to use guard
      conditions by setting $g$ to be $true$.

    \item $E \subseteq \mathbb{S} \times \mathbb{V} \times \mathbb{S} \times
      \mathbb{V} \times \Sigma \times  G \times \mathscr{P}(C)$ gives the set of
      transitions.\\ $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$
      represents that if the monitor is in state $s_i$ with variable assignments
      $v_i$, given the input tuple $(t, p)$ such that the guard $g$ is
      satisfied, the monitor can transition to a state $s_j$ with variable assignments
      $v_j$, and reset the clocks in $C'$ to 0.
  \end{itemize}
  \label{def:sm}
\end{definition}

A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
the state machine at time $t_i$. The monitor {\it rejects a trace} $Tr$ if there
exists a prefix of $Tr$ such that all states reachable after consuming the
prefix have no valid transitions for the next $(t, p)$ input.

As an example, the monitor state machine illustrated in
Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:

\begin{itemize}
  \item $\Sigma = \{DATA_i, DATA'_i, Ack\ |\ 0 \le i < N\}$.
  \item Clock variables $C = \{c\}$. The only clock variable $c$ is
    used for acknowledgment time out.
  \item $\mathbb{X} = \{i\}$, as a variable with ${\mathit log}(N) + 1$ bits to
    count from $0$ to $N$.  
  \item Guard constraints $G = \{ c \le T_o, c > T_o, T_o < c \le T_m\}$.
    $T_o$ is the acknowledgment time out value, and $T_m >
    T_o$ is the maximum delay allowed before the retransmission packet gets
    sent. $T_o$ can be arbitrary large but not infinity in order to check the
    liveness of the DUT.
\end{itemize}

The monitor state machine defines a \textit{timed language} $L$ which consists
of all valid packet traces that can be observed by the DUT.  We now give the
definition of protocol \textit{compliance} and \textit{violation}.

\begin{definition}[Violation and Compliance]
  Suppose $\mathbb{T}$ is the set of all possible packet traces collected from
  DUT, and $S$ is the state machine specified by the protocol. The DUT
  \textit{violates} the protocol specification if there exists an
  packet trace $Tr \in \mathbb{T}$ such that $S$ rejects $Tr$.
  Otherwise, the DUT is \textit{compliant} with the specification.
\end{definition}

The focus of this paper is to determine whether a \textit{given} sniffer trace
$Tr$ is evidence of a violation.
%
%We acknowledge that determining \textit{compliance} is a more challenging
%problem, as it requires enumerating every possible trace in $\mathbb{T}$, which
%is probably infinite.

\subsection{Mutation Trace}
\label{subsec:mutation}

As shown in the motivation example in Fig.~\ref{fig:sniffer_in_middle}, a
sniffer trace may either miss packets that are present in DUT trace, or contain
extra packets that are missing in DUT trace. Note that in the latter case, those
extra packets must be all sent \textit{to} the DUT. This is because it is
impossible for the sniffer to overhear packets sent from the DUT that were not
actually sent by the DUT.

We formally capture this relationship with the definition of mutation trace.

\begin{definition}[Mutation Trace]
  \label{def:mutation}
  A packet trace $Tr'$ is a mutation of sniffer trace $Tr$ w.r.t a DUT if for
  all $(t, p) \in Tr \setminus Tr'$, $p.dest = DUT$, where $p.dest$ is the
  destination of packet $p$.
\end{definition}

By definition, either $Tr' \supseteq Tr$ (hence $Tr \setminus Tr' = \emptyset$),
or those extra packets in $Tr$ but not in $Tr'$ are all sent to the DUT. Note
that $Tr'$ may contain extra packets that are either sent to or received by
the DUT.

A mutation trace $Tr'$ represents a \textit{guess} of the corresponding DUT
packet trace given sniffer trace $Tr$.  In fact, the DUT packet trace must
be one of the mutation traces of the sniffer trace $Tr$.

\begin{lemma}
  Let $Tr_{DUT}$ and $Tr$ be the DUT and sniffer packet trace captured during
  the same protocol operation session, and $\mathcal{M}(Tr)$ be the set of
  mutation traces of $Tr$ with respect to DUT, then $Tr_{DUT} \in \mathcal{M}(Tr)$.
  \label{lem:mutation}
\end{lemma}%
\begin{proof}
  Let $\Delta = Tr \setminus Tr_{DUT}$ be the set of packets that are in $Tr$
  but not in $Tr_{DUT}$. Recall that it is not possible for the sniffer to
  observe packets sent from the DUT that the DUT did not send. Therefore,
  all packets in $\Delta$ are sent \textit{to} the DUT. That is, for
  all $(t, p) \in \Delta,\ p.dest = DUT$. By Definition~\ref{def:mutation},
  $Tr_{DUT} \in \mathcal{M}(Tr)$.
  \qed
\end{proof}

\subsection{Problem Statement}
\label{subsec:problem}

Lemma~\ref{lem:mutation} shows that $\mathcal{M}(Tr)$ is a \textit{complete} set
of guesses of the DUT packet trace. Therefore, the problem of validating DUT
implementation given a sniffer trace can be formally defined as follows:

\begin{problem}
  \label{prob:validation}
  VALIDATION\\
  \textbf{instance} A protocol monitor state machine $S$ and a sniffer trace $Tr$.\\
  \textbf{question} Does there exist a mutation trace $Tr'$ of $Tr$ that satisfies $S$?
\end{problem}

If the answer is no, a definite violation of the DUT implementation can be
claimed. Nevertheless, if the answer is yes, $S$ may still reject $Tr_{DUT}$.
In other words, the conclusion of the validation can either be
\textit{definitely wrong} or \textit{probably correct}, but not
\textit{definitely correct}.  This is the fundamental limitation caused by the
uncertainty of sniffer traces.

### 2017-02-20 case.tex
init

\section{Case Studies}
\label{sec:case}

We present case studies on applying our validation framework on two protocols
implemented in the \ns{} network simulator: 802.11 data transmission and ARF
rate control algorithm. The goal is to demonstrate how our framework can avoid
false alarms and report true violations on incomplete sniffer traces and report
true violations. We also report the application of our framework to a commercial
product under development.

\input{./case/tx.tex}
\input{./case/arf.tex}
\input{./case/xbox.tex}

### 2017-02-20 case/tx.tex
init

\subsection{802.11 Data Transmission}
\label{subsec:tx}

In this section, we first show that our framework can improve validation
precision by inferring the missing or extra packets using the augmented
transition framework. We then demonstrate the ability of our framework to detect
true violations by manually introducing bugs in the \ns{} implementation and
show the precision and recall of validation results.

\subsubsection{Experimental Setup}

We set up two \wifi{} devices acting as the transmitter (DUT) and receiver
respectively. Another \wifi{} device is configured in monitor mode and acts as
the sniffer. During the experiments,we collect both the DUT packet trace
(the ground truth) and the sniffer trace.

\subsubsection{Verifying Unmodified Implementation}

In the original monitor state machine shown in Fig.~\ref{fig:dot11_tx_ta}, we
set acknowledgment timeout $T_o=334\mu$s, maximum retransmission delay
$T_m=15$ms according to the protocol. We also adapt the state machine to include
multiple retransmissions\footnote{The exact number of retransmissions is not
part of the protocol, and NS-3 implementation set this to be 7.} instead of one.

Let $Pr_{ds}$, $Pr_{es}$ and $Pr_{ed}$ be the packet loss probability between
the DUT and sniffer, endpoint and sniffer, endpoint and DUT respectively.
$Pr_{ed}$ represents the characteristics of the system being tested, while
$Pr_{ds}$ and $Pr_{es}$ represent the sniffer's quality in capturing packets.

We vary each of the three probabilities, $Pr_{ds}$, $Pr_{es}$ and $Pr_{ed}$, from
0 to 0.5 (both inclusive) with 0.05 step.  For each loss ratio combination, we
ran the experiment 5 times, and each run lasted
30 seconds. In total, 6655 ($11^3\times 5$) pairs of DUT and sniffer packet
traces were collected.

To establish the ground truth of violations, we first verify the DUT packet
traces using the \textit{original} state machine $S$.  This can be achieved by
disabling augmented transitions in our framework.  As expected, no violation is
detected in any DUT packet traces.

We then verify the sniffer traces using the augmented state machine $S^+$.  For
the $\mathit{GoBack}(k)$ heuristic, we set $k=7$, which is the maximum number of
transmissions of a single packet. For the $\mathit{NumMissing}(d, l, k)$ heuristic, we
set the sliding window size $l=100$, and $k=80$ such that no violation is
reported. The relationship of $k$ and validation precision is studied in next
section.

Next, we present detailed analysis of the augmented transitions on the sniffer
traces. The goal is to study for a given system packet loss probability
$Pr_{ed}$, how the sniffer packet loss properties ($Pr_{ds}$ and $Pr_{es}$)
affect the difference between the DUT trace and the mutation trace, which represents
a guess of the DUT trace by the augmented state machine based on the sniffer
trace.

For all following analysis, we divide the traces into three groups according to
$Pr_{ed}$: low ($0 \le Pr_{ed} \le 0.15$), medium ($0.20 \le Pr_{ed} \le 0.35$)
and high ($0.40 \le Pr_{ed} \le 0.50$).

The different between two packet traces can be quantified by the Jaccard distance
metric.

\begin{align}
  Jaccard(Tr_1, Tr_2) = \frac{\left\vert Tr_1 \ominus Tr_2\right\vert}{\left\vert
  Tr_1 \cup Tr_2\right\vert}
\end{align}%
where $\ominus$ is the symmetric difference operator.  The distance is 0 if the
two traces are identical, and is 1 when the two traces are completely different.
The smaller the distance is, the more similar the two traces are.

A na\"ive way to calculate the Jaccard distance of two traces is to use the hash
of the $(time, packet)$ pair for set intersection and union operation.  However,
it does not work for mutation trace which contains fabricated packets with no
actual payload.  Therefore, we use a protocol specific canonical representation
of packets when calculating the distance.  In particular, the string
\texttt{r\_DATA\_i\_t} represents the $t^{th}$ transmission of a data packet
with sequence number $i$, and $r$ represents the round of sequence numbers as it
wraps after 4096.  And similarly \texttt{r\_Ack\_i\_t} is the corresponding
acknowledgment packet.

\begin{figure*}[t!]
  \centering
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./figures/scripts/MutationDUTJaccard3DFigure_0_10.pdf}
    % \caption{$Pr_{ed} \in [0.05, 0.1, 0.15]$}
    \caption{$0.05 \le Pr_{ed} \le 0.15$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./figures/scripts/MutationDUTJaccard3DFigure_0_30.pdf}
    \caption{$0.2 \le Pr_{ed} \le 0.35$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./figures/scripts/MutationDUTJaccard3DFigure_0_50.pdf}
    \caption{$0.4 \le Pr_{ed} \le 0.5$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \caption{\textbf{Jaccard Distance Between Mutation and DUT Traces.} For each
  data point, the mean of the 5 runs is used.}
  \label{fig:mutation_dut}
\end{figure*}

Fig.~\ref{fig:mutation_dut} shows the Jaccard Distance between mutation and its
corresponding DUT trace. We make the following observations. First, for a given
system loss probability $Pr_{ed}$ (each sub-figure), the lower the sniffer
packet loss probability ($Pr_{ds}$ and $Pr_{es}$), the smaller Jaccard distance
between the DUT and mutation trace. Intuitively, this means a sniffer that
misses less packets can enable our framework to better reconstruct the DUT
trace.

Second, we observe a \textit{protocol-specific} trend that $Pr_{ds}$ is more
dominant than $Pr_{es}$.  This is because retransmission packets of the same
sequence number are identical, hence when the sniffer misses multiple
retransmission packets, our framework only needs to infer one retransmission
packet to continue state machine execution.

Finally, as the system loss probability $Pr_{ed}$ increases, the Jaccard
distance increases more rapidly as $Pr_{ds}$ increases.  This is because the
ratio of retransmission packet increases along with $Pr_{ed}$.

We then evaluate the cost of resolving uncertainty. In particular, we use
the Average Search Steps Per Packet (ASSPP) as a metric to quantify the search
cost.  It is calculated by dividing the total number of search steps by the
number of packets in the packet trace. For DUT traces, ASSPP is always 1 since
there is no uncertainty. For sniffer traces, however, multiple search steps must
be to conducted to resolve the potential uncertainty of each packet in the
sniffer trace.

\begin{figure*}[t!]
  \centering
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./figures/scripts/StepCount3DFigure_0_10.pdf}
    \caption{$Pr_{ed} \in [0.05, 0.10, 0.15]$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./figures/scripts/StepCount3DFigure_0_30.pdf}
    \caption{$Pr_{ed} \in [0.2, 0.25, 0.3, 0.35]$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./figures/scripts/StepCount3DFigure_0_50.pdf}
    \caption{$Pr_{ed} \in [0.40, 0.45, 0.50]$}
  \end{subfigure}
  \caption{\textbf{Average Searching Step Per Packet.} For each data point, the mean of 5 runs
  is used.}
  \label{fig:cost}
\end{figure*}

Fig.~\ref{fig:cost} shows the distribution of ASSPP at different $Pr_{ed}$.
Similar to the case in Fig.~\ref{fig:mutation_dut}, $Pr_{ds}$ plays a dominant
role in determine the searching cost. One interesting observation is that the
search cost peaks when $Pr_{ds}$ is high while $Pr_{es}$ is low. In such loss
probability combinations, sniffer misses many data packets from the DUT but
picks up lots of \textit{dangling} $Ack$ packets from the DUT. Because the $Ack$
packet has neither sequence numbers nor retry flag, the searching algorithm had
a hard time resolving such uncertainty.

\subsubsection{Introducing Bugs}

We have demonstrated that our framework can tolerate sniffer imperfections and
avoid raising false alarms.  The next question is, can it detect true
violations?  To answer this question, we manually introduce several bugs in
\ns{} implementation that concerns various aspects of 802.11 data transmission
protocol.  More specifically, the bugs are:

\begin{itemize}
  \item \textbf{Sequence Number}: the DUT does not assign sequence number
    correctly. For example, it may increase sequence by 2 instead of 1, or it
    does not increase sequence number after certain packet, etc. We choose one
    type of such bugs in each run.
  \item \textbf{Semantic}: the DUT may retransmit even
    after receiving $Ack$, or does not retransmit when not receiving $Ack$.
\end{itemize}

We instrument the \ns{} implementation to embed instances of bugs in each
category. At each experiment run, we randomly decide whether and which bug to
introduce for each category. We fix $Pr_{ds}=Pr_{es}=0.1$ and vary $Pr_{ed}$
from 0.0 to 0.5 with 0.01 step. For each $Pr_{ed}$ value, we ran the experiment
100 times, of which roughly 75 experiments contained bugs. In total, 5100 pairs of
DUT and sniffer traces were collected.

We use the DUT packet traces as ground truth of whether or not each experiment
run contains bugs.
For each $Pr_{ed}$ value, we calculate the precision and recall of violation
detection using the sniffer traces.%
\begin{align}
  \text{Precision} = \frac{\left\vert \{\text{Reported Bugs}\} \cap \{\text{True Bugs}\}\right\vert}{\left\vert
  \{\text{Reported Bugs}\}\right\vert}\\
  \text{Recall} = \frac{\left\vert \{\text{Reported Bugs}\} \cap \{\text{True Bugs}\}\right\vert}{\left\vert
  \{\text{True Bugs}\}\right\vert}
\end{align}%
The precision metric quantifies how \textit{useful} the validation results are ,
while the recall metric measures how \textit{complete} the validation results
are.

\begin{figure}[t!]
  \centering
  \begin{subfigure}{0.48\textwidth}
    \includegraphics[width=\textwidth]{./figures/scripts/PrecisionFigure.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \includegraphics[width=\textwidth]{./figures/scripts/RecallFigure.pdf}
  \end{subfigure}
  \caption{\textbf{Precision and Recall of Validation Results.}}
  \label{fig:precision}
\end{figure}

Fig.~\ref{fig:precision} shows the CDF of precision and recall of the 51
experiments for various $k$ values. For precision, as expected, the more
tolerant the search to sniffer losses (larger $k$), the more tolerant the
framework is to sniffer losses, and the more precise the violation detection. In
particular, when $k=30$, the precisions are 100\% for all $Pr_{ed}$ values.
Second, the recall is less sensitive to the choice of $k$.  Except for the
extreme case when $k=30$, all other thresholds can report almost all the
violations.

### 2017-02-20 case/xbox.tex
init

\subsection{Industrial Application}
\label{subsec:xbox}

\begin{table}[t!]
  \centering
  \begin{tabular}{lll}
    \toprule
    \textbf{Protocol Aspects} & \textbf{Traces} & \textbf{Violations (\%)}\\
    \midrule
    Sequence Number & 3049 & 1539 (50.48\%) \\
    Station Scheduling & 3046 & 2045 (67.14\%) \\
    Uplink Modulation & 3127 & 8 (0.26\%) \\
    Downlink Modulation & 3127 & 24 (0.77\%) \\
    \bottomrule
  \end{tabular}
  \caption{\textbf{Validation Results on Traces from the Gaming Wireless Protocol}}
  \label{tab:xbox}
\end{table}

We have applied our framework for runtime verification of the wireless protocol
implementation of a commercial product that has several millions of shipping
devices. The product runs a proprietary wireless protocol to reduce latency and
power consumption. Sniffer traces were collected in regular testing process, but
were only manually inspected previously.

We obtained 75 sniffer traces from the testing team for a new version of the
protocol that is under development and testing. This team has been testing the
implementation for a few weeks.  Each trace contained around 6 million packets
that were captured during 1 hour and 40 minutes of protocol operation.

We split the traces into \num{100000} packet segments, and applied our framework
to validate the DUT implementation. We found that the latest implementation of
the protocol under development had violations of the protocol specification.
Some of the implementation bugs we found related to the sequence number
management, station scheduling, and the changing of modulation rates adaptation.
Table~\ref{tab:xbox} summarizes the validation result. Note that if we disable
the augmented transitions, each trace will be flagged as violation because of
the missing packets, thereby reducing the usability of the tool.

The bugs related to packet sequence number and station scheduler were detected
in about half the traces, and the bug related to the rate control algorithm was
detected in only a few traces. This is because the previous two aspects are
essential in all protocol operations, while the bugs related to rate control
only manifest themselves under certain link conditions.

After communicating with the testing team, we confirmed that the sequence number
bug was already known, as it is relatively easy to detect even by manually
examining the traces. The bug related to station scheduling was also noticed
before, yet no quantitative results about how frequent this bug appears were
obtained because of the lack of automatized validating tools. Finally, the bug
related to rate control has been filed as a bug report.

### 2017-02-20 framework/augment.tex
init

\subsection{Augmented State Machine}
\label{subsec:augment}

To deal with the inherent uncertainty of sniffer traces, we propose to
systematically augment the original monitor state machine with non-deterministic
transitions to account for the difference between the sniffer and DUT traces.

Before formally defining the augmented state machine, we first use an example to
illustrate the basic idea. Fig.~\ref{fig:augment} shows the augmented state
machine for 802.11 transmitter state machine shown in
Fig.~\ref{fig:dot11_tx_ta}.  For each existing transition (e.g., $s_0\rightarrow
s_1$), we add an \textit{empty transition} with same clock guards and resetting
clocks.  This accounts for the possibility when such packet was observed by
the DUT but missed by the sniffer.  Additionally, for each transition triggered
by a \textit{receiving} packet (i.e., $p.dest = DUT$), such as $s_1\rightarrow
s_0$ and $s_2\rightarrow s_0$, we add a \textit{self transition} with the same
trigger packet and clock guards, but an empty set of resetting clocks and no
assignments to variables. This allows the state machine to make progress when
the sniffer missed such packets.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.7\textwidth]{./figures/dot11_tx_checker.pdf}
  \caption{\textbf{Augmented Monitor State Machine.} Augmented transitions are
  highlighted in bold face. $\overline{Pkt}$ means either $\epsilon$ or $Pkt$.}
  \label{fig:augment}
\end{figure}

There are two things to note. First, self transitions are added only for
packets sent \textit{to} the DUT, since the sniffer will not overhear packets
\textit{from} the DUT if they were not sent by the DUT. Second, no augmented
transitions are added for the packets that are sent to DUT yet are missed by both
the DUT and the sniffer, since such packets do not cause difference between the
DUT and sniffer traces.

The augmented state machine in Fig.~\ref{fig:augment} will accept the sniffer
packet traces $Tr_1$ and $Tr_2$ shown in Fig.~\ref{fig:sniffer_in_middle}.  For
instance, one accepting transition sequence on sniffer trace $Tr_1$ is
$s_0\rightarrow s_1 \rightarrow_s s_1\rightarrow s_2 \rightarrow s_0$, and the
sequence for $Tr_2$ is $s_0 \rightarrow s_1 \rightarrow_e s_2 \rightarrow s_0$,
where $\rightarrow$ is the transition from the original state machine,
$\rightarrow_e$ and $\rightarrow_s$ are the augmented empty and self transitions
respectively.

We now formally define the augmented state machine.

\begin{definition}[Augmented Monitor]
  \sloppy{
  An augmented state machine $S^+$ for a monitor state machine $S$ is a 7-tuple
  $\{\boldsymbol{\Sigma^+}, \mathbb{S}, \mathbb{X}, s_0, C, \boldsymbol{E^+},
  G\}$, where $\mathbb{S}, \mathbb{X}, s_0, C, G$ are the same as $S$.
  $\Sigma^+=\{\epsilon\} \cup \Sigma$ is the augmented input alphabet with the
  empty symbol, and $E^+ \supset E$ is the set of transitions, which includes:
}
  \begin{itemize}
    \item $E$: existing transitions (\textbf{Type-0}) in $S$.
    \item $E^+_1$: empty transitions (\textbf{Type-1}) for transitions in $E$.
    \item $E^+_2$: self transitions \textbf{(Type-2)} for transitions
      triggered by receiving packets.
  \end{itemize}
\end{definition}

\begin{figure}[t!]
\begin{algorithm}[H]
  \caption{Obtain Augmented Transitions $E^+$ from $E$}
  \label{alg:augment}
  \begin{algorithmic}[1]
    \Function{augment}{$E$}
    \Let{$E^+$}{$\emptyset$}
    \ForAll{$\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, p, g, C'\rangle\}$\Comment{Type-0}}
      \label{alg:augment:type0}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, \boldsymbol{\epsilon}, g, C'\rangle\}$\Comment{Type-1}}
      \label{alg:augment:type1}
      \If{$p.dest = DUT$}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, \boldsymbol{s_i},
      \boldsymbol{v_i}, p, g, \boldsymbol{\emptyset}\rangle\}$\Comment{Type-2}}
        \label{alg:augment:type2}
      \EndIf
    \EndFor
    \State \Return{$E^+$}
    \EndFunction
  \end{algorithmic}
\end{algorithm}
\end{figure}

Algorithm~\ref{alg:augment} describes the process of transforming $E$ into
$E^+$. In particular, Line~\ref{alg:augment:type0} adds existing transitions in
$E$ to $E^+$, while line~\ref{alg:augment:type1} and~\ref{alg:augment:type2} add
Type-1 and Type-2 transitions to $E^+$ respectively.  We have highlighted the
elements of the tuple that differ from the underlying Type-0 transition. Note
that in Type-2 transitions, both the state and the variables stay the same after
the transition.

With augmented state machine $S^+$, we can use Type-1 transitions to
non-deterministically infer packets missed by the sniffer, and use Type-2
transitions to consume extra packets captured by the sniffer but missed by the
DUT.

A accepting run of $S^+$ on sniffer trace $Tr$ yields a mutation trace $Tr'$
which represents one possibility of the DUT trace. Specifically, $Tr'$ can be
obtained by adding missing packets indicated by Type-1 transitions to $Tr$, and
removing extra packets indicated by Type-2 transitions from $Tr$

We show that the VALIDATION problem is equivalent to the
satisfiability problem of $Tr$ on $S^+$.

\begin{theorem}
  There exists a mutation trace $Tr' \in \mathcal{M}(Tr)$ that satisfies $S$ if
  and only if $Tr$ satisfies $S^+$.
 \label{the:equivalent}
\end{theorem}
\begin{proof}
  Assume $Tr$ satisfies $S^+$, and $P$ is a sequence of accepting transitions,
  we construct a mutation trace $Tr'$ using $P$ and show that $Tr'$ satisfies
  $S$.

  \sloppy{%
    Initially, let $Tr'=Tr$, then for each \textit{augmented} transition $\langle s_i,
  v_i, s_j, v_j, \sigma, g, C'\rangle \in P$:
  }
  \begin{itemize}
    \item If this is a Type-1 transition, add $(t, p)$ to $Tr'$, where $t$ is a
      timestamp that satisfies $g$ and $p$ is the missing packet.
    \item If this is a Type-2 transition, remove corresponding $(t, p)$ from
      $Tr'$.
  \end{itemize}
  It is obvious that $Tr'$ is a mutation trace of $Tr$, since only receiving
  packets are removed in the process.

  Now we show that there exists a accepting transition sequence $P'$ of $S^+$ on
  input $Tr'$ that does not contain augmented transitions.  In particular, $P'$
  can be obtained by substituting all Type-1 transitions with corresponding
  original transitions, and removing all Type-2 transition.  Since $P'$ does not
  contain augmented transitions, it is also an accepting transition sequence of
  $S$ on input $Tr'$, hence $Tr'$ satisfies $S$.

  On the other hand, assume $Tr' \in \mathcal{M}(Tr)$ and $Tr'$ satisfies $S$.
  Suppose $P'$ is the accepting transition sequences of $S$ on input $Tr'$.
  We first note that $P'$ is also the accepting transitions of $S^+$ on input
  $Tr'$, since $E \subset E^+$.

  We construct a accepting transition sequence $P$ of $S^+$ on input $Tr$ as
  follows.
  \begin{itemize}
    \item For each packet $p \in Tr' \setminus Tr$, substitute the transition
      $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle$ with the corresponding Type-1
      transition $\langle s_i, v_i, s_j, v_j, \epsilon, g, C'\rangle$.
    \item For each transition $\langle s_i, v_i, s_j, v_j, \sigma, g, C'\rangle$
      followed by packet $p \in Tr\setminus Tr'$, add a Type-2 self
      transition $\langle s_j, v_j, s_j, v_j, p, g, \emptyset\rangle$. This is
      possible since $Tr'$ is a mutation trace of $Tr$, thus  for all $p \in Tr'
      \setminus Tr$, $p.src \ne DUT$.
  \end{itemize}
  Therefore, $Tr$ satisfies $S^+$.
  \qed
\end{proof}

By Theorem~\ref{the:equivalent}, the inherent uncertainty of the sniffer traces
is explicitly represented by the augmented transitions, and can be
systematically explored using the well established theory of state machine.

\begin{comment}
One immediate observation can be drawn from Theorem~\ref{the:equivalent} by
contradiction.

\begin{corollary}
  If $S^+$ rejects $Tr$, then $S$ rejects $Tr_{DUT}$.
  \label{cor:false_pos}
\end{corollary}
In the context of validation where we raise a violation alarm when $S^+$
rejects $Tr$, Corollary~\ref{cor:false_pos} provides a {\it sufficient} condition for reporting incorrect traces. 
However, when $S^+$ accepts $Tr$, $S$ could still reject $Tr_{DUT}$.
In other words, the conclusion of the validation can either be \textit{definitely
wrong} or \textit{possibly correct}, but not \textit{definitely correct}.
This is the fundamental limitation caused by the uncertainty of sniffer traces.
\end{comment}

### 2017-02-20 framework/hardness.tex
init

\subsection{Problem Hardness}
\label{subsec:hard}

In this section, we show that the VALIDATION problem is NP-complete. In
fact, the problem is still NP-complete even with only one type of augmented
transitions.

Recall that Type-1 transitions are added because the sniffer may miss packets.
Suppose an imaginary sniffer that is able to capture \textit{every} packet ever
transmitted, then only Type-2 transitions are needed since the sniffer may still
overhear packets sent to the DUT. Similarly, suppose another special sniffer
that would not overhear any packets sent to the DUT, then only Type-1
transitions are needed to infer missing packets.

We refer the augmented state machine that only has Type-0 and Type-1
transitions as $S^+_1$, and the augmented state machine that only has Type-0 and
Type-2 transitions as $S^+_2$. And we show that each subproblem of determining
trace satisfiability is NP-complete.

\begin{problem}
  VALIDATION-1\\
  Given that $Tr\setminus Tr_{DUT}=\emptyset$ (sniffer does not overhear
  packets).\\
  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
  \textbf{question} Does $S^+_1$ accept $Tr$?
\end{problem}

\begin{problem}
  VALIDATION-2\\
  Given that $Tr_{DUT} \subseteq Tr$ (sniffer does not missing packets).\\
  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
  \textbf{question} Does $S^+_2$ accept $Tr$?
\end{problem}

\begin{lemma}
  \label{lem:sub}
  Both VALIDATION-1 and VALIDATION-2 are NP-complete.
\end{lemma}
\begin{proof}
  First, note that the length of mutation trace $Tr'$ is polynomial to the
  length of $Tr$ because of the discrete time stamp and non-overlapping packets
  assumption.
  Therefore, given a state transition sequence as witness, it can be verified in
  polynomial time whether or not it is an accepting transition sequence, hence
  both VALIDATION-1 and VALIDATION-2 are in NP.

  Next, we show how the SAT problem can be reduced to either one of the two
  problems.
  Consider an instance of SAT problem of a propositional formula $F$ with $n$
  variables $x_0,x_1,\ldots, x_{n-1}$, we construct a corresponding protocol and
  its monitor state machine as follows.

  The protocol involves two devices: the DUT (transmitter) and the endpoint
  (receiver).
  The DUT shall send a series of packets, $pkt_0, pkt_1,\ldots, pkt_{n-1}$.
  For each $pkt_i$, if the DUT receives the
  acknowledgment packet $ack_i$ from the endpoint, it sets boolean variable
  $x_i$ to be \texttt{true}.
  Otherwise $x_i$ remains to be \texttt{false}.
  After $n$ rounds, the DUT evaluate the formula $F$ using the assignments and
  sends a special packet, $pkt_{true}$, if $F$ is \texttt{true}.
  One round of the protocol operation can be completed in polynomial time since
  any witness of $F$ can be evaluated in polynomial time.

  \begin{figure}[t!]
    \centering
    \includegraphics[width=0.7\textwidth]{./figures/sat_sm.pdf}
    \caption{\textbf{Monitor State Machine for SAT Problem.}}
    \label{fig:sat}
  \end{figure}

  The protocol monitor state machine $S$ is shown in Fig.~\ref{fig:sat}.
  Initially, all $x_i$ is set to \texttt{false}.
  At state $s_0$, the DUT shall transmit $pkt_i$ within a unit time, transit to
  $s_1$ and reset the clock along the transition.
  At state $s_1$, either the DUT receives the $ack_i$ packet and
  set $x_i$ to be \texttt{true} ($s_1 \rightarrow s_0$), or the DUT continues to
  transmit the next packet $pkt_{i+1}$.
  After $n$ rounds, the state machine is $s_0$ or $s_1$ depending on whether
  $ack_{n-1}$ is received by the DUT.
  In either case, the DUT shall evaluate $F$ and transmit $pkt_{true}$ if $F$ is
  \texttt{true}.

  \sloppy{%
    Consider a sniffer trace $Tr_1=\{(0, pkt_0), (2, pkt_1), (4, pkt_2),\ldots,
  (2(n-1), pkt_{n-1}), (2n, pkt_{true})\}$.  That is, the sniffer only captures
  all $pkt_i$ plus the final $pkt_{true}$, but none of $ack_i$.  It is easy to see
  that $F$ is satisfiable if $S_1^+$ accepts $Tr_1$.  In particular, a successful
  run of $S_1^+$ on $Tr_1$ would have to guess, for each $pkt_i$, whether the
  Type-1 empty transitions should be used to infer the missing $ack_i$ packet,
  such that $F$ is \textit{true} at the end.  Note that for $Tr_1$, no Type-2 self
  transitions can be used since all packets in $Tr_1$ are sent from the DUT.
  Therefore, the SAT problem of $F$ can be reduced to the VALIDATION-1 problem of
  $S^+_1$ on sniffer trace $Tr_1$.
  }

  \sloppy{%
    On the other hand, consider another sniffer trace $Tr_2 =\{(0, pkt_0), (1,
    ack_0), (2, pkt_1), (3, ack_1),\ldots, (2n-2, pkt_{n-1}), (2n-1, ack_{n-1}),\\ (2n,
  pkt_{true}\}$. That is, the sniffer captures all $n$ pair of $pkt_i$ and $ack_i$
  packets and the final $pkt_{true}$ packet.  Similar to $Tr_1$, $F$ is
  satisfiable if $S^+_2$ accepts $Tr_2$. A successful transition sequence of
  $S^+_2$ on $Tr_2$ must decide for each $ack_i$ packet, whether Type-2 self
  transitions should be used, so that $F$ can be evaluated as true at the end.
  Therefore, the SAT problem of $F$ can also be reduced to the VALIDATION-2
  problem of $S^+_2$ on sniffer trace $Tr_2$.
}

Since SAT is known to be NP-complete, both the VALIDATION-1 and the
VALIDATION-2 problem are also NP-complete.  \qed
\end{proof}

The hardness statement of the general VALIDATION problem naturally follows
Lemma~\ref{lem:sub}.

\begin{theorem}
  VALIDATION is NP-complete.
\end{theorem}

### 2017-02-20 framework/heuristic.tex
init

\subsection{Pruning Heuristics}
\label{subsec:heuristic}

In the face of uncertainty between a possible
protocol violation and sniffer imperfection, augmented transitions provide the
ability to blame the latter. The exhaustive nature of
Algorithm~\ref{alg:search} means that it always tries to blame sniffer
imperfection whenever possible, making it reluctant to report true
violations.

Inspired by the \textit{directed model checking}~\cite{edelkamp2008survey}
technique which is to mitigate the state explosion problem, we propose to enforce extra
constraints on the mutation trace to restrict the search to only mutation traces
with high likelihood. The modified \texttt{EXTEND} function checks certain
likelihood constraints on the prefix of the mutation trace before continuing
(line~\ref{alg:search:stop}), and stops the current search branch immediately if
the prefix seems \textit{unlikely}.  Because of the recursive nature of the
algorithm, other branches which may have a higher likelihood can then be
explored.

The strictness of the likelihood constraint represents a trade-off between
precision and recall of validation. The more strict the constraints are, the
more false positive violations will potentially be reported, hence the lower the
precision yet higher recall. On the contrary, the more tractable the
constraints are, the more tolerant the search is to sniffer imperfection, hence
the more likely that it will report true violations, thus higher precision but
lower recall.

The exact forms of the constraints may depend on many factors, such as the
nature of the protocol, properties of the sniffer, or domain knowledge.  Next,
we propose two \textit{protocol oblivious} heuristics based on the sniffer loss
probabilities and general protocol operations. Both heuristic contains
parameters that can be fine tuned in practice.

\subsubsection{$\mathit{NumMissing}(d, l, k)$}

This heuristic states that the number of missing packets from device $d$ in any
sub mutation traces of length $l$ shall not exceed $k$ ($k \le l$).  The sliding
window of size $l$ serves two purposes. First, $l$ should be large enough for
the calculated packet loss ratio to be statistically meaningful.  Second, it
ensures that the packet losses are evenly distributed among the entire packet
trace.

The intuition behind this heuristic is that the sniffer's empirical packet loss
probability can usually be measured before validation. Therefore, the likelihood
that the sniffer misses more packets than prior measured loss ratio is quite
low. The value of $l$ and $k$ can then be configured such that $k/l$ is marginally
larger than the measured ratio.

\begin{comment}
In addition, we note that for a fixed $l$, a single threshold $k$ may not work
for traces collected by sniffers with different packet loss probabilities.
Intuitively, sniffers with high loss probabilities require a larger threshold.
Therefore, we perform a \textit{stratified} search with an increasing sequence
of thresholds $\{k_1, k_2,\ldots, k_n\}$ ($k_i < k_{i+1}$ for $1 \le i < n$).
At round $i$, the search is restricted by constraint $NumMissing(d, l, k_i)$.
If a mutation trace is found, the search is completed.  Otherwise, we increase
the round number and repeat the search process.  In this way, this heuristic can
automatically adapt to the underlying sniffer packet loss probabilities. And a
violation is declared when no mutation trace can be found with $NumMissing(d, l,
k_n)$.
\end{comment}

\subsubsection{$\mathit{GoBack}(k)$}

This heuristic states that the search should only
backtrack at most $k$ steps when the search gets stuck using only $E$.
The motivation is that many protocols operate as a sequence of independent
transactions, and the uncertainty of previous transactions often do not affect
the next transaction.
For instance, in 802.11 packet transmission protocol, each packet exchange,
include the original, retransmission and acknowledgment packets, constitute a
transaction.
And the retransmission status of previous packets has no effect on the packets
with subsequent sequence numbers, hence need not be explored when resolving the
uncertainty of the packets with new sequence numbers.  Note that we do not
require the protocol to specify an exact transaction boundary, but only need $k$
to be sufficiently large to cover a transaction.

\begin{comment}
\subsection{Confidence of Correctness}

A run of $S^+$ on sniffer packet trace $Tr$ may yield two conclusions about the
DUT implementation: either ``definitely wrong'' when $S^+$ rejects $Tr$, or
``possibly correct'' when $S^+$ accepts $Tr$. We now quantify the
\textit{confidence} of the implementation correctness in the later case.

Note that the augmented transitions in $S^+$ implicitly embed the assumptions
about packet losses. In particular, Type-1 transitions assume the sniffer
missed certain packets, while Type-2 transitions assume the sniffer
overheard packets that were missed by the DUT. Intuitively, the portion of such
assumptions (hence packet loss \textit{ratios}) in an accepting transition
sequence should match with the true packet loss \textit{probabilities}.
Otherwise, the correctness confidence, despite $S^+$ accepts $Tr$, is low.

Suppose $Tr$ is the sniffer trace and $Tr'$ is its mutation that satisfies $S$.
We can calculate the three packet loss ratios in our model shown in
Figure~\ref{fig:prob}.

\begin{align}
  &Pr_{ds} = \frac{\parallel \{(t,p) \in Tr'\setminus Tr \land p.src = DUT\}\parallel}
  {\parallel \{(t,p) \in Tr' \cup Tr \land p.src = DUT\}\parallel}\\
  &Pr_{es} = \frac{\parallel \{(t,p) \in Tr'\setminus Tr \land p.src = EP\}\parallel}
  {\parallel \{(t,p) \in Tr' \cup Tr \land p.src = EP\}\parallel}
\end{align}where $Tr'\setminus Tr$ is the set of inferred packets that were missed by the
sniffer (Type-1 transitions in $S^+$).

Note that we can not directly calculate the packet loss ratio from Endpoint to
DUT, since there might be packets that were missed by both the sniffer and the
DUT. We can, however, approximate $Pr_{ed}$ using conditional packet loss
probability.

\begin{align}
  \nonumber
  &(1-Pr_{es})\times Pr_{ed} =\\
  &\frac{\parallel \{(t,p) \in Tr\setminus Tr' \land p.src = EP \land p.dest = DUT\}\parallel}
  {\parallel \{(t,p) \in Tr \cup Tr' \land p.src = EP \land p.dest = DUT\}\parallel}
\end{align}The LHS is the probability that a packet sent from the Endpoint to the DUT was
missed by the DUT yet captured by the sniffer. And the RHS is the ratio of
such packets in the mutation trace.

Consider the 802.11 data transmission protocol and the example trace snippet
shown in Figure~\ref{fig:loss}, the packet loss ratios are:
\begin{align*}
  Pr_{ds} &= \frac{\parallel\{DATA'_0\}\parallel}{\parallel\{DATA_0, DATA'_0\}\parallel} = 0.5\\
  Pr_{es} &= \frac{\parallel\emptyset\parallel}{\parallel\{Ack, Ack\}\parallel} = 0\\
  Pr_{ed} &= \frac{\parallel\{Ack\}\parallel}{\parallel\{Ack, Ack\}\parallel\times (1-Pr_{es})} = 0.5
\end{align*}

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.4\textwidth]{./figures/loss.pdf}
  \caption{\textbf{Example Sniffer Trace and Its Mutation Trace.}}
  \label{fig:loss}
\end{figure}

Suppose we have the true packet loss probabilities $\overline{Pr_{ds}},
\overline{Pr_{es}}$ and $\overline{Pr_{ed}}$, the \textit{confidence} of this
mutation trace $Tr'$ can be defined as:

\begin{align}
  Confidence(Tr') = \prod_{i\in\{ds, es, ed\}}(1-|Pr_i-\overline{Pr_i}|)
\end{align}

By this definition, the trace confidence maximizes when all packet loss
ratios calculated from $Tr'$ matches the true packet
loss probabilities.

Suppose we have $M$ mutation traces that satisfies $S$, then
the overall \textit{confidence of correctness} from sniffer trace $Tr$ can be defined as:

\begin{align}
  Confidence(Tr) &= 1-\prod_{j=1}^{M}(1-Pr(Tr'_j))
\end{align}

Since $0 \le 1-Pr(Tr'_j) \le 1$ for all $j$, we have:

\begin{align}
  \nonumber
  Confidence(Tr) &\ge 1-(1-Pr(Tr'_j)\\
     &= Pr(Tr'_j)
\end{align}for all $j$.

This means if we can find a mutation trace $Tr'$ that satisfies $S$, then the
overall confidence of the correctness is no less than $Pr(Tr')$.
\end{comment}

### 2017-02-20 framework/search.tex
init

\subsection{Searching Strategies}
\label{subsec:search}

In this section, we present an \textit{exhaustive} search algorithm of the
accepting transition sequence of $S^+$ on sniffer trace $Tr$. It is guaranteed to
yield an accepting sequence if there exists one, thus is exhaustive.  In the next
sections, we present heuristics to limit the search to accepting sequences of
$S^+$ that require relatively fewer transitions from $E^+_1 \cup E^+_2$.  Due to
the NP-completeness of the problem, this also makes the algorithm meaningful in
practice.

\begin{figure}[t!]
  \begin{minipage}{\textwidth}
    \begin{algorithm}[H]
      \caption{Exhaustive search algorithm of $S^+$ on $Tr$.}
      \label{alg:search}

      \begin{algorithmic}[1]
        \Function{search}{S, Tr}
          \Let{$S^+$}{\Call{augment}{S}}
          \State\Return{\Call{extend}{[], Tr, $S^+.s_0$}}
        \EndFunction
        \Function{extend}{prefix, p::suffix, $s$}
          \If{not \Call{likely}{prefix}}
            \label{alg:search:stop}
            {\Return{\nil}}\footnote{This check should be ignored in the exhaustive algorithm.}
          \EndIf
          \For{i $\in$ [0, 1, 2]}
          \label{alg:search:order}
          \Let{mutation}{EXTEND-i(prefix, p::suffix, $s$)}
            \If{mutation $\ne$ \nil}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State\Return{\nil}
        \EndFunction
        \Function{extend-0}{prefix, p::suffix, $s$}
          \For{$\langle s, s', p\rangle\footnote{$\langle s, s', p \rangle$ is
              short for $\langle s, *, s', *, p, *, *\rangle$} \in E$}
            \If{suffix = \nil}
              {\Return{prefix@p}}
              \label{alg:search:terminate_0}
            \EndIf
            \Let{mutation}{\Call{extend}{prefix@p, suffix, $s'$}}
            \label{alg:search:append}
            \If{mutation $\ne$ \nil}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State{\Return{\nil}}
        \EndFunction
        \Function{extend-1}{prefix, p::suffix, $s$}
          \ForAll{$\langle s, s', q\rangle \in E^+_1$}
            \If{q.time $>$ p.time}
              \label{alg:search:type_1_limit}
              \State{\textbf{continue}}
            \EndIf
            \Let{mutation}{\Call{extend}{prefix@q, p::suffix, $s'$}}
            \label{alg:search:guess}
            \If{mutation $\ne \nil$}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State{\Return{\nil}}
        \EndFunction
        \Function{extend-2}{prefix, p::suffix, $s$}
          \ForAll{$\langle s, s, p \rangle \in E^+_2$}
            \If{suffix = \nil}
              {\Return{prefix}}
              \label{alg:search:terminate_2}
            \EndIf
            \Let{mutation}{\Call{extend}{prefix, suffix, $s$}}
            \label{alg:search:drop}
            \If{mutation $\ne$ \nil}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State{\Return{\nil}}
        \EndFunction
      \end{algorithmic}
    \end{algorithm}
    \renewcommand\footnoterule{}
  \end{minipage}
\end{figure}

The main routines of the algorithm are shown in Algorithm~\ref{alg:search}.  In
the top level \texttt{SEARCH} routine, we first obtain the augmented state
machine $S^+$, and then call the recursive \texttt{EXTEND} function with an empty
prefix, the sniffer trace, and the $S^+$'s initial state. 
In the \texttt{EXTEND} function, we try to consume the first packet in the
remaining trace using either Type-0, Type-1 or Type-2 transition.
Note that we always try to use Type-0 transitions before other two augmented
transitions (line~\ref{alg:search:order}).
This ensures the first found mutation trace will have the most number of Type-0
transitions among all possible mutation traces.
Intuitively, this means the search algorithm tries to utilize the sniffer's
observation as much as possible before being forced to make assumptions.

Each of the extend functions either returns the mutation trace $Tr'$, or
\textit{\nil} if the search fails.
In both \texttt{EXTEND-0} and
\texttt{EXTEND-2} function, if there is a valid transition, we try to consume
the next packet either by appending it to the prefix
(line~\ref{alg:search:append}) or dropping it (line~\ref{alg:search:drop}).
While in \texttt{EXTEND-1}, we guess a missing packet without consuming the next
real packet (line~\ref{alg:search:guess}).
Note that since only Type-0 and Type-2 consume packets, the recursion terminates
if there is a valid Type-0 or Type-2 transition for the last packet
(line~\ref{alg:search:terminate_0} and line~\ref{alg:search:terminate_2}).

It is not hard to see that Algorithm~\ref{alg:search} terminates on any sniffer
traces. Each node in the transition tree only has finite number of possible next
steps, and the depth of Type-1 transitions is limited by the time available
before the next packet (line~\ref{alg:search:type_1_limit}).

### 2017-02-20 introduction.tex
init

\section{Introduction}
\label{sec:intro}

Custom wireless protocols are often designed and deployed to meet the specific
performance and power needs of special-purpose wireless devices. Examples
include Google
Iris contact lenses~\cite{iris}, Xbox One wireless controllers~\cite{xbox}, and
Google Chromecast~\cite{chromecast}. Validating that device implementations work
correctly is critical to achieve the design goals of the wireless protocol and
also prevent bugs in shipped products~\cite{wifried,lollipop,surface}.

Runtime validation of the protocol implementations on such devices is
challenging because collecting traces from the device under test (DUT) is often
infeasible. The resource limitations of embedded or battery-powered devices may
cause them to not provide trace collecting capabilities. DUT may contain
proprietary hardware or firmware that hides the implementation details and
prevents testers from collecting traces through source code instrumentation.
Even when collecting trace directly from the DUT is possible, the overhead it causes
may alter the behavior of the DUT due to the observer
effect~\cite{mytkowicz2008observer}, threatening the validation results.

An attractive alternative is to use wireless
sniffers to record traffic generated by the DUT during testing.
Sniffers do not require direct access to the DUT or the need to alter its behavior.
However, due to the fundamentally unpredictable nature of wireless
communications, the packets captured by the sniffer will not exactly match
those received by the DUT.
The sniffer may miss packets that the DUT received, or receive packets that
the DUT missed.
This is true even when using multiple
sniffers~\cite{cheng2006jigsaw,mahajan2006analyzing,bahl2006enhancing}, a sniffer
with multiple antennas~\cite{omnipeek}, or in isolated wireless environments.

Since the sniffer trace may not perfectly match the actual trace, uncertainty
arises during protocol implementation validation.  For example, if the DUT fails
to respond correctly to a packet in the sniffer trace, it may either because the
DUT's implementation is incorrect, or the DUT did not actually receive the
packet, or the DUT's response was missed by the sniffer.  Whenever the DUT's
behavior does not match the specification, there are now two potential
explanations: either the DUT's implementation is wrong, or the sniffer trace is
inaccurate. Accurate validation requires  distinguishing between these two
causes.

We present a new technique than enables validation of protocol implementations using
wireless sniffers. Given a monitor state machine representing the protocol being
validated, we describe a systematic transformation that adds non-deterministic
transitions to incorporate uncertainty introduced by the sniffer. This
augmented validation state machine implicitly defines a set of
mutated traces, each satisfying the original state machine with a specific
likelihood. If the set is empty, the implementation definitely violates the
protocol. Searching over all the mutated traces is NP-complete, but the approach
can be made practical by applying protocol-oblivious heuristics that limit the
search to likely mutated traces.

Our paper makes the following contributions:
\begin{enumerate}
  \item To the best of our knowledge, we are the first to identify the
    uncertainty problem caused by sniffers in validating wireless protocol
    implementations.
  \item We formalize the problem using a nondeterministic state machine that
    systematically and completely encodes the uncertainty of the
    sniffer trace.
  \item We characterize the NP-completeness of the validation problem, and
    present two protocol-oblivious heuristics to prune the search space and make
    validation possible in practice.
  \item We implement the validation framework and evaluate it using the \ns{}
    network simulator~\cite{riley2010ns}. Our framework accurately identifies
    both synthetic and previously unknown violations in \ns{}'s implementations
    of the 802.11 and ARF protocols.  We also applied our framework to a
    commercial product under development and was able to found three latent bugs
    that were not observed previously.
\end{enumerate}

### 2017-02-20 motivation.tex
init

\section{Background and Motivating Example}
\label{sec:motivation}

We encountered the uncertainty problem while testing the protocol implementation
of a popular wireless game controller. A custom wireless protocol
was designed to meet the low latency and low power consumption goals. As is
common industry practice, the protocol specification was then handed over to
wireless chipset vendors for implementation. However, neither implementation
details nor trace collection capabilities are provided in the shipped firmware
due to intellectual property constraints and device resource limitation. Hence
using sniffers to validate the protocol implementation is the only option.

We initially developed a tool to validate certain protocol properties over the
sniffer trace, yet often found unacceptable amount of false alarms due to
the incompleteness of the sniffer traces, making the tool virtually useless. It
was clear that we needed to account for sniffer uncertainty.

\begin{figure}[H]
  \vspace*{-3mm}
  \centering
  \includegraphics[width=0.6\textwidth]{./figures/dot11_tx_ta.pdf}
  \caption{\textbf{Monitor State Machine for 802.11 Transmitter.}}
  \label{fig:dot11_tx_ta}
  \vspace*{-3mm}
\end{figure}

To better understand the incompleteness of sniffer trace, consider the IEEE
802.11 (also known as \wifi{}) transmitter (DUT) state machine shown in
Fig.~\ref{fig:dot11_tx_ta}. After the DUT sends $DATA_i$---a data packet with
sequence number $i$ ($s_0\rightarrow s_1)$, it starts a timer and waits for the
acknowledgment packet---$Ack$. The DUT either receives $Ack$ within time $T_o$
($s_1\rightarrow s_0$), or it sends $DATA'_i$---retransmission of $DATA_i$
($s_1\rightarrow s_2$). Similarly, the DUT either receives the $Ack$ within $T_o$
($s_2\rightarrow s_0$) or aborts transmission and moves on to next
packet\footnote{To represent the state machine succinctly, our example assumes
that the DUT retries at most once.} ($s_2\rightarrow s_1$).

Given a complete log of DUT's packet transmission and reception events, it is
trivial to feed such a log into the state machine in Fig.~\ref{fig:dot11_tx_ta}
and validate the correctness of DUT's protocol implementation. However, due to
DUT limitations we have described earlier, this complete log is not
available. As a result, we seek to validate the DUT implementation using
sniffers.

There are two fundamental properties in wireless communication that bring uncertainty to
sniffer's observation: packet loss and physical diversity. The sniffer could
either miss packets sent from or to the DUT due to packet loss, or overhear packets
that are sent to but missed by the DUT due to physical diversity.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.6\textwidth]{./figures/false_pos.pdf}
  \caption{\textbf{Uncertainty of Sniffer Observations.} $Tr_{OTA}$ is
    the chronological sequence of packets sent by the DUT and the receiver.
    $Tr_{DUT}$ is DUT's internal events. $Tr_1$ and $Tr_2$ are two examples of
    many possible sniffer traces.}
  \label{fig:sniffer_in_middle}
\end{figure}

Consider a correct packet exchange sequence shown in
Fig.~\ref{fig:sniffer_in_middle}. The DUT first sends $DATA_0$.  Suppose the
receiver receives $DATA_0$ and sends the $Ack$ which the DUT does not receive.
Eventually the DUT's timer fires and it sends $DATA'_0$.  This time the
$DATA'_0$ reaches receiver and the DUT also receives the $Ack$.

Now consider two possible traces that could have been overheard by a sniffer shown
in Fig.~\ref{fig:sniffer_in_middle}.
In first sniffer trace $Tr_1$ where the sniffer
\textit{overhears} the first $Ack$ packet, a validation \textit{uncertainty}
arises when the sniffer sees the $DATA'_0$: was the previous $Ack$ missed by the
DUT or is there a bug in DUT which causes it to retransmit even after receiving
the $Ack$?

Similarly, consider the second possible sniffer trace $Tr_2$
where both the $DATA'_0$ and $Ack$ packets were missed by the sniffer. During
this period of time, it appears the DUT neither receives $Ack$ for $DATA_0$ nor
sends $DATA'_0$. Again, without any additional information it is impossible to
disambiguate between the sniffer missing certain packets and a bug in DUT's
retransmission logic.

Informally, the question we set out to answer in this paper is: given the
protocol monitor state machine and the sniffer's observation with inherent
uncertainty, how to accurately validate that the DUT behaves as specified?

### 2017-02-20 conclusion.tex
conclusion

\section{Conclusions}
\label{sec:conclusion}

We formally define the uncertainty problem in validating wireless protocol
implementations using sniffers. We describe a systematic augmentation of the
protocol state machine to explicitly encode the uncertainty of sniffer traces.
We characterize the NP-completeness of the problem and propose both an
exhaustive search algorithm and heuristics to restrict the search to more likely
traces. We present two case studies using \ns{} network simulator to demonstrate
how our framework can improve validation precision and detect real bugs. We also
report the application of our framework on a commercial product under
development.

Finally, we discuss a few challenges and future
directions.

\textbf{Verification Coverage.} Given a single sniffer trace, it is possible
that not all the states in the state machine are visited during the verification
process. For instance, a rate control state machine based on certain consecutive
packet losses patterns can not be verified if no such consecutive losses appear
in the sniffer trace. In general, given a protocol state machine, it is
challenging to extract the packet patterns for each state to be reached and to
alter the testing such that such patterns can be observed.

\textbf{State Machine Generation.} We manually translated the protocols studied
in this paper into monitor state machines based on the source code, comments and
documentation. The process is time-consuming and error-prone. A more scalable
approach would be taking the protocol specification written in certain formal
language, and automatically translate such specification into state machines
that can be used for verification process.

### 2017-02-22 case/arf.tex
title page format

\subsection{ARF Rate Control Algorithm}
\label{subsec:arf}

Automatic Rate Fallback (ARF)~\cite{kamerman1997wavelan} is the first rate
control algorithm in literature. In ARF, the sender increases the bit rate after
$Th_1$ number of consecutive successes or $Th_2$ number of packets with at most
one retransmission. The sender decreases bit rate after two consecutive packet
failures or if the first packet sent after rate increase (commonly referred as
\textit{probing} packet) fails.

\sloppy{%
  Fig.~\ref{fig:arf_sm} shows the state machine $S$ for the packet trace
  collected at sender (DUT), where $DATA_i^r$ denotes a data packet
  with sequence number $i$ and bit rate $r$, $DATA_i^{r\prime}$ is a retransmission packet
  and $Ack$ is the acknowledgment packet. The \texttt{pkg\_succ} function is
  shown in Algorithm~\ref{alg:pkt_succ}.
}

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.8\textwidth]{./figures/arf_sm.pdf}
  \caption{\textbf{Monitor State Machine for ARF Rate Control Algorithm.} Timing
  constraints are omitted for succinctness.}
  \label{fig:arf_sm}
\end{figure}

The \texttt{succ} variable is used to track the number of consecutive packet
successes. It is increased after each packet success , and is reset to 0 after a
rate increase or upon a packet failure ($s_1\rightarrow s_2$).  Similarly,
\texttt{count} is to track the number of packets with at most one
retransmission, and is increased after packet success, or for the first packet
retransmission ($s_1\rightarrow s_2$). It is reset when there are two
consecutive packet failures ($s_2\rightarrow s_3$). Finally, the \texttt{probe}
flag is set upon rate increases to indicate the probing packet, and is cleared
upon packet success. The variable \texttt{r} is the current bit rate, which is
decreased if the probing packet fails ($s_1\rightarrow s_4$), or every two
consecutive failures ($s_2\rightarrow s_3$). If \texttt{r} is not the highest
rate, it is increased when either of the two thresholds are reached.

\begin{algorithm}[t!]
  \caption{\texttt{pkt\_succ} function}
  \label{alg:pkt_succ}
  \begin{algorithmic}[1]
    \Function{pkt\_succ}{}
    \Let{i}{(i+1)\%N}
    \Let{succ}{succ + 1}
    \Let{count}{count + 1}
    \Let{probe}{false}
    \If{r $< R$ and (succ $\ge Th_1$ or count $\ge Th_2$)}
    \Let{r}{r+1}
    \Let{succ}{0}
    \Let{count}{0}
    \Let{probe}{true}
    \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

In particular, the bug we found lies in the implementation's \texttt{pkt\_succ}
function in line 6. Instead of checking \texttt{count $\ge$ Th\_2}, the
implementation checks \texttt{count == Th\_2}. This bug also exists in the NS-3
implementation of Adaptive ARF (AARF) algorithm~\cite{lacage2004ieee} and the
pseudo code implementation of AARF in~\cite{lacage2004report}.

Note that the \texttt{count} variable is incremented twice if a packet succeed
after one retransmission: once in $s_1\rightarrow s_2$, once in the
\texttt{pkt\_succ} function for the retransmission packet. Therefore, if the
value of \texttt{count} is $Th_2-1$ and the next packet succeed after one
retransmission, the value of \texttt{count} will be $Th_2+1$, which would fail
the implementation's test of \texttt{count == Th\_2}.

We report a bug found in NS-3 ARF~\cite{kamerman1997wavelan} implementation
which causes the sender to get stuck at a lower rate even after enough number of
consecutive successes. The bug was detected using sniffer traces and
confirmed by both the DUT trace and source code inspection.

### 2017-02-22 introduction.tex
title page format

@@
 \end{enumerate}
+
+This paper is an extension of our work that appeared in RV 2016~\cite{Shi2016}.
+We added the proof for Lemma~\ref{lem:mutation}, \ref{lem:sub} and
+Theorem~\ref{the:equivalent}. We included new results that show the searching
+cost (Fig.~\ref{fig:cost}), and details of the ARF protocol violations
+(Section~\ref{subsec:arf}). Further, we reported the application of our
+framework to a commercial product under development in
+Section~\ref{subsec:xbox}.

### 2017-02-24 case/arf.tex
\it{DATA}

@@
   Fig.~\ref{fig:arf_sm} shows the state machine $S$ for the packet trace
-  collected at sender (DUT), where $DATA_i^r$ denotes a data packet
-  with sequence number $i$ and bit rate $r$, $DATA_i^{r\prime}$ is a retransmission packet
+  collected at sender (DUT), where $\it{DATA}_i^r$ denotes a data packet
+  with sequence number $i$ and bit rate $r$, $\it{DATA}_i^{r\prime}$ is a retransmission packet
   and $Ack$ is the acknowledgment packet. The \texttt{pkg\_succ} function is

### 2017-02-24 case/tx.tex
\it{DATA}

@@
 of packets when calculating the distance.  In particular, the string
-\texttt{r\_DATA\_i\_t} represents the $t^{th}$ transmission of a data packet
+\texttt{r\_\it{DATA}\_i\_t} represents the $t^{th}$ transmission of a data packet
 with sequence number $i$, and $r$ represents the round of sequence numbers as it

### 2017-02-24 framework/checker.tex
\it{DATA}

\subsection{Packet, Trace and Monitor}
\label{subsec:basic}

The observations of the DUT constitute packets sent and received at certain
timestamps.
%
Aspects of the protocol specification can be monitored using state machines
to validate the observations of the DUT.
%
In this section, we formally define these concepts.
%
More specifically, we use \textit{timed automata}~\cite{alur1994theory} to model
the expected behaviors of the DUT with correct protocol implementation.
%
A timed automata is a finite state machine with timing constraints on the
transitions: each transition can optionally start one or more timers, which can
later be used to assert certain events should be seen before or after the time
out event.
%
We use the term \textit{timed automata} and \textit{state machine}
interchangeably hereafter.

The input alphabet of the state machine is the finite set of all valid packets defined
by the protocol, denoted as $\mathbb{P}$. 
%
A packet is a binary string of finite number of bits, encoding interesting
protocol attributes such as \texttt{src}, \texttt{dest}, \texttt{type},
\texttt{flags}, and physical layer information, such as \texttt{channel},
\texttt{modulation}, etc.

Next, we define the \textit{observation} that corresponds to time-ordered
sequence of packets.
\begin{definition}
  \sloppy{%
    A \textit{packet trace} is a finite sequence of $(timestamp, packet)$ tuple:%
  }%
  \begin{align*}
    [(t_1, p_1), (t_2, p_2),\ldots,(t_n, p_n)]
  \end{align*}%
  where $t_i \in \mathbb{Z}^+$ is the \textit{discrete} timestamp and $p_i$ is the packet
  observed at time $t_i$. The timestamps are strictly monotonically increasing,
  $t_i < t_{i+1}$ for $1 \le i < n$.
\end{definition}%

In addition to timestamp monotonicity, we also require that adjacent packets do
not overlap in time, $t_{i+1}-t_i > \text{\texttt{airtime}}(p_i)$ for $1 \le i
< n$, where \texttt{airtime()} calculates the time taken to transmit a packet
based on its size and modulation scheme.

We now formally define the monitor state machine that models the expected
behavior of the DUT and serves as the protocol specification.

\begin{definition}
  A protocol monitor state machine $S$ is an 6-tuple $\{\Sigma, \mathbb{S}, s_0, C, E, G\}$, where:
  \begin{itemize}
    \item $\Sigma = \mathbb{P}$ is the finite input alphabet.
    \item $\mathbb{S}$ is a non-empty, finite set of states. $s_0 \in
      \mathbb{S}$ is the initial state.
    \item $C$ is the set of clock variables. A \textit{clock variable} can be
      reset along any state transitions. At any instant, reading a clock
      variable returns the time elapsed since last time it was reset.
    \item $G$ is the set of clock constraints defined inductively by
      \begin{align*}
        g := true\ |\ c \le T\ |\ c \ge T\ |\ \neg g\ |\ g_1 \land g_2
      \end{align*}%
      where $c \in C$ is a clock variable and $T$ is a constant. Note that an
      transition can choose not to use clock guard by setting $g$ to be $true$.
    \item $E \subseteq \mathbb{S} \times \mathbb{S} \times \Sigma \times  G \times \mathscr{P}(C)$
      gives the set of transitions. $\langle s_i, s_j, p, g, C'\rangle \in E$
      represents that if the monitor is in state $s_i$, given the input tuple
      $(t, p)$ such that the clock variables satisfies the guard $g$, the
      monitor transits to state $s_j$ and reset the clocks in $C'$ to 0.
  \end{itemize}
  \label{def:sm}
\end{definition}

A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
the state machine at time $t_i$.
%
The monitor rejects a trace $Tr$ if there exists a prefix of $Tr$ such that all
states reachable after consuming the prefix have no valid transitions for the
next $(t, p)$ input.
%
Note that the input to the state machine, time and packet, are all
observable events in the communication medium.
%
Validating the internal events of the DUT is beyond the
scope of this paper.

Next, we give an example monitor state machine for the transmitter in 802.11
data transmission protocol.
%
Without loss of generality, we assume the DUT only
retransmits the packet at most once if the first transmission failed.
%
The state machine can be easily adapted to accommodate multiple retransmissions.

There are only three types of packets to consider in this case: $\it{DATA}_i$ denotes
a packet sent by the transmitter with sequence number $i$, $\it{DATA}'_i$ denotes the
retransmission of $\it{DATA}_i$, and $Ack$ is the acknowledgment packet sent by the
receiver.
%
The sequence number is within the range of $[0, N)$.

The monitor state machine is illustrated in Fig.~\ref{fig:dot11_tx_ta}, and
the main components are:

\begin{itemize}
  \item $\Sigma = \{\it{DATA}_i, \it{DATA}'_i, Ack\ |\ 0 \le i < N\}$.
    %
  \item Clock variables $C = \{c\}$. The only clock variable $c$ is
    used for acknowledgment time out.
    %
  \item Guard constraints $G = \{ c \le T_o, c > T_o, T_o < c \le T_m\}$.
    $T_o$ is the acknowledgment time out value, and $T_m >
    T_o$ is the maximum delay allowed before the retransmission packet gets
    sent. $T_o$ can be arbitrary large but not infinity in order to check the
    liveness of the DUT.
\end{itemize}

To succinctly present the state machine, we use an internal variable $i$ to keep
track of the next sequence number.
%
One can easily eliminate this variable and
use only different states for different sequence numbers.
%
Therefore, this state machine is consistent with Definition~\ref{def:sm}.

The sequence number $i$ is initialized to 0 when the state machine is
initialized.
%
At state $s_0$, the monitor expects to see a $\it{DATA}_i$ packet.
%
It transits to $s_1$ when such event happens and resets acknowledgment timer $c$ along the transition.
%
At state $s_1$, two valid events can happen: either the DUT receives $Ack$ packet
within time $T_o$ ($s_1\rightarrow s_0$), or the DUT sends $\it{DATA}'_i$ after $T_o$
but before $T_m$ ($s_1 \rightarrow s_2$).
%
Similarly at state $s_2$, either the DUT receives $Ack$ within time $T_o$ ($s_2
\rightarrow s_0$), or otherwise the DUT concludes that the previous transmission
failed and continues to transmit the next packet $s_2\rightarrow s_1$.

The monitor state machine defines a \textit{timed language} $L$ which consists
all valid packet traces that can be observed by the DUT.
%
We now give the
definition of protocol \textit{compliance} and \textit{violation}.

\begin{definition}
  Suppose $\mathbb{T}$ is the set of all possible packet traces collected from
  DUT, and $S$ is the state machine specified by the protocol. The DUT
  \textit{violates} the protocol specification if there exists an
  packet trace $Tr \in \mathbb{T}$ such that $S$ rejects $Tr$.
  Otherwise, the DUT is \textit{compliant} with the specification.
\end{definition}

The focus of this paper is to determine whether a \textit{given} $Tr$ is
evidence of a violation.
%
We acknowledge that determining \textit{compliance} is a more challenging
problem, as it requires enumerating every possible trace in $\mathbb{T}$, which
is probably infinite.

### 2017-02-24 framework/heuristic.tex
\it{DATA}

@@
 \begin{align*}
-  Pr_{ds} &= \frac{\parallel\{DATA'_0\}\parallel}{\parallel\{DATA_0, DATA'_0\}\parallel} = 0.5\\
+  Pr_{ds} &= \frac{\parallel\{\it{DATA}'_0\}\parallel}{\parallel\{\it{DATA}_0, \it{DATA}'_0\}\parallel} = 0.5\\
   Pr_{es} &= \frac{\parallel\emptyset\parallel}{\parallel\{Ack, Ack\}\parallel} = 0\\

### 2017-02-24 motivation.tex
title page format
\it{DATA}

@@
 
-\begin{figure}[H]
-  \vspace*{-3mm}
+\begin{figure}[t!]
   \centering
@@
   \label{fig:dot11_tx_ta}
-  \vspace*{-3mm}
 \end{figure}
@@
 802.11 (also known as \wifi{}) transmitter (DUT) state machine shown in
-Fig.~\ref{fig:dot11_tx_ta}. After the DUT sends $DATA_i$---a data packet with
+Fig.~\ref{fig:dot11_tx_ta}. After the DUT sends $\it{DATA}_i$---a data packet with
 sequence number $i$ ($s_0\rightarrow s_1)$, it starts a timer and waits for the
 acknowledgment packet---$Ack$. The DUT either receives $Ack$ within time $T_o$
-($s_1\rightarrow s_0$), or it sends $DATA'_i$---retransmission of $DATA_i$
+($s_1\rightarrow s_0$), or it sends $\it{DATA}'_i$---retransmission of $\it{DATA}_i$
 ($s_1\rightarrow s_2$). Similarly, the DUT either receives the $Ack$ within $T_o$
@@
 Consider a correct packet exchange sequence shown in
-Fig.~\ref{fig:sniffer_in_middle}. The DUT first sends $DATA_0$.  Suppose the
-receiver receives $DATA_0$ and sends the $Ack$ which the DUT does not receive.
-Eventually the DUT's timer fires and it sends $DATA'_0$.  This time the
-$DATA'_0$ reaches receiver and the DUT also receives the $Ack$.
+Fig.~\ref{fig:sniffer_in_middle}. The DUT first sends $\it{DATA}_0$.  Suppose the
+receiver receives $\it{DATA}_0$ and sends the $Ack$ which the DUT does not receive.
+Eventually the DUT's timer fires and it sends $\it{DATA}'_0$.  This time the
+$\it{DATA}'_0$ reaches receiver and the DUT also receives the $Ack$.
 
@@
 \textit{overhears} the first $Ack$ packet, a validation \textit{uncertainty}
-arises when the sniffer sees the $DATA'_0$: was the previous $Ack$ missed by the
+arises when the sniffer sees the $\it{DATA}'_0$: was the previous $Ack$ missed by the
 DUT or is there a bug in DUT which causes it to retransmit even after receiving
@@
 Similarly, consider the second possible sniffer trace $Tr_2$
-where both the $DATA'_0$ and $Ack$ packets were missed by the sniffer. During
-this period of time, it appears the DUT neither receives $Ack$ for $DATA_0$ nor
-sends $DATA'_0$. Again, without any additional information it is impossible to
+where both the $\it{DATA}'_0$ and $Ack$ packets were missed by the sniffer. During
+this period of time, it appears the DUT neither receives $Ack$ for $\it{DATA}_0$ nor
+sends $\it{DATA}'_0$. Again, without any additional information it is impossible to
 disambiguate between the sniffer missing certain packets and a bug in DUT's

### 2017-02-24 case/xbox.tex
xbox and conclusion

@@
   \end{tabular}
-  \caption{\textbf{Validation Results on Traces from the Gaming Wireless Protocol}}
+  \caption{\textbf{Validation Results on Traces from the Gaming Wireless Protocol.}}
   \label{tab:xbox}
@@
 Some of the implementation bugs we found related to the sequence number
-management, station scheduling, and the changing of modulation rates adaptation.
+management, station scheduling during power saving mode, and modulation rates
+adaptation.
+
 Table~\ref{tab:xbox} summarizes the validation result. Note that if we disable
 the augmented transitions, each trace will be flagged as violation because of
-the missing packets, thereby reducing the usability of the tool.
-
-The bugs related to packet sequence number and station scheduler were detected
-in about half the traces, and the bug related to the rate control algorithm was
-detected in only a few traces. This is because the previous two aspects are
-essential in all protocol operations, while the bugs related to rate control
-only manifest themselves under certain link conditions.
+the missing packets, thereby reducing the usability of the tool. We also note
+that some bugs manifest more often than others. For instance, the bugs related
+to packet sequence number and station scheduler were detected in about half the
+traces, and the bug related to the rate control algorithm was detected in only a
+few traces. This is because the previous two aspects are essential in all
+protocol operations, while the bugs related to rate control only manifest
+themselves under certain link conditions.
 
@@
 obtained because of the lack of automatized validating tools. Finally, the bug
-related to rate control has been filed as a bug report.
+related to rate control has been filed as a bug report. All reported bugs were
+fixed before the next release the product.

### 2017-02-24 conclusion.tex
xbox and conclusion

@@
 
-\textbf{State Machine Generation.} We manually translated the protocols studied
-in this paper into monitor state machines based on the source code, comments and
-documentation. The process is time-consuming and error-prone. A more scalable
-approach would be taking the protocol specification written in certain formal
-language, and automatically translate such specification into state machines
-that can be used for verification process.
+\textbf{Automated State Machine Construction.} We manually constructed the
+protocol monitor state machines for the protocols studied in this paper based on
+the source code, comments and documentation. The process involves lots of labor
+effort and is time-consuming. A potential alternative is to automatically learn
+the sketch of the monitor state machines from the sniffer traces. Domain
+knowledge can then be leveraged to improve the sketch state machines.

### 2017-03-02 case/tx.tex
xbox and conclusion
all-in-one

@@
     \centering
-    \includegraphics[width=\textwidth]{./figures/scripts/MutationDUTJaccard3DFigure_0_10.pdf}
+    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_10.pdf}
     % \caption{$Pr_{ed} \in [0.05, 0.1, 0.15]$}
@@
     \centering
-    \includegraphics[width=\textwidth]{./figures/scripts/MutationDUTJaccard3DFigure_0_30.pdf}
+    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_30.pdf}
     \caption{$0.2 \le Pr_{ed} \le 0.35$}
@@
     \centering
-    \includegraphics[width=\textwidth]{./figures/scripts/MutationDUTJaccard3DFigure_0_50.pdf}
+    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_50.pdf}
     \caption{$0.4 \le Pr_{ed} \le 0.5$}
@@
     \centering
-    \includegraphics[width=\textwidth]{./figures/scripts/StepCount3DFigure_0_10.pdf}
+    \includegraphics[width=\textwidth]{StepCount3DFigure_0_10.pdf}
     \caption{$Pr_{ed} \in [0.05, 0.10, 0.15]$}
@@
     \centering
-    \includegraphics[width=\textwidth]{./figures/scripts/StepCount3DFigure_0_30.pdf}
+    \includegraphics[width=\textwidth]{StepCount3DFigure_0_30.pdf}
     \caption{$Pr_{ed} \in [0.2, 0.25, 0.3, 0.35]$}
@@
     \centering
-    \includegraphics[width=\textwidth]{./figures/scripts/StepCount3DFigure_0_50.pdf}
+    \includegraphics[width=\textwidth]{StepCount3DFigure_0_50.pdf}
     \caption{$Pr_{ed} \in [0.40, 0.45, 0.50]$}
@@
 Similar to the case in Fig.~\ref{fig:mutation_dut}, $Pr_{ds}$ plays a dominant
-role in determine the searching cost. One interesting observation is that the
+role in determine the searching cost. One interesting observation for this
+particular protocol is that the
 search cost peaks when $Pr_{ds}$ is high while $Pr_{es}$ is low. In such loss
@@
   \begin{subfigure}{0.48\textwidth}
-    \includegraphics[width=\textwidth]{./figures/scripts/PrecisionFigure.pdf}
+    \includegraphics[width=\textwidth]{PrecisionFigure.pdf}
   \end{subfigure}
   \begin{subfigure}{0.48\textwidth}
-    \includegraphics[width=\textwidth]{./figures/scripts/RecallFigure.pdf}
+    \includegraphics[width=\textwidth]{RecallFigure.pdf}
   \end{subfigure}

### 2017-03-02 submitted/all-in-one.tex
all-in-one

\makeatletter{}\RequirePackage{fix-cm}
\documentclass[smallextended]{svjour3}

\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{comment}
\usepackage{mathrsfs}
\usepackage{balance}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[noend]{algpseudocode}
\usepackage[group-separator={,}]{siunitx}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{bm}
\usepackage[all]{hypcap}
\usepackage[absolute]{textpos}
\usepackage{wrapfig}
\usepackage[title]{appendix}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\LineComment}[1]{\State{\(\triangleright\) #1}}
\newcommand*\Let[2]{\State #1 := #2}

\smartqed

\usepackage{graphicx}
\graphicspath{{./figures/}}

\hypersetup{  bookmarks=true,
  unicode=true,
  pdftoolbar=true,
  pdfmenubar=true,
  pdffitwindow=true,
  pdfstartview={FitV},
  pdfnewwindow=true,
  colorlinks=false,
  pdfdisplaydoctitle=true,
  pdfborder={0 0 0}
}

\setlength{\TPHorizModule}{1in}
\setlength{\TPVertModule}{1in}
\textblockorigin{0.75in}{0.875in}

\setlist[itemize]{leftmargin=*,partopsep=5pt}
\setlist[enumerate]{leftmargin=*,partopsep=5pt}

\paperheight 11in
\paperwidth 8.5in

\usepackage{dcolumn}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\newcommand\nil{\textit{nil}}

\usepackage{tikz}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
      \node[shape=circle,draw,inner sep=0pt] (char) {#1};}}

\makeatletter
\newcommand*{\rome}[1]{\expandafter\@slowromancap\romannumeral #1@}
\makeatother

\makeatletter{}
\newcommand{\wifi}{Wi-Fi}
\newcommand{\codename}{\textsc{Verifier}}
\newcommand{\us}{$\mu$s}
\newcommand{\ns}{\textsc{NS-3}}

\newcommand{\shuvendu}[1]{{\bf [[SKL:{#1}]]}}

\def\thetitle{Wireless Protocol Validation Under Uncertainty}
\def\theauthors{}

\authorrunning{J. Shi, S. K. Lahiri, R. Chandra, G. Challen}

\journalname{Formal Methods in System Design}

\begin{document}

\title{  \thetitle\thanks{This work was published, in part, in Runtime Verification
    (RV) 2016~\cite{Shi2016}.}
}

\author{  Jinghao Shi$^1$ \and
  Shuvendu K. Lahiri$^2$ \and
  Ranveer Chandra$^2$ \and
  Geoffrey Challen$^1$
}

\institute{
  Jinghao Shi\\
  jinghaos@buffalo.edu\\\\
  Shuvendu K. Lahiri\\
  shuvendu@microsoft.com\\\\
  Ranveer Chandra\\
  ranveer@microsoft.com\\\\
  Geoffrey Challen\\
  challen@buffalo.edu\\\\
  $^1$University at Buffalo, Buffalo, NY 14260, USA\\
  $^2$Microsoft Research, Redmond, WA 98052, USA
}

\hypersetup{  pdfinfo={    Title={\thetitle},
    Author={\theauthors},
  }
}

\maketitle

\makeatletter{}\begin{abstract}
  Runtime validation of wireless protocol implementations cannot always employ
  direct instrumentation of the device under test (DUT). The DUT may not
  implement the required instrumentation, or the instrumentation may alter the
  DUT's behavior when enabled. Wireless sniffers can monitor the DUT's
  behavior without instrumentation, but they introduce new validation challenges.
  Losses caused by wireless propagation prevent sniffers
  from perfectly reconstructing the actual DUT packet trace. As a result,
  accurate validation requires distinguishing between
  specification deviations that represent implementation errors and those
  caused by sniffer uncertainty.

  \sloppy{
  We present a new approach enabling sniffer-based validation of wireless
  protocol implementations. Beginning with the original protocol monitor state
  machine, we automatically and completely encode sniffer uncertainty by
  selectively adding non-deterministic transitions. We characterize the
  NP-completeness of the resulting decision problem and provide an exhaustive
  algorithm for searching over all mutated traces. We also present
  practical protocol-oblivious heuristics for searching over the most likely
  mutated traces.  We have implemented our framework and show that it can
  accurately identify implementation errors in the face of uncertainty.
}

  \keywords{
    Runtime Verification \and
    Wireless Protocol \and
    Sniffer \and
  Uncertainty}
\end{abstract}

\makeatletter{}\section{Introduction}
\label{sec:intro}

Custom wireless protocols are often designed and deployed to meet the specific
performance and power needs of special-purpose wireless devices. Examples
include Google
Iris contact lenses~\cite{iris}, Xbox One wireless controllers~\cite{xbox}, and
Google Chromecast~\cite{chromecast}. Validating that device implementations work
correctly is critical to achieve the design goals of the wireless protocol and
also prevent bugs in shipped products~\cite{wifried,lollipop,surface}.

Runtime validation of the protocol implementations on such devices is
challenging because collecting traces from the device under test (DUT) is often
infeasible. The resource limitations of embedded or battery-powered devices may
cause them to not provide trace collecting capabilities. DUT may contain
proprietary hardware or firmware that hides the implementation details and
prevents testers from collecting traces through source code instrumentation.
Even when collecting trace directly from the DUT is possible, the overhead it causes
may alter the behavior of the DUT due to the observer
effect~\cite{mytkowicz2008observer}, threatening the validation results.

An attractive alternative is to use wireless
sniffers to record traffic generated by the DUT during testing.
Sniffers do not require direct access to the DUT or the need to alter its behavior.
However, due to the fundamentally unpredictable nature of wireless
communications, the packets captured by the sniffer will not exactly match
those received by the DUT.
The sniffer may miss packets that the DUT received, or receive packets that
the DUT missed.
This is true even when using multiple
sniffers~\cite{cheng2006jigsaw,mahajan2006analyzing,bahl2006enhancing}, a sniffer
with multiple antennas~\cite{omnipeek}, or in isolated wireless environments.

Since the sniffer trace may not perfectly match the actual trace, uncertainty
arises during protocol implementation validation.  For example, if the DUT fails
to respond correctly to a packet in the sniffer trace, it may either because the
DUT's implementation is incorrect, or the DUT did not actually receive the
packet, or the DUT's response was missed by the sniffer.  Whenever the DUT's
behavior does not match the specification, there are now two potential
explanations: either the DUT's implementation is wrong, or the sniffer trace is
inaccurate. Accurate validation requires  distinguishing between these two
causes.

We present a new technique than enables validation of protocol implementations using
wireless sniffers. Given a monitor state machine representing the protocol being
validated, we describe a systematic transformation that adds non-deterministic
transitions to incorporate uncertainty introduced by the sniffer. This
augmented validation state machine implicitly defines a set of
mutated traces, each satisfying the original state machine with a specific
likelihood. If the set is empty, the implementation definitely violates the
protocol. Searching over all the mutated traces is NP-complete, but the approach
can be made practical by applying protocol-oblivious heuristics that limit the
search to likely mutated traces.

Our paper makes the following contributions:
\begin{enumerate}
  \item To the best of our knowledge, we are the first to identify the
    uncertainty problem caused by sniffers in validating wireless protocol
    implementations.
  \item We formalize the problem using a nondeterministic state machine that
    systematically and completely encodes the uncertainty of the
    sniffer trace.
  \item We characterize the NP-completeness of the validation problem, and
    present two protocol-oblivious heuristics to prune the search space and make
    validation possible in practice.
  \item We implement the validation framework and evaluate it using the \ns{}
    network simulator~\cite{riley2010ns}. Our framework accurately identifies
    both synthetic and previously unknown violations in \ns{}'s implementations
    of the 802.11 and ARF protocols.  We also applied our framework to a
    commercial product under development and was able to found three latent bugs
    that were not observed previously.
\end{enumerate}

This paper is an extension of our work that appeared in RV 2016~\cite{Shi2016}.
We added the proof for Lemma~\ref{lem:mutation}, \ref{lem:sub} and
Theorem~\ref{the:equivalent}. We included new results that show the searching
cost (Fig.~\ref{fig:cost}), and details of the ARF protocol violations
(Section~\ref{subsec:arf}). Further, we reported the application of our
framework to a commercial product under development in
Section~\ref{subsec:xbox}.

\makeatletter{}\section{Background and Motivating Example}
\label{sec:motivation}

We encountered the uncertainty problem while testing the protocol implementation
of a popular wireless game controller. A custom wireless protocol
was designed to meet the low latency and low power consumption goals. As is
common industry practice, the protocol specification was then handed over to
wireless chipset vendors for implementation. However, neither implementation
details nor trace collection capabilities are provided in the shipped firmware
due to intellectual property constraints and device resource limitation. Hence
using sniffers to validate the protocol implementation is the only option.

We initially developed a tool to validate certain protocol properties over the
sniffer trace, yet often found unacceptable amount of false alarms due to
the incompleteness of the sniffer traces, making the tool virtually useless. It
was clear that we needed to account for sniffer uncertainty.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.6\textwidth]{dot11_tx_ta.pdf}
  \caption{\textbf{Monitor State Machine for 802.11 Transmitter.}}
  \label{fig:dot11_tx_ta}
\end{figure}

To better understand the incompleteness of sniffer trace, consider the IEEE
802.11 (also known as \wifi{}) transmitter (DUT) state machine shown in
Fig.~\ref{fig:dot11_tx_ta}. After the DUT sends $\it{DATA}_i$---a data packet with
sequence number $i$ ($s_0\rightarrow s_1)$, it starts a timer and waits for the
acknowledgment packet---$Ack$. The DUT either receives $Ack$ within time $T_o$
($s_1\rightarrow s_0$), or it sends $\it{DATA}'_i$---retransmission of $\it{DATA}_i$
($s_1\rightarrow s_2$). Similarly, the DUT either receives the $Ack$ within $T_o$
($s_2\rightarrow s_0$) or aborts transmission and moves on to next
packet\footnote{To represent the state machine succinctly, our example assumes
that the DUT retries at most once.} ($s_2\rightarrow s_1$).

Given a complete log of DUT's packet transmission and reception events, it is
trivial to feed such a log into the state machine in Fig.~\ref{fig:dot11_tx_ta}
and validate the correctness of DUT's protocol implementation. However, due to
DUT limitations we have described earlier, this complete log is not
available. As a result, we seek to validate the DUT implementation using
sniffers.

There are two fundamental properties in wireless communication that bring uncertainty to
sniffer's observation: packet loss and physical diversity. The sniffer could
either miss packets sent from or to the DUT due to packet loss, or overhear packets
that are sent to but missed by the DUT due to physical diversity.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.6\textwidth]{false_pos.pdf}
  \caption{\textbf{Uncertainty of Sniffer Observations.} $Tr_{OTA}$ is
    the chronological sequence of packets sent by the DUT and the receiver.
    $Tr_{DUT}$ is DUT's internal events. $Tr_1$ and $Tr_2$ are two examples of
    many possible sniffer traces.}
  \label{fig:sniffer_in_middle}
\end{figure}

Consider a correct packet exchange sequence shown in
Fig.~\ref{fig:sniffer_in_middle}. The DUT first sends $\it{DATA}_0$.  Suppose the
receiver receives $\it{DATA}_0$ and sends the $Ack$ which the DUT does not receive.
Eventually the DUT's timer fires and it sends $\it{DATA}'_0$.  This time the
$\it{DATA}'_0$ reaches receiver and the DUT also receives the $Ack$.

Now consider two possible traces that could have been overheard by a sniffer shown
in Fig.~\ref{fig:sniffer_in_middle}.
In first sniffer trace $Tr_1$ where the sniffer
\textit{overhears} the first $Ack$ packet, a validation \textit{uncertainty}
arises when the sniffer sees the $\it{DATA}'_0$: was the previous $Ack$ missed by the
DUT or is there a bug in DUT which causes it to retransmit even after receiving
the $Ack$?

Similarly, consider the second possible sniffer trace $Tr_2$
where both the $\it{DATA}'_0$ and $Ack$ packets were missed by the sniffer. During
this period of time, it appears the DUT neither receives $Ack$ for $\it{DATA}_0$ nor
sends $\it{DATA}'_0$. Again, without any additional information it is impossible to
disambiguate between the sniffer missing certain packets and a bug in DUT's
retransmission logic.

Informally, the question we set out to answer in this paper is: given the
protocol monitor state machine and the sniffer's observation with inherent
uncertainty, how to accurately validate that the DUT behaves as specified?

\makeatletter{}\section{Prerequisites and Problem Statement}
\label{sec:background}

\subsection{Packet, Trace and Monitor State Machine}
\label{subsec:basic}

The alphabet of the monitor state machine is the finite set of all valid packets
defined by the protocol, denoted as $\mathbb{P}$.  A packet is a binary string
of a finite number of bits, encoding interesting protocol attributes such as
\texttt{src}, \texttt{dest}, \texttt{type}, \texttt{flags}, and physical layer
information, such as \texttt{channel}, \texttt{modulation}, etc. The input of
the state machine then corresponds to a time-ordered sequence of packets.

\begin{definition}[Packet Trace]
  \sloppy{
  A \textit{packet trace} is a finite sequence of $(timestamp, packet)$ tuple:
  $[(t_1, p_1), (t_2, p_2),\ldots,(t_n, p_n)]$ where $t_i \in \mathbb{Z}^+$ is
  the \textit{discrete} timestamp and $p_i$ is the packet observed at time
  $t_i$. The timestamps are strictly monotonically increasing: $t_i < t_{i+1}$
  for $1 \le i < n$.
}
\end{definition}
In addition to timestamp monotonicity, we also require that adjacent packets do
not overlap in time, $t_{i+1}-t_i > \text{\texttt{airtime}}(p_i)$ for $1 \le i <
n$, where \texttt{airtime()} calculates the time taken to transmit a packet.
The timestamp represents the observer's local clock ticks, and need not to be
synchronized among devices.

We use \textit{timed automata}~\cite{alur1994theory} to model the expected
behaviors of the DUT. A timed automata is a finite state machine with timing
constraints on the transitions: each transition can optionally start one or more
timers, which can later be used to assert certain events should be seen before
or after the time out event. We refer the readers to~\cite{alur1994theory} for
more details about timed automata.

\begin{definition}[Monitor]
  \sloppy{    A protocol monitor state machine $S$ is a 7-tuple
    $\{\Sigma, \mathbb{S}, \mathbb{X}, s_0, C, E, G\}$, where:
  }
  \begin{itemize}
    \item $\Sigma = \mathbb{P}$ is the finite input alphabet.

    \item $\mathbb{S}$ is a non-empty, finite set of states. $s_0 \in
      \mathbb{S}$ is the initial state.

    \item $\mathbb{X}$ is the set of boolean variables. We use $v = \{x
      \leftarrow true/false\ |\ x \in \mathbb{X}\}$ to denote an assignment of
      the variables. Let $\mathbb{V}$ be the set of such values $v$.

    \item $C$ is the set of clock variables. A \textit{clock variable} can be
      reset along any state transitions. At any instant, reading a clock
      variable returns the time elapsed since last time it was reset.

    \item $G$ is the set of guard conditions defined inductively by
      \begin{align*}
        g := true\ |\ c \le T\ |\ c \ge T\ |\ x\ |\ \neg g\ |\ g_1 \land g_2
      \end{align*}      where $c \in C$ is a clock variable, $T$ is a constant, and $x$ is a
      variable in $\mathbb{X}$.  A transition can choose not to use guard
      conditions by setting $g$ to be $true$.

    \item $E \subseteq \mathbb{S} \times \mathbb{V} \times \mathbb{S} \times
      \mathbb{V} \times \Sigma \times  G \times \mathscr{P}(C)$ gives the set of
      transitions.\\ $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$
      represents that if the monitor is in state $s_i$ with variable assignments
      $v_i$, given the input tuple $(t, p)$ such that the guard $g$ is
      satisfied, the monitor can transition to a state $s_j$ with variable assignments
      $v_j$, and reset the clocks in $C'$ to 0.
  \end{itemize}
  \label{def:sm}
\end{definition}

A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
the state machine at time $t_i$. The monitor {\it rejects a trace} $Tr$ if there
exists a prefix of $Tr$ such that all states reachable after consuming the
prefix have no valid transitions for the next $(t, p)$ input.

As an example, the monitor state machine illustrated in
Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:

\begin{itemize}
  \item $\Sigma = \{\it{DATA}_i, \it{DATA}'_i, Ack\ |\ 0 \le i < N\}$.
  \item Clock variables $C = \{c\}$. The only clock variable $c$ is
    used for acknowledgment time out.
  \item $\mathbb{X} = \{i\}$, as a variable with ${\mathit log}(N) + 1$ bits to
    count from $0$ to $N$.  
  \item Guard constraints $G = \{ c \le T_o, c > T_o, T_o < c \le T_m\}$.
    $T_o$ is the acknowledgment time out value, and $T_m >
    T_o$ is the maximum delay allowed before the retransmission packet gets
    sent. $T_o$ can be arbitrary large but not infinity in order to check the
    liveness of the DUT.
\end{itemize}

The monitor state machine defines a \textit{timed language} $L$ which consists
of all valid packet traces that can be observed by the DUT.  We now give the
definition of protocol \textit{compliance} and \textit{violation}.

\begin{definition}[Violation and Compliance]
  Suppose $\mathbb{T}$ is the set of all possible packet traces collected from
  DUT, and $S$ is the state machine specified by the protocol. The DUT
  \textit{violates} the protocol specification if there exists an
  packet trace $Tr \in \mathbb{T}$ such that $S$ rejects $Tr$.
  Otherwise, the DUT is \textit{compliant} with the specification.
\end{definition}

The focus of this paper is to determine whether a \textit{given} sniffer trace
$Tr$ is evidence of a violation.

\subsection{Mutation Trace}
\label{subsec:mutation}

As shown in the motivation example in Fig.~\ref{fig:sniffer_in_middle}, a
sniffer trace may either miss packets that are present in DUT trace, or contain
extra packets that are missing in DUT trace. Note that in the latter case, those
extra packets must be all sent \textit{to} the DUT. This is because it is
impossible for the sniffer to overhear packets sent from the DUT that were not
actually sent by the DUT.

We formally capture this relationship with the definition of mutation trace.

\begin{definition}[Mutation Trace]
  \label{def:mutation}
  A packet trace $Tr'$ is a mutation of sniffer trace $Tr$ w.r.t a DUT if for
  all $(t, p) \in Tr \setminus Tr'$, $p.dest = DUT$, where $p.dest$ is the
  destination of packet $p$.
\end{definition}

By definition, either $Tr' \supseteq Tr$ (hence $Tr \setminus Tr' = \emptyset$),
or those extra packets in $Tr$ but not in $Tr'$ are all sent to the DUT. Note
that $Tr'$ may contain extra packets that are either sent to or received by
the DUT.

A mutation trace $Tr'$ represents a \textit{guess} of the corresponding DUT
packet trace given sniffer trace $Tr$.  In fact, the DUT packet trace must
be one of the mutation traces of the sniffer trace $Tr$.

\begin{lemma}
  Let $Tr_{DUT}$ and $Tr$ be the DUT and sniffer packet trace captured during
  the same protocol operation session, and $\mathcal{M}(Tr)$ be the set of
  mutation traces of $Tr$ with respect to DUT, then $Tr_{DUT} \in \mathcal{M}(Tr)$.
  \label{lem:mutation}
\end{lemma}\begin{proof}
  Let $\Delta = Tr \setminus Tr_{DUT}$ be the set of packets that are in $Tr$
  but not in $Tr_{DUT}$. Recall that it is not possible for the sniffer to
  observe packets sent from the DUT that the DUT did not send. Therefore,
  all packets in $\Delta$ are sent \textit{to} the DUT. That is, for
  all $(t, p) \in \Delta,\ p.dest = DUT$. By Definition~\ref{def:mutation},
  $Tr_{DUT} \in \mathcal{M}(Tr)$.
  \qed
\end{proof}

\subsection{Problem Statement}
\label{subsec:problem}

Lemma~\ref{lem:mutation} shows that $\mathcal{M}(Tr)$ is a \textit{complete} set
of guesses of the DUT packet trace. Therefore, the problem of validating DUT
implementation given a sniffer trace can be formally defined as follows:

\begin{problem}
  \label{prob:validation}
  VALIDATION\\
  \textbf{instance} A protocol monitor state machine $S$ and a sniffer trace $Tr$.\\
  \textbf{question} Does there exist a mutation trace $Tr'$ of $Tr$ that satisfies $S$?
\end{problem}

If the answer is no, a definite violation of the DUT implementation can be
claimed. Nevertheless, if the answer is yes, $S$ may still reject $Tr_{DUT}$.
In other words, the conclusion of the validation can either be
\textit{definitely wrong} or \textit{probably correct}, but not
\textit{definitely correct}.  This is the fundamental limitation caused by the
uncertainty of sniffer traces.

\makeatletter{}\section{Validation Framework}
\label{sec:framework}

\makeatletter{}\subsection{Augmented State Machine}
\label{subsec:augment}

To deal with the inherent uncertainty of sniffer traces, we propose to
systematically augment the original monitor state machine with non-deterministic
transitions to account for the difference between the sniffer and DUT traces.

Before formally defining the augmented state machine, we first use an example to
illustrate the basic idea. Fig.~\ref{fig:augment} shows the augmented state
machine for 802.11 transmitter state machine shown in
Fig.~\ref{fig:dot11_tx_ta}.  For each existing transition (e.g., $s_0\rightarrow
s_1$), we add an \textit{empty transition} with same clock guards and resetting
clocks.  This accounts for the possibility when such packet was observed by
the DUT but missed by the sniffer.  Additionally, for each transition triggered
by a \textit{receiving} packet (i.e., $p.dest = DUT$), such as $s_1\rightarrow
s_0$ and $s_2\rightarrow s_0$, we add a \textit{self transition} with the same
trigger packet and clock guards, but an empty set of resetting clocks and no
assignments to variables. This allows the state machine to make progress when
the sniffer missed such packets.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.7\textwidth]{dot11_tx_checker.pdf}
  \caption{\textbf{Augmented Monitor State Machine.} Augmented transitions are
  highlighted in bold face. $\overline{Pkt}$ means either $\epsilon$ or $Pkt$.}
  \label{fig:augment}
\end{figure}

There are two things to note. First, self transitions are added only for
packets sent \textit{to} the DUT, since the sniffer will not overhear packets
\textit{from} the DUT if they were not sent by the DUT. Second, no augmented
transitions are added for the packets that are sent to DUT yet are missed by both
the DUT and the sniffer, since such packets do not cause difference between the
DUT and sniffer traces.

The augmented state machine in Fig.~\ref{fig:augment} will accept the sniffer
packet traces $Tr_1$ and $Tr_2$ shown in Fig.~\ref{fig:sniffer_in_middle}.  For
instance, one accepting transition sequence on sniffer trace $Tr_1$ is
$s_0\rightarrow s_1 \rightarrow_s s_1\rightarrow s_2 \rightarrow s_0$, and the
sequence for $Tr_2$ is $s_0 \rightarrow s_1 \rightarrow_e s_2 \rightarrow s_0$,
where $\rightarrow$ is the transition from the original state machine,
$\rightarrow_e$ and $\rightarrow_s$ are the augmented empty and self transitions
respectively.

We now formally define the augmented state machine.

\begin{definition}[Augmented Monitor]
  \sloppy{
  An augmented state machine $S^+$ for a monitor state machine $S$ is a 7-tuple
  $\{\boldsymbol{\Sigma^+}, \mathbb{S}, \mathbb{X}, s_0, C, \boldsymbol{E^+},
  G\}$, where $\mathbb{S}, \mathbb{X}, s_0, C, G$ are the same as $S$.
  $\Sigma^+=\{\epsilon\} \cup \Sigma$ is the augmented input alphabet with the
  empty symbol, and $E^+ \supset E$ is the set of transitions, which includes:
}
  \begin{itemize}
    \item $E$: existing transitions (\textbf{Type-0}) in $S$.
    \item $E^+_1$: empty transitions (\textbf{Type-1}) for transitions in $E$.
    \item $E^+_2$: self transitions \textbf{(Type-2)} for transitions
      triggered by receiving packets.
  \end{itemize}
\end{definition}

\begin{figure}[t!]
\begin{algorithm}[H]
  \caption{Obtain Augmented Transitions $E^+$ from $E$}
  \label{alg:augment}
  \begin{algorithmic}[1]
    \Function{augment}{$E$}
    \Let{$E^+$}{$\emptyset$}
    \ForAll{$\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, p, g, C'\rangle\}$\Comment{Type-0}}
      \label{alg:augment:type0}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, \boldsymbol{\epsilon}, g, C'\rangle\}$\Comment{Type-1}}
      \label{alg:augment:type1}
      \If{$p.dest = DUT$}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, \boldsymbol{s_i},
      \boldsymbol{v_i}, p, g, \boldsymbol{\emptyset}\rangle\}$\Comment{Type-2}}
        \label{alg:augment:type2}
      \EndIf
    \EndFor
    \State \Return{$E^+$}
    \EndFunction
  \end{algorithmic}
\end{algorithm}
\end{figure}

Algorithm~\ref{alg:augment} describes the process of transforming $E$ into
$E^+$. In particular, Line~\ref{alg:augment:type0} adds existing transitions in
$E$ to $E^+$, while line~\ref{alg:augment:type1} and~\ref{alg:augment:type2} add
Type-1 and Type-2 transitions to $E^+$ respectively.  We have highlighted the
elements of the tuple that differ from the underlying Type-0 transition. Note
that in Type-2 transitions, both the state and the variables stay the same after
the transition.

With augmented state machine $S^+$, we can use Type-1 transitions to
non-deterministically infer packets missed by the sniffer, and use Type-2
transitions to consume extra packets captured by the sniffer but missed by the
DUT.

A accepting run of $S^+$ on sniffer trace $Tr$ yields a mutation trace $Tr'$
which represents one possibility of the DUT trace. Specifically, $Tr'$ can be
obtained by adding missing packets indicated by Type-1 transitions to $Tr$, and
removing extra packets indicated by Type-2 transitions from $Tr$

We show that the VALIDATION problem is equivalent to the
satisfiability problem of $Tr$ on $S^+$.

\begin{theorem}
  There exists a mutation trace $Tr' \in \mathcal{M}(Tr)$ that satisfies $S$ if
  and only if $Tr$ satisfies $S^+$.
 \label{the:equivalent}
\end{theorem}
\begin{proof}
  Assume $Tr$ satisfies $S^+$, and $P$ is a sequence of accepting transitions,
  we construct a mutation trace $Tr'$ using $P$ and show that $Tr'$ satisfies
  $S$.

  \sloppy{    Initially, let $Tr'=Tr$, then for each \textit{augmented} transition $\langle s_i,
  v_i, s_j, v_j, \sigma, g, C'\rangle \in P$:
  }
  \begin{itemize}
    \item If this is a Type-1 transition, add $(t, p)$ to $Tr'$, where $t$ is a
      timestamp that satisfies $g$ and $p$ is the missing packet.
    \item If this is a Type-2 transition, remove corresponding $(t, p)$ from
      $Tr'$.
  \end{itemize}
  It is obvious that $Tr'$ is a mutation trace of $Tr$, since only receiving
  packets are removed in the process.

  Now we show that there exists a accepting transition sequence $P'$ of $S^+$ on
  input $Tr'$ that does not contain augmented transitions.  In particular, $P'$
  can be obtained by substituting all Type-1 transitions with corresponding
  original transitions, and removing all Type-2 transition.  Since $P'$ does not
  contain augmented transitions, it is also an accepting transition sequence of
  $S$ on input $Tr'$, hence $Tr'$ satisfies $S$.

  On the other hand, assume $Tr' \in \mathcal{M}(Tr)$ and $Tr'$ satisfies $S$.
  Suppose $P'$ is the accepting transition sequences of $S$ on input $Tr'$.
  We first note that $P'$ is also the accepting transitions of $S^+$ on input
  $Tr'$, since $E \subset E^+$.

  We construct a accepting transition sequence $P$ of $S^+$ on input $Tr$ as
  follows.
  \begin{itemize}
    \item For each packet $p \in Tr' \setminus Tr$, substitute the transition
      $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle$ with the corresponding Type-1
      transition $\langle s_i, v_i, s_j, v_j, \epsilon, g, C'\rangle$.
    \item For each transition $\langle s_i, v_i, s_j, v_j, \sigma, g, C'\rangle$
      followed by packet $p \in Tr\setminus Tr'$, add a Type-2 self
      transition $\langle s_j, v_j, s_j, v_j, p, g, \emptyset\rangle$. This is
      possible since $Tr'$ is a mutation trace of $Tr$, thus  for all $p \in Tr'
      \setminus Tr$, $p.src \ne DUT$.
  \end{itemize}
  Therefore, $Tr$ satisfies $S^+$.
  \qed
\end{proof}

By Theorem~\ref{the:equivalent}, the inherent uncertainty of the sniffer traces
is explicitly represented by the augmented transitions, and can be
systematically explored using the well established theory of state machine.

\begin{comment}
One immediate observation can be drawn from Theorem~\ref{the:equivalent} by
contradiction.

\begin{corollary}
  If $S^+$ rejects $Tr$, then $S$ rejects $Tr_{DUT}$.
  \label{cor:false_pos}
\end{corollary}
In the context of validation where we raise a violation alarm when $S^+$
rejects $Tr$, Corollary~\ref{cor:false_pos} provides a {\it sufficient} condition for reporting incorrect traces. 
However, when $S^+$ accepts $Tr$, $S$ could still reject $Tr_{DUT}$.
In other words, the conclusion of the validation can either be \textit{definitely
wrong} or \textit{possibly correct}, but not \textit{definitely correct}.
This is the fundamental limitation caused by the uncertainty of sniffer traces.
\end{comment}

\makeatletter{}\subsection{Problem Hardness}
\label{subsec:hard}

In this section, we show that the VALIDATION problem is NP-complete. In
fact, the problem is still NP-complete even with only one type of augmented
transitions.

Recall that Type-1 transitions are added because the sniffer may miss packets.
Suppose an imaginary sniffer that is able to capture \textit{every} packet ever
transmitted, then only Type-2 transitions are needed since the sniffer may still
overhear packets sent to the DUT. Similarly, suppose another special sniffer
that would not overhear any packets sent to the DUT, then only Type-1
transitions are needed to infer missing packets.

We refer the augmented state machine that only has Type-0 and Type-1
transitions as $S^+_1$, and the augmented state machine that only has Type-0 and
Type-2 transitions as $S^+_2$. And we show that each subproblem of determining
trace satisfiability is NP-complete.

\begin{problem}
  VALIDATION-1\\
  Given that $Tr\setminus Tr_{DUT}=\emptyset$ (sniffer does not overhear
  packets).\\
  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
  \textbf{question} Does $S^+_1$ accept $Tr$?
\end{problem}

\begin{problem}
  VALIDATION-2\\
  Given that $Tr_{DUT} \subseteq Tr$ (sniffer does not missing packets).\\
  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
  \textbf{question} Does $S^+_2$ accept $Tr$?
\end{problem}

\begin{lemma}
  \label{lem:sub}
  Both VALIDATION-1 and VALIDATION-2 are NP-complete.
\end{lemma}
\begin{proof}
  First, note that the length of mutation trace $Tr'$ is polynomial to the
  length of $Tr$ because of the discrete time stamp and non-overlapping packets
  assumption.
  Therefore, given a state transition sequence as witness, it can be verified in
  polynomial time whether or not it is an accepting transition sequence, hence
  both VALIDATION-1 and VALIDATION-2 are in NP.

  Next, we show how the SAT problem can be reduced to either one of the two
  problems.
  Consider an instance of SAT problem of a propositional formula $F$ with $n$
  variables $x_0,x_1,\ldots, x_{n-1}$, we construct a corresponding protocol and
  its monitor state machine as follows.

  The protocol involves two devices: the DUT (transmitter) and the endpoint
  (receiver).
  The DUT shall send a series of packets, $pkt_0, pkt_1,\ldots, pkt_{n-1}$.
  For each $pkt_i$, if the DUT receives the
  acknowledgment packet $ack_i$ from the endpoint, it sets boolean variable
  $x_i$ to be \texttt{true}.
  Otherwise $x_i$ remains to be \texttt{false}.
  After $n$ rounds, the DUT evaluate the formula $F$ using the assignments and
  sends a special packet, $pkt_{true}$, if $F$ is \texttt{true}.
  One round of the protocol operation can be completed in polynomial time since
  any witness of $F$ can be evaluated in polynomial time.

  \begin{figure}[t!]
    \centering
    \includegraphics[width=0.7\textwidth]{sat_sm.pdf}
    \caption{\textbf{Monitor State Machine for SAT Problem.}}
    \label{fig:sat}
  \end{figure}

  The protocol monitor state machine $S$ is shown in Fig.~\ref{fig:sat}.
  Initially, all $x_i$ is set to \texttt{false}.
  At state $s_0$, the DUT shall transmit $pkt_i$ within a unit time, transit to
  $s_1$ and reset the clock along the transition.
  At state $s_1$, either the DUT receives the $ack_i$ packet and
  set $x_i$ to be \texttt{true} ($s_1 \rightarrow s_0$), or the DUT continues to
  transmit the next packet $pkt_{i+1}$.
  After $n$ rounds, the state machine is $s_0$ or $s_1$ depending on whether
  $ack_{n-1}$ is received by the DUT.
  In either case, the DUT shall evaluate $F$ and transmit $pkt_{true}$ if $F$ is
  \texttt{true}.

  \sloppy{    Consider a sniffer trace $Tr_1=\{(0, pkt_0), (2, pkt_1), (4, pkt_2),\ldots,
  (2(n-1), pkt_{n-1}), (2n, pkt_{true})\}$.  That is, the sniffer only captures
  all $pkt_i$ plus the final $pkt_{true}$, but none of $ack_i$.  It is easy to see
  that $F$ is satisfiable if $S_1^+$ accepts $Tr_1$.  In particular, a successful
  run of $S_1^+$ on $Tr_1$ would have to guess, for each $pkt_i$, whether the
  Type-1 empty transitions should be used to infer the missing $ack_i$ packet,
  such that $F$ is \textit{true} at the end.  Note that for $Tr_1$, no Type-2 self
  transitions can be used since all packets in $Tr_1$ are sent from the DUT.
  Therefore, the SAT problem of $F$ can be reduced to the VALIDATION-1 problem of
  $S^+_1$ on sniffer trace $Tr_1$.
  }

  \sloppy{    On the other hand, consider another sniffer trace $Tr_2 =\{(0, pkt_0), (1,
    ack_0), (2, pkt_1), (3, ack_1),\ldots, (2n-2, pkt_{n-1}), (2n-1, ack_{n-1}),\\ (2n,
  pkt_{true}\}$. That is, the sniffer captures all $n$ pair of $pkt_i$ and $ack_i$
  packets and the final $pkt_{true}$ packet.  Similar to $Tr_1$, $F$ is
  satisfiable if $S^+_2$ accepts $Tr_2$. A successful transition sequence of
  $S^+_2$ on $Tr_2$ must decide for each $ack_i$ packet, whether Type-2 self
  transitions should be used, so that $F$ can be evaluated as true at the end.
  Therefore, the SAT problem of $F$ can also be reduced to the VALIDATION-2
  problem of $S^+_2$ on sniffer trace $Tr_2$.
}

Since SAT is known to be NP-complete, both the VALIDATION-1 and the
VALIDATION-2 problem are also NP-complete.  \qed
\end{proof}

The hardness statement of the general VALIDATION problem naturally follows
Lemma~\ref{lem:sub}.

\begin{theorem}
  VALIDATION is NP-complete.
\end{theorem}

\makeatletter{}\subsection{Searching Strategies}
\label{subsec:search}

In this section, we present an \textit{exhaustive} search algorithm of the
accepting transition sequence of $S^+$ on sniffer trace $Tr$. It is guaranteed to
yield an accepting sequence if there exists one, thus is exhaustive.  In the next
sections, we present heuristics to limit the search to accepting sequences of
$S^+$ that require relatively fewer transitions from $E^+_1 \cup E^+_2$.  Due to
the NP-completeness of the problem, this also makes the algorithm meaningful in
practice.

\begin{figure}[t!]
  \begin{minipage}{\textwidth}
    \begin{algorithm}[H]
      \caption{Exhaustive search algorithm of $S^+$ on $Tr$.}
      \label{alg:search}

      \begin{algorithmic}[1]
        \Function{search}{S, Tr}
          \Let{$S^+$}{\Call{augment}{S}}
          \State\Return{\Call{extend}{[], Tr, $S^+.s_0$}}
        \EndFunction
        \Function{extend}{prefix, p::suffix, $s$}
          \If{not \Call{likely}{prefix}}
            \label{alg:search:stop}
            {\Return{\nil}}\footnote{This check should be ignored in the exhaustive algorithm.}
          \EndIf
          \For{i $\in$ [0, 1, 2]}
          \label{alg:search:order}
          \Let{mutation}{EXTEND-i(prefix, p::suffix, $s$)}
            \If{mutation $\ne$ \nil}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State\Return{\nil}
        \EndFunction
        \Function{extend-0}{prefix, p::suffix, $s$}
          \For{$\langle s, s', p\rangle\footnote{$\langle s, s', p \rangle$ is
              short for $\langle s, *, s', *, p, *, *\rangle$} \in E$}
            \If{suffix = \nil}
              {\Return{prefix@p}}
              \label{alg:search:terminate_0}
            \EndIf
            \Let{mutation}{\Call{extend}{prefix@p, suffix, $s'$}}
            \label{alg:search:append}
            \If{mutation $\ne$ \nil}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State{\Return{\nil}}
        \EndFunction
        \Function{extend-1}{prefix, p::suffix, $s$}
          \ForAll{$\langle s, s', q\rangle \in E^+_1$}
            \If{q.time $>$ p.time}
              \label{alg:search:type_1_limit}
              \State{\textbf{continue}}
            \EndIf
            \Let{mutation}{\Call{extend}{prefix@q, p::suffix, $s'$}}
            \label{alg:search:guess}
            \If{mutation $\ne \nil$}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State{\Return{\nil}}
        \EndFunction
        \Function{extend-2}{prefix, p::suffix, $s$}
          \ForAll{$\langle s, s, p \rangle \in E^+_2$}
            \If{suffix = \nil}
              {\Return{prefix}}
              \label{alg:search:terminate_2}
            \EndIf
            \Let{mutation}{\Call{extend}{prefix, suffix, $s$}}
            \label{alg:search:drop}
            \If{mutation $\ne$ \nil}
              {\Return{mutation}}
            \EndIf
          \EndFor
          \State{\Return{\nil}}
        \EndFunction
      \end{algorithmic}
    \end{algorithm}
    \renewcommand\footnoterule{}
  \end{minipage}
\end{figure}

The main routines of the algorithm are shown in Algorithm~\ref{alg:search}.  In
the top level \texttt{SEARCH} routine, we first obtain the augmented state
machine $S^+$, and then call the recursive \texttt{EXTEND} function with an empty
prefix, the sniffer trace, and the $S^+$'s initial state. 
In the \texttt{EXTEND} function, we try to consume the first packet in the
remaining trace using either Type-0, Type-1 or Type-2 transition.
Note that we always try to use Type-0 transitions before other two augmented
transitions (line~\ref{alg:search:order}).
This ensures the first found mutation trace will have the most number of Type-0
transitions among all possible mutation traces.
Intuitively, this means the search algorithm tries to utilize the sniffer's
observation as much as possible before being forced to make assumptions.

Each of the extend functions either returns the mutation trace $Tr'$, or
\textit{\nil} if the search fails.
In both \texttt{EXTEND-0} and
\texttt{EXTEND-2} function, if there is a valid transition, we try to consume
the next packet either by appending it to the prefix
(line~\ref{alg:search:append}) or dropping it (line~\ref{alg:search:drop}).
While in \texttt{EXTEND-1}, we guess a missing packet without consuming the next
real packet (line~\ref{alg:search:guess}).
Note that since only Type-0 and Type-2 consume packets, the recursion terminates
if there is a valid Type-0 or Type-2 transition for the last packet
(line~\ref{alg:search:terminate_0} and line~\ref{alg:search:terminate_2}).

It is not hard to see that Algorithm~\ref{alg:search} terminates on any sniffer
traces. Each node in the transition tree only has finite number of possible next
steps, and the depth of Type-1 transitions is limited by the time available
before the next packet (line~\ref{alg:search:type_1_limit}).

\makeatletter{}\subsection{Pruning Heuristics}
\label{subsec:heuristic}

In the face of uncertainty between a possible
protocol violation and sniffer imperfection, augmented transitions provide the
ability to blame the latter. The exhaustive nature of
Algorithm~\ref{alg:search} means that it always tries to blame sniffer
imperfection whenever possible, making it reluctant to report true
violations.

Inspired by the \textit{directed model checking}~\cite{edelkamp2008survey}
technique which is to mitigate the state explosion problem, we propose to enforce extra
constraints on the mutation trace to restrict the search to only mutation traces
with high likelihood. The modified \texttt{EXTEND} function checks certain
likelihood constraints on the prefix of the mutation trace before continuing
(line~\ref{alg:search:stop}), and stops the current search branch immediately if
the prefix seems \textit{unlikely}.  Because of the recursive nature of the
algorithm, other branches which may have a higher likelihood can then be
explored.

The strictness of the likelihood constraint represents a trade-off between
precision and recall of validation. The more strict the constraints are, the
more false positive violations will potentially be reported, hence the lower the
precision yet higher recall. On the contrary, the more tractable the
constraints are, the more tolerant the search is to sniffer imperfection, hence
the more likely that it will report true violations, thus higher precision but
lower recall.

The exact forms of the constraints may depend on many factors, such as the
nature of the protocol, properties of the sniffer, or domain knowledge.  Next,
we propose two \textit{protocol oblivious} heuristics based on the sniffer loss
probabilities and general protocol operations. Both heuristic contains
parameters that can be fine tuned in practice.

\subsubsection{$\mathit{NumMissing}(d, l, k)$}

This heuristic states that the number of missing packets from device $d$ in any
sub mutation traces of length $l$ shall not exceed $k$ ($k \le l$).  The sliding
window of size $l$ serves two purposes. First, $l$ should be large enough for
the calculated packet loss ratio to be statistically meaningful.  Second, it
ensures that the packet losses are evenly distributed among the entire packet
trace.

The intuition behind this heuristic is that the sniffer's empirical packet loss
probability can usually be measured before validation. Therefore, the likelihood
that the sniffer misses more packets than prior measured loss ratio is quite
low. The value of $l$ and $k$ can then be configured such that $k/l$ is marginally
larger than the measured ratio.

\begin{comment}
In addition, we note that for a fixed $l$, a single threshold $k$ may not work
for traces collected by sniffers with different packet loss probabilities.
Intuitively, sniffers with high loss probabilities require a larger threshold.
Therefore, we perform a \textit{stratified} search with an increasing sequence
of thresholds $\{k_1, k_2,\ldots, k_n\}$ ($k_i < k_{i+1}$ for $1 \le i < n$).
At round $i$, the search is restricted by constraint $NumMissing(d, l, k_i)$.
If a mutation trace is found, the search is completed.  Otherwise, we increase
the round number and repeat the search process.  In this way, this heuristic can
automatically adapt to the underlying sniffer packet loss probabilities. And a
violation is declared when no mutation trace can be found with $NumMissing(d, l,
k_n)$.
\end{comment}

\subsubsection{$\mathit{GoBack}(k)$}

This heuristic states that the search should only
backtrack at most $k$ steps when the search gets stuck using only $E$.
The motivation is that many protocols operate as a sequence of independent
transactions, and the uncertainty of previous transactions often do not affect
the next transaction.
For instance, in 802.11 packet transmission protocol, each packet exchange,
include the original, retransmission and acknowledgment packets, constitute a
transaction.
And the retransmission status of previous packets has no effect on the packets
with subsequent sequence numbers, hence need not be explored when resolving the
uncertainty of the packets with new sequence numbers.  Note that we do not
require the protocol to specify an exact transaction boundary, but only need $k$
to be sufficiently large to cover a transaction.

\begin{comment}
\subsection{Confidence of Correctness}

A run of $S^+$ on sniffer packet trace $Tr$ may yield two conclusions about the
DUT implementation: either ``definitely wrong'' when $S^+$ rejects $Tr$, or
``possibly correct'' when $S^+$ accepts $Tr$. We now quantify the
\textit{confidence} of the implementation correctness in the later case.

Note that the augmented transitions in $S^+$ implicitly embed the assumptions
about packet losses. In particular, Type-1 transitions assume the sniffer
missed certain packets, while Type-2 transitions assume the sniffer
overheard packets that were missed by the DUT. Intuitively, the portion of such
assumptions (hence packet loss \textit{ratios}) in an accepting transition
sequence should match with the true packet loss \textit{probabilities}.
Otherwise, the correctness confidence, despite $S^+$ accepts $Tr$, is low.

Suppose $Tr$ is the sniffer trace and $Tr'$ is its mutation that satisfies $S$.
We can calculate the three packet loss ratios in our model shown in
Figure~\ref{fig:prob}.

\begin{align}
  &Pr_{ds} = \frac{\parallel \{(t,p) \in Tr'\setminus Tr \land p.src = DUT\}\parallel}
  {\parallel \{(t,p) \in Tr' \cup Tr \land p.src = DUT\}\parallel}\\
  &Pr_{es} = \frac{\parallel \{(t,p) \in Tr'\setminus Tr \land p.src = EP\}\parallel}
  {\parallel \{(t,p) \in Tr' \cup Tr \land p.src = EP\}\parallel}
\end{align}where $Tr'\setminus Tr$ is the set of inferred packets that were missed by the
sniffer (Type-1 transitions in $S^+$).

Note that we can not directly calculate the packet loss ratio from Endpoint to
DUT, since there might be packets that were missed by both the sniffer and the
DUT. We can, however, approximate $Pr_{ed}$ using conditional packet loss
probability.

\begin{align}
  \nonumber
  &(1-Pr_{es})\times Pr_{ed} =\\
  &\frac{\parallel \{(t,p) \in Tr\setminus Tr' \land p.src = EP \land p.dest = DUT\}\parallel}
  {\parallel \{(t,p) \in Tr \cup Tr' \land p.src = EP \land p.dest = DUT\}\parallel}
\end{align}The LHS is the probability that a packet sent from the Endpoint to the DUT was
missed by the DUT yet captured by the sniffer. And the RHS is the ratio of
such packets in the mutation trace.

Consider the 802.11 data transmission protocol and the example trace snippet
shown in Figure~\ref{fig:loss}, the packet loss ratios are:
\begin{align*}
  Pr_{ds} &= \frac{\parallel\{\it{DATA}'_0\}\parallel}{\parallel\{\it{DATA}_0, \it{DATA}'_0\}\parallel} = 0.5\\
  Pr_{es} &= \frac{\parallel\emptyset\parallel}{\parallel\{Ack, Ack\}\parallel} = 0\\
  Pr_{ed} &= \frac{\parallel\{Ack\}\parallel}{\parallel\{Ack, Ack\}\parallel\times (1-Pr_{es})} = 0.5
\end{align*}

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.4\textwidth]{loss.pdf}
  \caption{\textbf{Example Sniffer Trace and Its Mutation Trace.}}
  \label{fig:loss}
\end{figure}

Suppose we have the true packet loss probabilities $\overline{Pr_{ds}},
\overline{Pr_{es}}$ and $\overline{Pr_{ed}}$, the \textit{confidence} of this
mutation trace $Tr'$ can be defined as:

\begin{align}
  Confidence(Tr') = \prod_{i\in\{ds, es, ed\}}(1-|Pr_i-\overline{Pr_i}|)
\end{align}

By this definition, the trace confidence maximizes when all packet loss
ratios calculated from $Tr'$ matches the true packet
loss probabilities.

Suppose we have $M$ mutation traces that satisfies $S$, then
the overall \textit{confidence of correctness} from sniffer trace $Tr$ can be defined as:

\begin{align}
  Confidence(Tr) &= 1-\prod_{j=1}^{M}(1-Pr(Tr'_j))
\end{align}

Since $0 \le 1-Pr(Tr'_j) \le 1$ for all $j$, we have:

\begin{align}
  \nonumber
  Confidence(Tr) &\ge 1-(1-Pr(Tr'_j)\\
     &= Pr(Tr'_j)
\end{align}for all $j$.

This means if we can find a mutation trace $Tr'$ that satisfies $S$, then the
overall confidence of the correctness is no less than $Pr(Tr')$.
\end{comment}

\makeatletter{}\section{Case Studies}
\label{sec:case}

We present case studies on applying our validation framework on two protocols
implemented in the \ns{} network simulator: 802.11 data transmission and ARF
rate control algorithm. The goal is to demonstrate how our framework can avoid
false alarms and report true violations on incomplete sniffer traces and report
true violations. We also report the application of our framework to a commercial
product under development.

\makeatletter{}\subsection{802.11 Data Transmission}
\label{subsec:tx}

In this section, we first show that our framework can improve validation
precision by inferring the missing or extra packets using the augmented
transition framework. We then demonstrate the ability of our framework to detect
true violations by manually introducing bugs in the \ns{} implementation and
show the precision and recall of validation results.

\subsubsection{Experimental Setup}

We set up two \wifi{} devices acting as the transmitter (DUT) and receiver
respectively. Another \wifi{} device is configured in monitor mode and acts as
the sniffer. During the experiments,we collect both the DUT packet trace
(the ground truth) and the sniffer trace.

\subsubsection{Verifying Unmodified Implementation}

In the original monitor state machine shown in Fig.~\ref{fig:dot11_tx_ta}, we
set acknowledgment timeout $T_o=334\mu$s, maximum retransmission delay
$T_m=15$ms according to the protocol. We also adapt the state machine to include
multiple retransmissions\footnote{The exact number of retransmissions is not
part of the protocol, and NS-3 implementation set this to be 7.} instead of one.

Let $Pr_{ds}$, $Pr_{es}$ and $Pr_{ed}$ be the packet loss probability between
the DUT and sniffer, endpoint and sniffer, endpoint and DUT respectively.
$Pr_{ed}$ represents the characteristics of the system being tested, while
$Pr_{ds}$ and $Pr_{es}$ represent the sniffer's quality in capturing packets.

We vary each of the three probabilities, $Pr_{ds}$, $Pr_{es}$ and $Pr_{ed}$, from
0 to 0.5 (both inclusive) with 0.05 step.  For each loss ratio combination, we
ran the experiment 5 times, and each run lasted
30 seconds. In total, 6655 ($11^3\times 5$) pairs of DUT and sniffer packet
traces were collected.

To establish the ground truth of violations, we first verify the DUT packet
traces using the \textit{original} state machine $S$.  This can be achieved by
disabling augmented transitions in our framework.  As expected, no violation is
detected in any DUT packet traces.

We then verify the sniffer traces using the augmented state machine $S^+$.  For
the $\mathit{GoBack}(k)$ heuristic, we set $k=7$, which is the maximum number of
transmissions of a single packet. For the $\mathit{NumMissing}(d, l, k)$ heuristic, we
set the sliding window size $l=100$, and $k=80$ such that no violation is
reported. The relationship of $k$ and validation precision is studied in next
section.

Next, we present detailed analysis of the augmented transitions on the sniffer
traces. The goal is to study for a given system packet loss probability
$Pr_{ed}$, how the sniffer packet loss properties ($Pr_{ds}$ and $Pr_{es}$)
affect the difference between the DUT trace and the mutation trace, which represents
a guess of the DUT trace by the augmented state machine based on the sniffer
trace.

For all following analysis, we divide the traces into three groups according to
$Pr_{ed}$: low ($0 \le Pr_{ed} \le 0.15$), medium ($0.20 \le Pr_{ed} \le 0.35$)
and high ($0.40 \le Pr_{ed} \le 0.50$).

The different between two packet traces can be quantified by the Jaccard distance
metric.

\begin{align}
  Jaccard(Tr_1, Tr_2) = \frac{\left\vert Tr_1 \ominus Tr_2\right\vert}{\left\vert
  Tr_1 \cup Tr_2\right\vert}
\end{align}where $\ominus$ is the symmetric difference operator.  The distance is 0 if the
two traces are identical, and is 1 when the two traces are completely different.
The smaller the distance is, the more similar the two traces are.

A na\"ive way to calculate the Jaccard distance of two traces is to use the hash
of the $(time, packet)$ pair for set intersection and union operation.  However,
it does not work for mutation trace which contains fabricated packets with no
actual payload.  Therefore, we use a protocol specific canonical representation
of packets when calculating the distance.  In particular, the string
\texttt{r\_\it{DATA}\_i\_t} represents the $t^{th}$ transmission of a data packet
with sequence number $i$, and $r$ represents the round of sequence numbers as it
wraps after 4096.  And similarly \texttt{r\_Ack\_i\_t} is the corresponding
acknowledgment packet.

\begin{figure*}[t!]
  \centering
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_10.pdf}
        \caption{$0.05 \le Pr_{ed} \le 0.15$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_30.pdf}
    \caption{$0.2 \le Pr_{ed} \le 0.35$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_50.pdf}
    \caption{$0.4 \le Pr_{ed} \le 0.5$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \caption{\textbf{Jaccard Distance Between Mutation and DUT Traces.} For each
  data point, the mean of the 5 runs is used.}
  \label{fig:mutation_dut}
\end{figure*}

Fig.~\ref{fig:mutation_dut} shows the Jaccard Distance between mutation and its
corresponding DUT trace. We make the following observations. First, for a given
system loss probability $Pr_{ed}$ (each sub-figure), the lower the sniffer
packet loss probability ($Pr_{ds}$ and $Pr_{es}$), the smaller Jaccard distance
between the DUT and mutation trace. Intuitively, this means a sniffer that
misses less packets can enable our framework to better reconstruct the DUT
trace.

Second, we observe a \textit{protocol-specific} trend that $Pr_{ds}$ is more
dominant than $Pr_{es}$.  This is because retransmission packets of the same
sequence number are identical, hence when the sniffer misses multiple
retransmission packets, our framework only needs to infer one retransmission
packet to continue state machine execution.

Finally, as the system loss probability $Pr_{ed}$ increases, the Jaccard
distance increases more rapidly as $Pr_{ds}$ increases.  This is because the
ratio of retransmission packet increases along with $Pr_{ed}$.

We then evaluate the cost of resolving uncertainty. In particular, we use
the Average Search Steps Per Packet (ASSPP) as a metric to quantify the search
cost.  It is calculated by dividing the total number of search steps by the
number of packets in the packet trace. For DUT traces, ASSPP is always 1 since
there is no uncertainty. For sniffer traces, however, multiple search steps must
be to conducted to resolve the potential uncertainty of each packet in the
sniffer trace.

\begin{figure*}[t!]
  \centering
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{StepCount3DFigure_0_10.pdf}
    \caption{$Pr_{ed} \in [0.05, 0.10, 0.15]$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{StepCount3DFigure_0_30.pdf}
    \caption{$Pr_{ed} \in [0.2, 0.25, 0.3, 0.35]$}
  \end{subfigure}\hspace*{0.01\textwidth}
  \begin{subfigure}{0.33\textwidth}
    \centering
    \includegraphics[width=\textwidth]{StepCount3DFigure_0_50.pdf}
    \caption{$Pr_{ed} \in [0.40, 0.45, 0.50]$}
  \end{subfigure}
  \caption{\textbf{Average Searching Step Per Packet.} For each data point, the mean of 5 runs
  is used.}
  \label{fig:cost}
\end{figure*}

Fig.~\ref{fig:cost} shows the distribution of ASSPP at different $Pr_{ed}$.
Similar to the case in Fig.~\ref{fig:mutation_dut}, $Pr_{ds}$ plays a dominant
role in determine the searching cost. One interesting observation for this
particular protocol is that the
search cost peaks when $Pr_{ds}$ is high while $Pr_{es}$ is low. In such loss
probability combinations, sniffer misses many data packets from the DUT but
picks up lots of \textit{dangling} $Ack$ packets from the DUT. Because the $Ack$
packet has neither sequence numbers nor retry flag, the searching algorithm had
a hard time resolving such uncertainty.

\subsubsection{Introducing Bugs}

We have demonstrated that our framework can tolerate sniffer imperfections and
avoid raising false alarms.  The next question is, can it detect true
violations?  To answer this question, we manually introduce several bugs in
\ns{} implementation that concerns various aspects of 802.11 data transmission
protocol.  More specifically, the bugs are:

\begin{itemize}
  \item \textbf{Sequence Number}: the DUT does not assign sequence number
    correctly. For example, it may increase sequence by 2 instead of 1, or it
    does not increase sequence number after certain packet, etc. We choose one
    type of such bugs in each run.
  \item \textbf{Semantic}: the DUT may retransmit even
    after receiving $Ack$, or does not retransmit when not receiving $Ack$.
\end{itemize}

We instrument the \ns{} implementation to embed instances of bugs in each
category. At each experiment run, we randomly decide whether and which bug to
introduce for each category. We fix $Pr_{ds}=Pr_{es}=0.1$ and vary $Pr_{ed}$
from 0.0 to 0.5 with 0.01 step. For each $Pr_{ed}$ value, we ran the experiment
100 times, of which roughly 75 experiments contained bugs. In total, 5100 pairs of
DUT and sniffer traces were collected.

We use the DUT packet traces as ground truth of whether or not each experiment
run contains bugs.
For each $Pr_{ed}$ value, we calculate the precision and recall of violation
detection using the sniffer traces.\begin{align}
  \text{Precision} = \frac{\left\vert \{\text{Reported Bugs}\} \cap \{\text{True Bugs}\}\right\vert}{\left\vert
  \{\text{Reported Bugs}\}\right\vert}\\
  \text{Recall} = \frac{\left\vert \{\text{Reported Bugs}\} \cap \{\text{True Bugs}\}\right\vert}{\left\vert
  \{\text{True Bugs}\}\right\vert}
\end{align}The precision metric quantifies how \textit{useful} the validation results are ,
while the recall metric measures how \textit{complete} the validation results
are.

\begin{figure}[t!]
  \centering
  \begin{subfigure}{0.48\textwidth}
    \includegraphics[width=\textwidth]{PrecisionFigure.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \includegraphics[width=\textwidth]{RecallFigure.pdf}
  \end{subfigure}
  \caption{\textbf{Precision and Recall of Validation Results.}}
  \label{fig:precision}
\end{figure}

Fig.~\ref{fig:precision} shows the CDF of precision and recall of the 51
experiments for various $k$ values. For precision, as expected, the more
tolerant the search to sniffer losses (larger $k$), the more tolerant the
framework is to sniffer losses, and the more precise the violation detection. In
particular, when $k=30$, the precisions are 100\% for all $Pr_{ed}$ values.
Second, the recall is less sensitive to the choice of $k$.  Except for the
extreme case when $k=30$, all other thresholds can report almost all the
violations.

\makeatletter{}\subsection{ARF Rate Control Algorithm}
\label{subsec:arf}

Automatic Rate Fallback (ARF)~\cite{kamerman1997wavelan} is the first rate
control algorithm in literature. In ARF, the sender increases the bit rate after
$Th_1$ number of consecutive successes or $Th_2$ number of packets with at most
one retransmission. The sender decreases bit rate after two consecutive packet
failures or if the first packet sent after rate increase (commonly referred as
\textit{probing} packet) fails.

\sloppy{  Fig.~\ref{fig:arf_sm} shows the state machine $S$ for the packet trace
  collected at sender (DUT), where $\it{DATA}_i^r$ denotes a data packet
  with sequence number $i$ and bit rate $r$, $\it{DATA}_i^{r\prime}$ is a retransmission packet
  and $Ack$ is the acknowledgment packet. The \texttt{pkg\_succ} function is
  shown in Algorithm~\ref{alg:pkt_succ}.
}

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.8\textwidth]{arf_sm.pdf}
  \caption{\textbf{Monitor State Machine for ARF Rate Control Algorithm.} Timing
  constraints are omitted for succinctness.}
  \label{fig:arf_sm}
\end{figure}

The \texttt{succ} variable is used to track the number of consecutive packet
successes. It is increased after each packet success , and is reset to 0 after a
rate increase or upon a packet failure ($s_1\rightarrow s_2$).  Similarly,
\texttt{count} is to track the number of packets with at most one
retransmission, and is increased after packet success, or for the first packet
retransmission ($s_1\rightarrow s_2$). It is reset when there are two
consecutive packet failures ($s_2\rightarrow s_3$). Finally, the \texttt{probe}
flag is set upon rate increases to indicate the probing packet, and is cleared
upon packet success. The variable \texttt{r} is the current bit rate, which is
decreased if the probing packet fails ($s_1\rightarrow s_4$), or every two
consecutive failures ($s_2\rightarrow s_3$). If \texttt{r} is not the highest
rate, it is increased when either of the two thresholds are reached.

\begin{algorithm}[t!]
  \caption{\texttt{pkt\_succ} function}
  \label{alg:pkt_succ}
  \begin{algorithmic}[1]
    \Function{pkt\_succ}{}
    \Let{i}{(i+1)\%N}
    \Let{succ}{succ + 1}
    \Let{count}{count + 1}
    \Let{probe}{false}
    \If{r $< R$ and (succ $\ge Th_1$ or count $\ge Th_2$)}
    \Let{r}{r+1}
    \Let{succ}{0}
    \Let{count}{0}
    \Let{probe}{true}
    \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

In particular, the bug we found lies in the implementation's \texttt{pkt\_succ}
function in line 6. Instead of checking \texttt{count $\ge$ Th\_2}, the
implementation checks \texttt{count == Th\_2}. This bug also exists in the NS-3
implementation of Adaptive ARF (AARF) algorithm~\cite{lacage2004ieee} and the
pseudo code implementation of AARF in~\cite{lacage2004report}.

Note that the \texttt{count} variable is incremented twice if a packet succeed
after one retransmission: once in $s_1\rightarrow s_2$, once in the
\texttt{pkt\_succ} function for the retransmission packet. Therefore, if the
value of \texttt{count} is $Th_2-1$ and the next packet succeed after one
retransmission, the value of \texttt{count} will be $Th_2+1$, which would fail
the implementation's test of \texttt{count == Th\_2}.

We report a bug found in NS-3 ARF~\cite{kamerman1997wavelan} implementation
which causes the sender to get stuck at a lower rate even after enough number of
consecutive successes. The bug was detected using sniffer traces and
confirmed by both the DUT trace and source code inspection.

\makeatletter{}\subsection{Industrial Application}
\label{subsec:xbox}

\begin{table}[t!]
  \centering
  \begin{tabular}{lll}
    \toprule
    \textbf{Protocol Aspects} & \textbf{Traces} & \textbf{Violations (\%)}\\
    \midrule
    Sequence Number & 3049 & 1539 (50.48\%) \\
    Station Scheduling & 3046 & 2045 (67.14\%) \\
    Uplink Modulation & 3127 & 8 (0.26\%) \\
    Downlink Modulation & 3127 & 24 (0.77\%) \\
    \bottomrule
  \end{tabular}
  \caption{\textbf{Validation Results on Traces from the Gaming Wireless Protocol.}}
  \label{tab:xbox}
\end{table}

We have applied our framework for runtime verification of the wireless protocol
implementation of a commercial product that has several millions of shipping
devices. The product runs a proprietary wireless protocol to reduce latency and
power consumption. Sniffer traces were collected in regular testing process, but
were only manually inspected previously.

We obtained 75 sniffer traces from the testing team for a new version of the
protocol that is under development and testing. This team has been testing the
implementation for a few weeks.  Each trace contained around 6 million packets
that were captured during 1 hour and 40 minutes of protocol operation.

We split the traces into \num{100000} packet segments, and applied our framework
to validate the DUT implementation. We found that the latest implementation of
the protocol under development had violations of the protocol specification.
Some of the implementation bugs we found related to the sequence number
management, station scheduling during power saving mode, and modulation rates
adaptation.

Table~\ref{tab:xbox} summarizes the validation result. Note that if we disable
the augmented transitions, each trace will be flagged as violation because of
the missing packets, thereby reducing the usability of the tool. We also note
that some bugs manifest more often than others. For instance, the bugs related
to packet sequence number and station scheduler were detected in about half the
traces, and the bug related to the rate control algorithm was detected in only a
few traces. This is because the previous two aspects are essential in all
protocol operations, while the bugs related to rate control only manifest
themselves under certain link conditions.

After communicating with the testing team, we confirmed that the sequence number
bug was already known, as it is relatively easy to detect even by manually
examining the traces. The bug related to station scheduling was also noticed
before, yet no quantitative results about how frequent this bug appears were
obtained because of the lack of automatized validating tools. Finally, the bug
related to rate control has been filed as a bug report. All reported bugs were
fixed before the next release the product.

\makeatletter{}\section{Related Work}
\label{sec:related}

\textbf{Hidden Markov Model (HMM) Approach.} When considering the whole system
under test (both DUT and endpoint), the sniffer only captures a subset of the
all the packets (events).  This is similar to the event sampling problem in
runtime
verification~\cite{bonakdarpour2011sampling,hauswirth2004low,arnold2008qvm,fei2006artemis,basin2012monitoring}.
Stoller \textit{et al}~\cite{stoller2011runtime} used HMM-based state estimation
techniques to calculate the confidence that the temporal property is satisfied
in the presence of gaps in observation.

While it seems possible to adapt the
method in~\cite{stoller2011runtime} to our problem, we note several advantages
of our approach. First, the
automatically augmented state machine precisely encodes the protocol
specification and the uncertainty. This is intuitive to design and natural for
reporting the evidence for a trace being successful. We do not require a user
to specify the number of states of the underlying HMM, or accurately provide
underlying probabilities. Second, we use timed automata to monitor the timing
constraints which are common in wireless protocols. It may be non-trivial to
encode such timing information in HMM. Finally, we can exploit domain knowledge
to devise effective pruning heuristics to rule out unlikely sequences during the
exhaustive search.

\textbf{Network Protocol Validation.} Lee \textit{et al}~\cite{lee1997passive}
studied the problem of passive network testing of network management. The system
input/output behavior is only partially observable. However, the uncertainty
only lies in missing events in the observation, while in the context of wireless
protocol verification, the uncertainty could also be caused by extra events not
observed by the tested system. Additionally, they do not provide any formal
guarantees even for cases when we report a definite bug.  Software model
checking techniques~\cite{musuvathi2002cmc,godefroid1997model} have also been
used to verify network protocols. Our problem is unique because of the
observation uncertainty caused by sniffers. Our framework shares similarity
with {\it angelic verification}~\cite{das-cav15} where the program verifier
reports a warning only when no acceptable specification exists on unknowns.

\begin{comment}
\textbf{Sniffer Trace Analysis.} Wireless sniffers has been widely used to
analyze MAC layer behaviors of enterprise wireless
networks~\cite{sheng:wicom2008,tan:tmc2014,yeo-wise04,yeo:witmemo2005}.
Jigsaw~\cite{Cheng:2006:JSP:1159913.1159920} is a larger scale wireless network
monitoring infrastructure. 150~radio monitors were deployed in a campus
building. Traces collected from multiple sniffers were merged and synchronized
to restructure the link and transportation layer conversations. Protocol
specific heuristics were developed to infer the missing packets. The work
in~\cite{Mahajan:2006:AMB:1159913.1159923} shared the same idea of trace merging
with Jigsaw, but uses a FSM to infer packet reception. These works assume the
correctness of the protocol implementation in order to infer missing packets,
while we systematically encode the uncertainty of sniffer traces for
verification purpose.

\textbf{ Testing Under Uncertainty}. The position paper by Rosemblum \textit{et
al}~\cite{Elbaum:2014:KUT:2635868.2666608} contains excellent motivation for the
need to combat uncertainty foundationally when testing systems. McKinley
\textit{et al}~\cite{bornholt2014uncertain,sampson2014expressing} deals with checking
assertions in programs dealing with noisy data from sensors. Instead of checking
the truth or falsity of assertions, they model the probability distribution of
the assertion conditions and perform Monte-carlo based simulations to estimate
the probabilities. Our work can be seen as leveraging non-determinism to weaken
the specification logically to precisely define the problem complexity, and use
probabilities to guide the search for likely mutations.  Other works have used
sampling to find data-race bugs~\cite{marino2009literace}, and ensure that the
sampling does not lead to spurious alarms.
\end{comment}

\makeatletter{}\section{Conclusions}
\label{sec:conclusion}

We formally define the uncertainty problem in validating wireless protocol
implementations using sniffers. We describe a systematic augmentation of the
protocol state machine to explicitly encode the uncertainty of sniffer traces.
We characterize the NP-completeness of the problem and propose both an
exhaustive search algorithm and heuristics to restrict the search to more likely
traces. We present two case studies using \ns{} network simulator to demonstrate
how our framework can improve validation precision and detect real bugs. We also
report the application of our framework on a commercial product under
development.

Finally, we discuss a few challenges and future
directions.

\textbf{Verification Coverage.} Given a single sniffer trace, it is possible
that not all the states in the state machine are visited during the verification
process. For instance, a rate control state machine based on certain consecutive
packet losses patterns can not be verified if no such consecutive losses appear
in the sniffer trace. In general, given a protocol state machine, it is
challenging to extract the packet patterns for each state to be reached and to
alter the testing such that such patterns can be observed.

\textbf{Automated State Machine Construction.} We manually constructed the
protocol monitor state machines for the protocols studied in this paper based on
the source code, comments and documentation. The process involves lots of labor
effort and is time-consuming. A potential alternative is to automatically learn
the sketch of the monitor state machines from the sniffer traces. Domain
knowledge can then be leveraged to improve the sketch state machines.

{\footnotesize
  \bibliographystyle{abbrv}
  \begin{thebibliography}{10}

\bibitem{alur1994theory}
R.~Alur and D.~L. Dill.
\newblock A theory of timed automata.
\newblock {\em Theoretical computer science}, 126(2):183--235, 1994.

\bibitem{arnold2008qvm}
M.~Arnold, M.~Vechev, and E.~Yahav.
\newblock Qvm: an efficient runtime for detecting defects in deployed systems.
\newblock In {\em ACM Sigplan Notices}, volume~43, pages 143--162. ACM, 2008.

\bibitem{bahl2006enhancing}
P.~Bahl, R.~Chandra, J.~Padhye, L.~Ravindranath, M.~Singh, A.~Wolman, and
  B.~Zill.
\newblock Enhancing the security of corporate \wifi{} networks using {DAIR}.
\newblock In {\em Proceedings of the 4th international conference on Mobile
  systems, applications and services}, pages 1--14. ACM, 2006.

\bibitem{basin2012monitoring}
D.~Basin, F.~Klaedtke, S.~Marinovic, and E.~Z{\u{a}}linescu.
\newblock Monitoring compliance policies over incomplete and disagreeing logs.
\newblock In {\em International Conference on Runtime Verification}, pages
  151--167. Springer, 2012.

\bibitem{bonakdarpour2011sampling}
B.~Bonakdarpour, S.~Navabpour, and S.~Fischmeister.
\newblock Sampling-based runtime verification.
\newblock In {\em FM 2011: Formal Methods}, pages 88--102. Springer, 2011.

\bibitem{cheng2006jigsaw}
Y.-C. Cheng, J.~Bellardo, P.~Benk{\"o}, A.~C. Snoeren, G.~M. Voelker, and
  S.~Savage.
\newblock {\em Jigsaw: solving the puzzle of enterprise 802.11 analysis},
  volume~36.
\newblock ACM, 2006.

\bibitem{wifried}
M.~Ciabarra.
\newblock {WiFried: iOS 8 WiFi Issue}.
\newblock \url{https://goo.gl/KtRDqk}.

\bibitem{das-cav15}
A.~Das, S.~K. Lahiri, A.~Lal, and Y.~Li.
\newblock Angelic verification: Precise verification modulo unknowns.
\newblock In {\em Computer Aided Verification - 27th International Conference,
  {CAV} 2015, San Francisco, CA, USA, July 18-24, 2015, Proceedings, Part {I}},
  pages 324--342, 2015.

\bibitem{surface}
digitalmediaphile.
\newblock Windows 10 wifi issues with surface pro 3 and surface 3.
\newblock \url{http://goo.gl/vBqiEo}.

\bibitem{edelkamp2008survey}
S.~Edelkamp, V.~Schuppan, D.~Bo{\v{s}}na{\v{c}}ki, A.~Wijs, A.~Fehnker, and
  H.~Aljazzar.
\newblock Survey on directed model checking.
\newblock In {\em International Workshop on Model Checking and Artificial
  Intelligence}, pages 65--89. Springer, 2008.

\bibitem{fei2006artemis}
L.~Fei and S.~P. Midkiff.
\newblock Artemis: Practical runtime monitoring of applications for execution
  anomalies.
\newblock In {\em ACM SIGPLAN Notices}, volume~41, pages 84--95. ACM, 2006.

\bibitem{lollipop}
Gizmodo.
\newblock The worst bugs in android 5.0 lollipop and how to fix them.
\newblock \url{http://goo.gl/akDcvA}.

\bibitem{godefroid1997model}
P.~Godefroid.
\newblock Model checking for programming languages using verisoft.
\newblock In {\em Proceedings of the 24th ACM SIGPLAN-SIGACT symposium on
  Principles of programming languages}, pages 174--186. ACM, 1997.

\bibitem{iris}
Google.
\newblock Google contact lens.
\newblock \url{https://en.wikipedia.org/wiki/Google_Contact_Lens}.

\bibitem{hauswirth2004low}
M.~Hauswirth and T.~M. Chilimbi.
\newblock Low-overhead memory leak detection using adaptive statistical
  profiling.
\newblock In {\em Acm Sigplan Notices}, volume~39, pages 156--164. ACM, 2004.

\bibitem{kamerman1997wavelan}
A.~Kamerman and L.~Monteban.
\newblock Wavelan{\textregistered}-ii: a high-performance wireless lan for the
  unlicensed band.
\newblock {\em Bell Labs technical journal}, 2(3):118--133, 1997.

\bibitem{lacage2004ieee}
M.~Lacage, M.~H. Manshaei, and T.~Turletti.
\newblock {IEEE 802.11 rate adaptation: a practical approach}.
\newblock In {\em Proceedings of the 7th ACM international symposium on
  Modeling, analysis and simulation of wireless and mobile systems}, pages
  126--134. ACM, 2004.

\bibitem{lacage2004report}
M.~Lacage, M.~H. Manshaei, and T.~Turletti.
\newblock {IEEE 802.11 rate adaptation: a practical approach}.
\newblock {\em [Research Report] RR-5208}, (<inria-00070784>):25, 2004.

\bibitem{lee1997passive}
D.~Lee, A.~N. Netravali, K.~K. Sabnani, B.~Sugla, and A.~John.
\newblock Passive testing and applications to network management.
\newblock In {\em Network Protocols, 1997. Proceedings., 1997 International
  Conference on}, pages 113--122. IEEE, 1997.

\bibitem{mahajan2006analyzing}
R.~Mahajan, M.~Rodrig, D.~Wetherall, and J.~Zahorjan.
\newblock Analyzing the mac-level behavior of wireless networks in the wild.
\newblock In {\em ACM SIGCOMM Computer Communication Review}, volume~36, pages
  75--86. ACM, 2006.

\bibitem{musuvathi2002cmc}
M.~Musuvathi, D.~Y. Park, A.~Chou, D.~R. Engler, and D.~L. Dill.
\newblock {CMC}: A pragmatic approach to model checking real code.
\newblock {\em ACM SIGOPS Operating Systems Review}, 36(SI):75--88, 2002.

\bibitem{mytkowicz2008observer}
T.~Mytkowicz, P.~F. Sweeney, M.~Hauswirth, and A.~Diwan.
\newblock Observer effect and measurement bias in performance analysis.
\newblock 2008.

\bibitem{riley2010ns}
G.~F. Riley and T.~R. Henderson.
\newblock The ns-3 network simulator.
\newblock In {\em Modeling and Tools for Network Simulation}, pages 15--34.
  Springer, 2010.

\bibitem{omnipeek}
{Savvius Inc.}
\newblock Savvius \wifi{} adapters.
\newblock \url{https://goo.gl/l3VXSx}.

\bibitem{Shi2016}
J.~Shi, S.~K. Lahiri, R.~Chandra, and G.~Challen.
\newblock {\em Wireless Protocol Validation Under Uncertainty}, pages 351--367.
\newblock Springer International Publishing, Cham, 2016.

\bibitem{stoller2011runtime}
S.~D. Stoller, E.~Bartocci, J.~Seyster, R.~Grosu, K.~Havelund, S.~A. Smolka,
  and E.~Zadok.
\newblock Runtime verification with state estimation.
\newblock In {\em Runtime Verification}, pages 193--207. Springer, 2011.

\bibitem{chromecast}
Wikipedia.
\newblock {Chromecast}.
\newblock \url{https://en.wikipedia.org/wiki/Chromecast}.

\bibitem{xbox}
Wikipedia.
\newblock {Xbox One} controller.
\newblock \url{https://en.wikipedia.org/wiki/Xbox_One_Controller}.

\end{thebibliography}

}

\end{document}

### 2017-07-09 background.tex
\it{DATA}
comments

@@
 A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
-the state machine at time $t_i$. The monitor {\it rejects a trace} $Tr$ if there
-exists a prefix of $Tr$ such that all states reachable after consuming the
-prefix have no valid transitions for the next $(t, p)$ input.
-
-As an example, the monitor state machine illustrated in
-Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:
+the state machine at time $t_i$.  As an example, the monitor state machine
+illustrated in Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:
 
 \begin{itemize}
-  \item $\Sigma = \{DATA_i, DATA'_i, Ack\ |\ 0 \le i < N\}$.
+  \item $\Sigma = \{\it{DATA}_i, \it{DATA}'_i, Ack\ |\ 0 \le i < N\}$.
   \item Clock variables $C = \{c\}$. The only clock variable $c$ is
@@
     T_o$ is the maximum delay allowed before the retransmission packet gets
-    sent. $T_o$ can be arbitrary large but not infinity in order to check the
-    liveness of the DUT.
+    sent. In this particular case, $T_o$ can be arbitrary large but not infinity
+        in order to check the liveness of the DUT.
 \end{itemize}
+
+\begin{definition}[Trace Rejection and Acceptance]
+    A monitor state machine $S$ {\it rejects a trace} $Tr$ if there
+    exists a prefix of $Tr$ such that all states reachable after consuming the
+    prefix have no valid transitions for the next $(t, p)$ input. Otherwise, $S$
+    {\it accepts} $Tr$ (or $Tr$ {\it satisfies} $S$).
+\end{definition}
 

### 2017-07-09 case/tx.tex
comments

@@
 (the ground truth) and the sniffer trace.
+
+The DUT and endpoint are configured to use the IEEE 802.11g standard with both
+RTS/CTS and fragmentation disabled.  A Constant Bit Rate (CBR) UDP traffic
+(54~Mbps) is generated from the DUT to the endpoint.  The UDP packet size is
+1436 bytes, which results in a 1500 bytes \wifi{} packet.
+
+In order to control the packet loss ratios between each pair of devices, we
+developed a new propagation loss model for \ns{} called
+\texttt{MatrixRandomPropagationLossModel}.
+\sloppy{%
+  Instead of a constant propagation
+  loss as in existing \texttt{MatrixPropagationLossModel}, the signal
+  propagation loss between a pair of nodes is determined by a binary random
+  variable of two  values: 0~dB (no loss) and 1000~dB (complete loss).
+  }
+Therefore, any packet loss
+probability can be achieved by adjusting the random variable distribution.
+The
+model supports both symmetric and asymmetric propagation losses.
+We use
+symmetric propagation loss in all our experiments.
+Finally, pcap capture is enabled in both the DUT and the sniffer devices.
 
@@
 
-The different between two packet traces can be quantified by the Jaccard distance
+The difference between two packet traces can be quantified by the Jaccard distance
 metric.
@@
 of packets when calculating the distance.  In particular, the string
-\texttt{r\_\it{DATA}\_i\_t} represents the $t^{th}$ transmission of a data packet
+\textit{r\_DATA\_i\_t} represents the $t^{th}$ transmission of a data packet
 with sequence number $i$, and $r$ represents the round of sequence numbers as it
-wraps after 4096.  And similarly \texttt{r\_Ack\_i\_t} is the corresponding
+wraps after 4096.  And similarly \textit{r\_ACK\_i\_t} is the corresponding
 acknowledgment packet.
@@
 there is no uncertainty. For sniffer traces, however, multiple search steps must
-be to conducted to resolve the potential uncertainty of each packet in the
+be conducted to resolve the potential uncertainty of each packet in the
 sniffer trace.

### 2017-07-09 case/xbox.tex
comments

@@
   \end{tabular}
-  \caption{\textbf{Validation Results on Traces from the Gaming Wireless Protocol.}}
+  \caption{\textbf{Validation Results on Traces from the Gaming Controller Wireless Protocol.}}
   \label{tab:xbox}
@@
 
-We split the traces into \num{100000} packet segments, and applied our framework
-to validate the DUT implementation. We found that the latest implementation of
-the protocol under development had violations of the protocol specification.
-Some of the implementation bugs we found related to the sequence number
-management, station scheduling during power saving mode, and modulation rates
-adaptation.
+We first split the traces into \num{100000} packet segments, which yields \num{3127}
+traces for testing.  We then applied our framework on the traces to validate the
+DUT implementation. We found that the latest implementation of the protocol
+under development had violations of the protocol specification.  Some of the
+implementation bugs we found related to the sequence number management, station
+scheduling during power saving mode, and modulation rates adaptation.
+Table~\ref{tab:xbox} summarizes the validation results. The \textbf{Traces}
+column shows the number of traces that we have successfully validated for each
+category, and the third column shows the fraction of traces that contain at
+least one violation in that category.
 
-Table~\ref{tab:xbox} summarizes the validation result. Note that if we disable
+Note that if we disable
 the augmented transitions, each trace will be flagged as violation because of
@@
 obtained because of the lack of automatized validating tools. Finally, the bug
-related to rate control has been filed as a bug report. All reported bugs were
-fixed before the next release the product.
+related to rate control, which was unknown previously, has been filed as a bug
+report. All reported bugs were fixed before the next release the product.

### 2017-07-09 framework/augment.tex
all-in-one
comments

@@
   \centering
-  \includegraphics[width=0.7\textwidth]{./figures/dot11_tx_checker.pdf}
+  \includegraphics[width=0.7\textwidth]{dot11_tx_checker.pdf}
   \caption{\textbf{Augmented Monitor State Machine.} Augmented transitions are
@@
 
-A accepting run of $S^+$ on sniffer trace $Tr$ yields a mutation trace $Tr'$
+An accepting run of $S^+$ on sniffer trace $Tr$ yields a mutation trace $Tr'$
 which represents one possibility of the DUT trace. Specifically, $Tr'$ can be
@@
 \begin{theorem}
+    Let $Tr$ be a sniffer trace, $\mathcal{M}(Tr)$ be $Tr$'s mutation traces, $S$
+    and $S^+$ be the original and augmented monitor state machine respectively.
   There exists a mutation trace $Tr' \in \mathcal{M}(Tr)$ that satisfies $S$ if
@@
 
-  Now we show that there exists a accepting transition sequence $P'$ of $S^+$ on
+  Now we show that there exists an accepting transition sequence $P'$ of $S^+$ on
   input $Tr'$ that does not contain augmented transitions.  In particular, $P'$
   can be obtained by substituting all Type-1 transitions with corresponding
-  original transitions, and removing all Type-2 transition.  Since $P'$ does not
+  original transitions, and removing all Type-2 transitions.  Since $P'$ does not
   contain augmented transitions, it is also an accepting transition sequence of
@@
 
-  We construct a accepting transition sequence $P$ of $S^+$ on input $Tr$ as
+  We construct an accepting transition sequence $P$ of $S^+$ on input $Tr$ as
   follows.

### 2017-07-09 framework/search.tex
comments

@@
 Intuitively, this means the search algorithm tries to utilize the sniffer's
-observation as much as possible before being forced to make assumptions.
+observation as much as possible before considering the packet loss by the
+sniffer or DUT.
 

### 2017-07-09 related.tex
comments

--- 2017-03-02 submitted/all-in-one.tex
@@
-\makeatletter{}\RequirePackage{fix-cm}
-\documentclass[smallextended]{svjour3}
-
-\usepackage{booktabs}
-\usepackage{subcaption}
-\usepackage{comment}
-\usepackage{mathrsfs}
-\usepackage{balance}
-\usepackage{enumitem}
-\usepackage{hyperref}
-\usepackage[noend]{algpseudocode}
-\usepackage[group-separator={,}]{siunitx}
-\usepackage{caption}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{listings}
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage{bm}
-\usepackage[all]{hypcap}
-\usepackage[absolute]{textpos}
-\usepackage{wrapfig}
-\usepackage[title]{appendix}
-
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
-\algnewcommand{\LineComment}[1]{\State{\(\triangleright\) #1}}
-\newcommand*\Let[2]{\State #1 := #2}
-
-\smartqed
-
-\usepackage{graphicx}
-\graphicspath{{./figures/}}
-
-\hypersetup{  bookmarks=true,
-  unicode=true,
-  pdftoolbar=true,
-  pdfmenubar=true,
-  pdffitwindow=true,
-  pdfstartview={FitV},
-  pdfnewwindow=true,
-  colorlinks=false,
-  pdfdisplaydoctitle=true,
-  pdfborder={0 0 0}
-}
-
-\setlength{\TPHorizModule}{1in}
-\setlength{\TPVertModule}{1in}
-\textblockorigin{0.75in}{0.875in}
-
-\setlist[itemize]{leftmargin=*,partopsep=5pt}
-\setlist[enumerate]{leftmargin=*,partopsep=5pt}
-
-\paperheight 11in
-\paperwidth 8.5in
-
-\usepackage{dcolumn}
-\newcolumntype{d}[1]{D{.}{.}{#1}}
-
-\newcommand\nil{\textit{nil}}
-
-\usepackage{tikz}
-\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-      \node[shape=circle,draw,inner sep=0pt] (char) {#1};}}
-
-\makeatletter
-\newcommand*{\rome}[1]{\expandafter\@slowromancap\romannumeral #1@}
-\makeatother
-
-\makeatletter{}
-\newcommand{\wifi}{Wi-Fi}
-\newcommand{\codename}{\textsc{Verifier}}
-\newcommand{\us}{$\mu$s}
-\newcommand{\ns}{\textsc{NS-3}}
-
-\newcommand{\shuvendu}[1]{{\bf [[SKL:{#1}]]}}
-
-\def\thetitle{Wireless Protocol Validation Under Uncertainty}
-\def\theauthors{}
-
-\authorrunning{J. Shi, S. K. Lahiri, R. Chandra, G. Challen}
-
-\journalname{Formal Methods in System Design}
-
-\begin{document}
-
-\title{  \thetitle\thanks{This work was published, in part, in Runtime Verification
-    (RV) 2016~\cite{Shi2016}.}
-}
-
-\author{  Jinghao Shi$^1$ \and
-  Shuvendu K. Lahiri$^2$ \and
-  Ranveer Chandra$^2$ \and
-  Geoffrey Challen$^1$
-}
-
-\institute{
-  Jinghao Shi\\
-  jinghaos@buffalo.edu\\\\
-  Shuvendu K. Lahiri\\
-  shuvendu@microsoft.com\\\\
-  Ranveer Chandra\\
-  ranveer@microsoft.com\\\\
-  Geoffrey Challen\\
-  challen@buffalo.edu\\\\
-  $^1$University at Buffalo, Buffalo, NY 14260, USA\\
-  $^2$Microsoft Research, Redmond, WA 98052, USA
-}
-
-\hypersetup{  pdfinfo={    Title={\thetitle},
-    Author={\theauthors},
-  }
-}
-
-\maketitle
-
-\makeatletter{}\begin{abstract}
-  Runtime validation of wireless protocol implementations cannot always employ
-  direct instrumentation of the device under test (DUT). The DUT may not
-  implement the required instrumentation, or the instrumentation may alter the
-  DUT's behavior when enabled. Wireless sniffers can monitor the DUT's
-  behavior without instrumentation, but they introduce new validation challenges.
-  Losses caused by wireless propagation prevent sniffers
-  from perfectly reconstructing the actual DUT packet trace. As a result,
-  accurate validation requires distinguishing between
-  specification deviations that represent implementation errors and those
-  caused by sniffer uncertainty.
-
-  \sloppy{
-  We present a new approach enabling sniffer-based validation of wireless
-  protocol implementations. Beginning with the original protocol monitor state
-  machine, we automatically and completely encode sniffer uncertainty by
-  selectively adding non-deterministic transitions. We characterize the
-  NP-completeness of the resulting decision problem and provide an exhaustive
-  algorithm for searching over all mutated traces. We also present
-  practical protocol-oblivious heuristics for searching over the most likely
-  mutated traces.  We have implemented our framework and show that it can
-  accurately identify implementation errors in the face of uncertainty.
-}
-
-  \keywords{
-    Runtime Verification \and
-    Wireless Protocol \and
-    Sniffer \and
-  Uncertainty}
-\end{abstract}
-
-\makeatletter{}\section{Introduction}
-\label{sec:intro}
-
-Custom wireless protocols are often designed and deployed to meet the specific
-performance and power needs of special-purpose wireless devices. Examples
-include Google
-Iris contact lenses~\cite{iris}, Xbox One wireless controllers~\cite{xbox}, and
-Google Chromecast~\cite{chromecast}. Validating that device implementations work
-correctly is critical to achieve the design goals of the wireless protocol and
-also prevent bugs in shipped products~\cite{wifried,lollipop,surface}.
-
-Runtime validation of the protocol implementations on such devices is
-challenging because collecting traces from the device under test (DUT) is often
-infeasible. The resource limitations of embedded or battery-powered devices may
-cause them to not provide trace collecting capabilities. DUT may contain
-proprietary hardware or firmware that hides the implementation details and
-prevents testers from collecting traces through source code instrumentation.
-Even when collecting trace directly from the DUT is possible, the overhead it causes
-may alter the behavior of the DUT due to the observer
-effect~\cite{mytkowicz2008observer}, threatening the validation results.
-
-An attractive alternative is to use wireless
-sniffers to record traffic generated by the DUT during testing.
-Sniffers do not require direct access to the DUT or the need to alter its behavior.
-However, due to the fundamentally unpredictable nature of wireless
-communications, the packets captured by the sniffer will not exactly match
-those received by the DUT.
-The sniffer may miss packets that the DUT received, or receive packets that
-the DUT missed.
-This is true even when using multiple
-sniffers~\cite{cheng2006jigsaw,mahajan2006analyzing,bahl2006enhancing}, a sniffer
-with multiple antennas~\cite{omnipeek}, or in isolated wireless environments.
-
-Since the sniffer trace may not perfectly match the actual trace, uncertainty
-arises during protocol implementation validation.  For example, if the DUT fails
-to respond correctly to a packet in the sniffer trace, it may either because the
-DUT's implementation is incorrect, or the DUT did not actually receive the
-packet, or the DUT's response was missed by the sniffer.  Whenever the DUT's
-behavior does not match the specification, there are now two potential
-explanations: either the DUT's implementation is wrong, or the sniffer trace is
-inaccurate. Accurate validation requires  distinguishing between these two
-causes.
-
-We present a new technique than enables validation of protocol implementations using
-wireless sniffers. Given a monitor state machine representing the protocol being
-validated, we describe a systematic transformation that adds non-deterministic
-transitions to incorporate uncertainty introduced by the sniffer. This
-augmented validation state machine implicitly defines a set of
-mutated traces, each satisfying the original state machine with a specific
-likelihood. If the set is empty, the implementation definitely violates the
-protocol. Searching over all the mutated traces is NP-complete, but the approach
-can be made practical by applying protocol-oblivious heuristics that limit the
-search to likely mutated traces.
-
-Our paper makes the following contributions:
-\begin{enumerate}
-  \item To the best of our knowledge, we are the first to identify the
-    uncertainty problem caused by sniffers in validating wireless protocol
-    implementations.
-  \item We formalize the problem using a nondeterministic state machine that
-    systematically and completely encodes the uncertainty of the
-    sniffer trace.
-  \item We characterize the NP-completeness of the validation problem, and
-    present two protocol-oblivious heuristics to prune the search space and make
-    validation possible in practice.
-  \item We implement the validation framework and evaluate it using the \ns{}
-    network simulator~\cite{riley2010ns}. Our framework accurately identifies
-    both synthetic and previously unknown violations in \ns{}'s implementations
-    of the 802.11 and ARF protocols.  We also applied our framework to a
-    commercial product under development and was able to found three latent bugs
-    that were not observed previously.
-\end{enumerate}
-
-This paper is an extension of our work that appeared in RV 2016~\cite{Shi2016}.
-We added the proof for Lemma~\ref{lem:mutation}, \ref{lem:sub} and
-Theorem~\ref{the:equivalent}. We included new results that show the searching
-cost (Fig.~\ref{fig:cost}), and details of the ARF protocol violations
-(Section~\ref{subsec:arf}). Further, we reported the application of our
-framework to a commercial product under development in
-Section~\ref{subsec:xbox}.
-
-\makeatletter{}\section{Background and Motivating Example}
-\label{sec:motivation}
-
-We encountered the uncertainty problem while testing the protocol implementation
-of a popular wireless game controller. A custom wireless protocol
-was designed to meet the low latency and low power consumption goals. As is
-common industry practice, the protocol specification was then handed over to
-wireless chipset vendors for implementation. However, neither implementation
-details nor trace collection capabilities are provided in the shipped firmware
-due to intellectual property constraints and device resource limitation. Hence
-using sniffers to validate the protocol implementation is the only option.
-
-We initially developed a tool to validate certain protocol properties over the
-sniffer trace, yet often found unacceptable amount of false alarms due to
-the incompleteness of the sniffer traces, making the tool virtually useless. It
-was clear that we needed to account for sniffer uncertainty.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=0.6\textwidth]{dot11_tx_ta.pdf}
-  \caption{\textbf{Monitor State Machine for 802.11 Transmitter.}}
-  \label{fig:dot11_tx_ta}
-\end{figure}
-
-To better understand the incompleteness of sniffer trace, consider the IEEE
-802.11 (also known as \wifi{}) transmitter (DUT) state machine shown in
-Fig.~\ref{fig:dot11_tx_ta}. After the DUT sends $\it{DATA}_i$---a data packet with
-sequence number $i$ ($s_0\rightarrow s_1)$, it starts a timer and waits for the
-acknowledgment packet---$Ack$. The DUT either receives $Ack$ within time $T_o$
-($s_1\rightarrow s_0$), or it sends $\it{DATA}'_i$---retransmission of $\it{DATA}_i$
-($s_1\rightarrow s_2$). Similarly, the DUT either receives the $Ack$ within $T_o$
-($s_2\rightarrow s_0$) or aborts transmission and moves on to next
-packet\footnote{To represent the state machine succinctly, our example assumes
-that the DUT retries at most once.} ($s_2\rightarrow s_1$).
-
-Given a complete log of DUT's packet transmission and reception events, it is
-trivial to feed such a log into the state machine in Fig.~\ref{fig:dot11_tx_ta}
-and validate the correctness of DUT's protocol implementation. However, due to
-DUT limitations we have described earlier, this complete log is not
-available. As a result, we seek to validate the DUT implementation using
-sniffers.
-
-There are two fundamental properties in wireless communication that bring uncertainty to
-sniffer's observation: packet loss and physical diversity. The sniffer could
-either miss packets sent from or to the DUT due to packet loss, or overhear packets
-that are sent to but missed by the DUT due to physical diversity.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=0.6\textwidth]{false_pos.pdf}
-  \caption{\textbf{Uncertainty of Sniffer Observations.} $Tr_{OTA}$ is
-    the chronological sequence of packets sent by the DUT and the receiver.
-    $Tr_{DUT}$ is DUT's internal events. $Tr_1$ and $Tr_2$ are two examples of
-    many possible sniffer traces.}
-  \label{fig:sniffer_in_middle}
-\end{figure}
-
-Consider a correct packet exchange sequence shown in
-Fig.~\ref{fig:sniffer_in_middle}. The DUT first sends $\it{DATA}_0$.  Suppose the
-receiver receives $\it{DATA}_0$ and sends the $Ack$ which the DUT does not receive.
-Eventually the DUT's timer fires and it sends $\it{DATA}'_0$.  This time the
-$\it{DATA}'_0$ reaches receiver and the DUT also receives the $Ack$.
-
-Now consider two possible traces that could have been overheard by a sniffer shown
-in Fig.~\ref{fig:sniffer_in_middle}.
-In first sniffer trace $Tr_1$ where the sniffer
-\textit{overhears} the first $Ack$ packet, a validation \textit{uncertainty}
-arises when the sniffer sees the $\it{DATA}'_0$: was the previous $Ack$ missed by the
-DUT or is there a bug in DUT which causes it to retransmit even after receiving
-the $Ack$?
-
-Similarly, consider the second possible sniffer trace $Tr_2$
-where both the $\it{DATA}'_0$ and $Ack$ packets were missed by the sniffer. During
-this period of time, it appears the DUT neither receives $Ack$ for $\it{DATA}_0$ nor
-sends $\it{DATA}'_0$. Again, without any additional information it is impossible to
-disambiguate between the sniffer missing certain packets and a bug in DUT's
-retransmission logic.
-
-Informally, the question we set out to answer in this paper is: given the
-protocol monitor state machine and the sniffer's observation with inherent
-uncertainty, how to accurately validate that the DUT behaves as specified?
-
-\makeatletter{}\section{Prerequisites and Problem Statement}
-\label{sec:background}
-
-\subsection{Packet, Trace and Monitor State Machine}
-\label{subsec:basic}
-
-The alphabet of the monitor state machine is the finite set of all valid packets
-defined by the protocol, denoted as $\mathbb{P}$.  A packet is a binary string
-of a finite number of bits, encoding interesting protocol attributes such as
-\texttt{src}, \texttt{dest}, \texttt{type}, \texttt{flags}, and physical layer
-information, such as \texttt{channel}, \texttt{modulation}, etc. The input of
-the state machine then corresponds to a time-ordered sequence of packets.
-
-\begin{definition}[Packet Trace]
-  \sloppy{
-  A \textit{packet trace} is a finite sequence of $(timestamp, packet)$ tuple:
-  $[(t_1, p_1), (t_2, p_2),\ldots,(t_n, p_n)]$ where $t_i \in \mathbb{Z}^+$ is
-  the \textit{discrete} timestamp and $p_i$ is the packet observed at time
-  $t_i$. The timestamps are strictly monotonically increasing: $t_i < t_{i+1}$
-  for $1 \le i < n$.
-}
-\end{definition}
-In addition to timestamp monotonicity, we also require that adjacent packets do
-not overlap in time, $t_{i+1}-t_i > \text{\texttt{airtime}}(p_i)$ for $1 \le i <
-n$, where \texttt{airtime()} calculates the time taken to transmit a packet.
-The timestamp represents the observer's local clock ticks, and need not to be
-synchronized among devices.
-
-We use \textit{timed automata}~\cite{alur1994theory} to model the expected
-behaviors of the DUT. A timed automata is a finite state machine with timing
-constraints on the transitions: each transition can optionally start one or more
-timers, which can later be used to assert certain events should be seen before
-or after the time out event. We refer the readers to~\cite{alur1994theory} for
-more details about timed automata.
-
-\begin{definition}[Monitor]
-  \sloppy{    A protocol monitor state machine $S$ is a 7-tuple
-    $\{\Sigma, \mathbb{S}, \mathbb{X}, s_0, C, E, G\}$, where:
-  }
-  \begin{itemize}
-    \item $\Sigma = \mathbb{P}$ is the finite input alphabet.
-
-    \item $\mathbb{S}$ is a non-empty, finite set of states. $s_0 \in
-      \mathbb{S}$ is the initial state.
-
-    \item $\mathbb{X}$ is the set of boolean variables. We use $v = \{x
-      \leftarrow true/false\ |\ x \in \mathbb{X}\}$ to denote an assignment of
-      the variables. Let $\mathbb{V}$ be the set of such values $v$.
-
-    \item $C$ is the set of clock variables. A \textit{clock variable} can be
-      reset along any state transitions. At any instant, reading a clock
-      variable returns the time elapsed since last time it was reset.
-
-    \item $G$ is the set of guard conditions defined inductively by
-      \begin{align*}
-        g := true\ |\ c \le T\ |\ c \ge T\ |\ x\ |\ \neg g\ |\ g_1 \land g_2
-      \end{align*}      where $c \in C$ is a clock variable, $T$ is a constant, and $x$ is a
-      variable in $\mathbb{X}$.  A transition can choose not to use guard
-      conditions by setting $g$ to be $true$.
-
-    \item $E \subseteq \mathbb{S} \times \mathbb{V} \times \mathbb{S} \times
-      \mathbb{V} \times \Sigma \times  G \times \mathscr{P}(C)$ gives the set of
-      transitions.\\ $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$
-      represents that if the monitor is in state $s_i$ with variable assignments
-      $v_i$, given the input tuple $(t, p)$ such that the guard $g$ is
-      satisfied, the monitor can transition to a state $s_j$ with variable assignments
-      $v_j$, and reset the clocks in $C'$ to 0.
-  \end{itemize}
-  \label{def:sm}
-\end{definition}
-
-A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
-the state machine at time $t_i$. The monitor {\it rejects a trace} $Tr$ if there
-exists a prefix of $Tr$ such that all states reachable after consuming the
-prefix have no valid transitions for the next $(t, p)$ input.
-
-As an example, the monitor state machine illustrated in
-Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:
-
-\begin{itemize}
-  \item $\Sigma = \{\it{DATA}_i, \it{DATA}'_i, Ack\ |\ 0 \le i < N\}$.
-  \item Clock variables $C = \{c\}$. The only clock variable $c$ is
-    used for acknowledgment time out.
-  \item $\mathbb{X} = \{i\}$, as a variable with ${\mathit log}(N) + 1$ bits to
-    count from $0$ to $N$.  
-  \item Guard constraints $G = \{ c \le T_o, c > T_o, T_o < c \le T_m\}$.
-    $T_o$ is the acknowledgment time out value, and $T_m >
-    T_o$ is the maximum delay allowed before the retransmission packet gets
-    sent. $T_o$ can be arbitrary large but not infinity in order to check the
-    liveness of the DUT.
-\end{itemize}
-
-The monitor state machine defines a \textit{timed language} $L$ which consists
-of all valid packet traces that can be observed by the DUT.  We now give the
-definition of protocol \textit{compliance} and \textit{violation}.
-
-\begin{definition}[Violation and Compliance]
-  Suppose $\mathbb{T}$ is the set of all possible packet traces collected from
-  DUT, and $S$ is the state machine specified by the protocol. The DUT
-  \textit{violates} the protocol specification if there exists an
-  packet trace $Tr \in \mathbb{T}$ such that $S$ rejects $Tr$.
-  Otherwise, the DUT is \textit{compliant} with the specification.
-\end{definition}
-
-The focus of this paper is to determine whether a \textit{given} sniffer trace
-$Tr$ is evidence of a violation.
-
-\subsection{Mutation Trace}
-\label{subsec:mutation}
-
-As shown in the motivation example in Fig.~\ref{fig:sniffer_in_middle}, a
-sniffer trace may either miss packets that are present in DUT trace, or contain
-extra packets that are missing in DUT trace. Note that in the latter case, those
-extra packets must be all sent \textit{to} the DUT. This is because it is
-impossible for the sniffer to overhear packets sent from the DUT that were not
-actually sent by the DUT.
-
-We formally capture this relationship with the definition of mutation trace.
-
-\begin{definition}[Mutation Trace]
-  \label{def:mutation}
-  A packet trace $Tr'$ is a mutation of sniffer trace $Tr$ w.r.t a DUT if for
-  all $(t, p) \in Tr \setminus Tr'$, $p.dest = DUT$, where $p.dest$ is the
-  destination of packet $p$.
-\end{definition}
-
-By definition, either $Tr' \supseteq Tr$ (hence $Tr \setminus Tr' = \emptyset$),
-or those extra packets in $Tr$ but not in $Tr'$ are all sent to the DUT. Note
-that $Tr'$ may contain extra packets that are either sent to or received by
-the DUT.
-
-A mutation trace $Tr'$ represents a \textit{guess} of the corresponding DUT
-packet trace given sniffer trace $Tr$.  In fact, the DUT packet trace must
-be one of the mutation traces of the sniffer trace $Tr$.
-
-\begin{lemma}
-  Let $Tr_{DUT}$ and $Tr$ be the DUT and sniffer packet trace captured during
-  the same protocol operation session, and $\mathcal{M}(Tr)$ be the set of
-  mutation traces of $Tr$ with respect to DUT, then $Tr_{DUT} \in \mathcal{M}(Tr)$.
-  \label{lem:mutation}
-\end{lemma}\begin{proof}
-  Let $\Delta = Tr \setminus Tr_{DUT}$ be the set of packets that are in $Tr$
-  but not in $Tr_{DUT}$. Recall that it is not possible for the sniffer to
-  observe packets sent from the DUT that the DUT did not send. Therefore,
-  all packets in $\Delta$ are sent \textit{to} the DUT. That is, for
-  all $(t, p) \in \Delta,\ p.dest = DUT$. By Definition~\ref{def:mutation},
-  $Tr_{DUT} \in \mathcal{M}(Tr)$.
-  \qed
-\end{proof}
-
-\subsection{Problem Statement}
-\label{subsec:problem}
-
-Lemma~\ref{lem:mutation} shows that $\mathcal{M}(Tr)$ is a \textit{complete} set
-of guesses of the DUT packet trace. Therefore, the problem of validating DUT
-implementation given a sniffer trace can be formally defined as follows:
-
-\begin{problem}
-  \label{prob:validation}
-  VALIDATION\\
-  \textbf{instance} A protocol monitor state machine $S$ and a sniffer trace $Tr$.\\
-  \textbf{question} Does there exist a mutation trace $Tr'$ of $Tr$ that satisfies $S$?
-\end{problem}
-
-If the answer is no, a definite violation of the DUT implementation can be
-claimed. Nevertheless, if the answer is yes, $S$ may still reject $Tr_{DUT}$.
-In other words, the conclusion of the validation can either be
-\textit{definitely wrong} or \textit{probably correct}, but not
-\textit{definitely correct}.  This is the fundamental limitation caused by the
-uncertainty of sniffer traces.
-
-\makeatletter{}\section{Validation Framework}
-\label{sec:framework}
-
-\makeatletter{}\subsection{Augmented State Machine}
-\label{subsec:augment}
-
-To deal with the inherent uncertainty of sniffer traces, we propose to
-systematically augment the original monitor state machine with non-deterministic
-transitions to account for the difference between the sniffer and DUT traces.
-
-Before formally defining the augmented state machine, we first use an example to
-illustrate the basic idea. Fig.~\ref{fig:augment} shows the augmented state
-machine for 802.11 transmitter state machine shown in
-Fig.~\ref{fig:dot11_tx_ta}.  For each existing transition (e.g., $s_0\rightarrow
-s_1$), we add an \textit{empty transition} with same clock guards and resetting
-clocks.  This accounts for the possibility when such packet was observed by
-the DUT but missed by the sniffer.  Additionally, for each transition triggered
-by a \textit{receiving} packet (i.e., $p.dest = DUT$), such as $s_1\rightarrow
-s_0$ and $s_2\rightarrow s_0$, we add a \textit{self transition} with the same
-trigger packet and clock guards, but an empty set of resetting clocks and no
-assignments to variables. This allows the state machine to make progress when
-the sniffer missed such packets.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=0.7\textwidth]{dot11_tx_checker.pdf}
-  \caption{\textbf{Augmented Monitor State Machine.} Augmented transitions are
-  highlighted in bold face. $\overline{Pkt}$ means either $\epsilon$ or $Pkt$.}
-  \label{fig:augment}
-\end{figure}
-
-There are two things to note. First, self transitions are added only for
-packets sent \textit{to} the DUT, since the sniffer will not overhear packets
-\textit{from} the DUT if they were not sent by the DUT. Second, no augmented
-transitions are added for the packets that are sent to DUT yet are missed by both
-the DUT and the sniffer, since such packets do not cause difference between the
-DUT and sniffer traces.
-
-The augmented state machine in Fig.~\ref{fig:augment} will accept the sniffer
-packet traces $Tr_1$ and $Tr_2$ shown in Fig.~\ref{fig:sniffer_in_middle}.  For
-instance, one accepting transition sequence on sniffer trace $Tr_1$ is
-$s_0\rightarrow s_1 \rightarrow_s s_1\rightarrow s_2 \rightarrow s_0$, and the
-sequence for $Tr_2$ is $s_0 \rightarrow s_1 \rightarrow_e s_2 \rightarrow s_0$,
-where $\rightarrow$ is the transition from the original state machine,
-$\rightarrow_e$ and $\rightarrow_s$ are the augmented empty and self transitions
-respectively.
-
-We now formally define the augmented state machine.
-
-\begin{definition}[Augmented Monitor]
-  \sloppy{
-  An augmented state machine $S^+$ for a monitor state machine $S$ is a 7-tuple
-  $\{\boldsymbol{\Sigma^+}, \mathbb{S}, \mathbb{X}, s_0, C, \boldsymbol{E^+},
-  G\}$, where $\mathbb{S}, \mathbb{X}, s_0, C, G$ are the same as $S$.
-  $\Sigma^+=\{\epsilon\} \cup \Sigma$ is the augmented input alphabet with the
-  empty symbol, and $E^+ \supset E$ is the set of transitions, which includes:
-}
-  \begin{itemize}
-    \item $E$: existing transitions (\textbf{Type-0}) in $S$.
-    \item $E^+_1$: empty transitions (\textbf{Type-1}) for transitions in $E$.
-    \item $E^+_2$: self transitions \textbf{(Type-2)} for transitions
-      triggered by receiving packets.
-  \end{itemize}
-\end{definition}
-
-\begin{figure}[t!]
-\begin{algorithm}[H]
-  \caption{Obtain Augmented Transitions $E^+$ from $E$}
-  \label{alg:augment}
-  \begin{algorithmic}[1]
-    \Function{augment}{$E$}
-    \Let{$E^+$}{$\emptyset$}
-    \ForAll{$\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$}
-      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, p, g, C'\rangle\}$\Comment{Type-0}}
-      \label{alg:augment:type0}
-      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, \boldsymbol{\epsilon}, g, C'\rangle\}$\Comment{Type-1}}
-      \label{alg:augment:type1}
-      \If{$p.dest = DUT$}
-      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, \boldsymbol{s_i},
-      \boldsymbol{v_i}, p, g, \boldsymbol{\emptyset}\rangle\}$\Comment{Type-2}}
-        \label{alg:augment:type2}
-      \EndIf
-    \EndFor
-    \State \Return{$E^+$}
-    \EndFunction
-  \end{algorithmic}
-\end{algorithm}
-\end{figure}
-
-Algorithm~\ref{alg:augment} describes the process of transforming $E$ into
-$E^+$. In particular, Line~\ref{alg:augment:type0} adds existing transitions in
-$E$ to $E^+$, while line~\ref{alg:augment:type1} and~\ref{alg:augment:type2} add
-Type-1 and Type-2 transitions to $E^+$ respectively.  We have highlighted the
-elements of the tuple that differ from the underlying Type-0 transition. Note
-that in Type-2 transitions, both the state and the variables stay the same after
-the transition.
-
-With augmented state machine $S^+$, we can use Type-1 transitions to
-non-deterministically infer packets missed by the sniffer, and use Type-2
-transitions to consume extra packets captured by the sniffer but missed by the
-DUT.
-
-A accepting run of $S^+$ on sniffer trace $Tr$ yields a mutation trace $Tr'$
-which represents one possibility of the DUT trace. Specifically, $Tr'$ can be
-obtained by adding missing packets indicated by Type-1 transitions to $Tr$, and
-removing extra packets indicated by Type-2 transitions from $Tr$
-
-We show that the VALIDATION problem is equivalent to the
-satisfiability problem of $Tr$ on $S^+$.
-
-\begin{theorem}
-  There exists a mutation trace $Tr' \in \mathcal{M}(Tr)$ that satisfies $S$ if
-  and only if $Tr$ satisfies $S^+$.
- \label{the:equivalent}
-\end{theorem}
-\begin{proof}
-  Assume $Tr$ satisfies $S^+$, and $P$ is a sequence of accepting transitions,
-  we construct a mutation trace $Tr'$ using $P$ and show that $Tr'$ satisfies
-  $S$.
-
-  \sloppy{    Initially, let $Tr'=Tr$, then for each \textit{augmented} transition $\langle s_i,
-  v_i, s_j, v_j, \sigma, g, C'\rangle \in P$:
-  }
-  \begin{itemize}
-    \item If this is a Type-1 transition, add $(t, p)$ to $Tr'$, where $t$ is a
-      timestamp that satisfies $g$ and $p$ is the missing packet.
-    \item If this is a Type-2 transition, remove corresponding $(t, p)$ from
-      $Tr'$.
-  \end{itemize}
-  It is obvious that $Tr'$ is a mutation trace of $Tr$, since only receiving
-  packets are removed in the process.
-
-  Now we show that there exists a accepting transition sequence $P'$ of $S^+$ on
-  input $Tr'$ that does not contain augmented transitions.  In particular, $P'$
-  can be obtained by substituting all Type-1 transitions with corresponding
-  original transitions, and removing all Type-2 transition.  Since $P'$ does not
-  contain augmented transitions, it is also an accepting transition sequence of
-  $S$ on input $Tr'$, hence $Tr'$ satisfies $S$.
-
-  On the other hand, assume $Tr' \in \mathcal{M}(Tr)$ and $Tr'$ satisfies $S$.
-  Suppose $P'$ is the accepting transition sequences of $S$ on input $Tr'$.
-  We first note that $P'$ is also the accepting transitions of $S^+$ on input
-  $Tr'$, since $E \subset E^+$.
-
-  We construct a accepting transition sequence $P$ of $S^+$ on input $Tr$ as
-  follows.
-  \begin{itemize}
-    \item For each packet $p \in Tr' \setminus Tr$, substitute the transition
-      $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle$ with the corresponding Type-1
-      transition $\langle s_i, v_i, s_j, v_j, \epsilon, g, C'\rangle$.
-    \item For each transition $\langle s_i, v_i, s_j, v_j, \sigma, g, C'\rangle$
-      followed by packet $p \in Tr\setminus Tr'$, add a Type-2 self
-      transition $\langle s_j, v_j, s_j, v_j, p, g, \emptyset\rangle$. This is
-      possible since $Tr'$ is a mutation trace of $Tr$, thus  for all $p \in Tr'
-      \setminus Tr$, $p.src \ne DUT$.
-  \end{itemize}
-  Therefore, $Tr$ satisfies $S^+$.
-  \qed
-\end{proof}
-
-By Theorem~\ref{the:equivalent}, the inherent uncertainty of the sniffer traces
-is explicitly represented by the augmented transitions, and can be
-systematically explored using the well established theory of state machine.
-
-\begin{comment}
-One immediate observation can be drawn from Theorem~\ref{the:equivalent} by
-contradiction.
-
-\begin{corollary}
-  If $S^+$ rejects $Tr$, then $S$ rejects $Tr_{DUT}$.
-  \label{cor:false_pos}
-\end{corollary}
-In the context of validation where we raise a violation alarm when $S^+$
-rejects $Tr$, Corollary~\ref{cor:false_pos} provides a {\it sufficient} condition for reporting incorrect traces. 
-However, when $S^+$ accepts $Tr$, $S$ could still reject $Tr_{DUT}$.
-In other words, the conclusion of the validation can either be \textit{definitely
-wrong} or \textit{possibly correct}, but not \textit{definitely correct}.
-This is the fundamental limitation caused by the uncertainty of sniffer traces.
-\end{comment}
-
-\makeatletter{}\subsection{Problem Hardness}
-\label{subsec:hard}
-
-In this section, we show that the VALIDATION problem is NP-complete. In
-fact, the problem is still NP-complete even with only one type of augmented
-transitions.
-
-Recall that Type-1 transitions are added because the sniffer may miss packets.
-Suppose an imaginary sniffer that is able to capture \textit{every} packet ever
-transmitted, then only Type-2 transitions are needed since the sniffer may still
-overhear packets sent to the DUT. Similarly, suppose another special sniffer
-that would not overhear any packets sent to the DUT, then only Type-1
-transitions are needed to infer missing packets.
-
-We refer the augmented state machine that only has Type-0 and Type-1
-transitions as $S^+_1$, and the augmented state machine that only has Type-0 and
-Type-2 transitions as $S^+_2$. And we show that each subproblem of determining
-trace satisfiability is NP-complete.
-
-\begin{problem}
-  VALIDATION-1\\
-  Given that $Tr\setminus Tr_{DUT}=\emptyset$ (sniffer does not overhear
-  packets).\\
-  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
-  \textbf{question} Does $S^+_1$ accept $Tr$?
-\end{problem}
-
-\begin{problem}
-  VALIDATION-2\\
-  Given that $Tr_{DUT} \subseteq Tr$ (sniffer does not missing packets).\\
-  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
-  \textbf{question} Does $S^+_2$ accept $Tr$?
-\end{problem}
-
-\begin{lemma}
-  \label{lem:sub}
-  Both VALIDATION-1 and VALIDATION-2 are NP-complete.
-\end{lemma}
-\begin{proof}
-  First, note that the length of mutation trace $Tr'$ is polynomial to the
-  length of $Tr$ because of the discrete time stamp and non-overlapping packets
-  assumption.
-  Therefore, given a state transition sequence as witness, it can be verified in
-  polynomial time whether or not it is an accepting transition sequence, hence
-  both VALIDATION-1 and VALIDATION-2 are in NP.
-
-  Next, we show how the SAT problem can be reduced to either one of the two
-  problems.
-  Consider an instance of SAT problem of a propositional formula $F$ with $n$
-  variables $x_0,x_1,\ldots, x_{n-1}$, we construct a corresponding protocol and
-  its monitor state machine as follows.
-
-  The protocol involves two devices: the DUT (transmitter) and the endpoint
-  (receiver).
-  The DUT shall send a series of packets, $pkt_0, pkt_1,\ldots, pkt_{n-1}$.
-  For each $pkt_i$, if the DUT receives the
-  acknowledgment packet $ack_i$ from the endpoint, it sets boolean variable
-  $x_i$ to be \texttt{true}.
-  Otherwise $x_i$ remains to be \texttt{false}.
-  After $n$ rounds, the DUT evaluate the formula $F$ using the assignments and
-  sends a special packet, $pkt_{true}$, if $F$ is \texttt{true}.
-  One round of the protocol operation can be completed in polynomial time since
-  any witness of $F$ can be evaluated in polynomial time.
-
-  \begin{figure}[t!]
-    \centering
-    \includegraphics[width=0.7\textwidth]{sat_sm.pdf}
-    \caption{\textbf{Monitor State Machine for SAT Problem.}}
-    \label{fig:sat}
-  \end{figure}
-
-  The protocol monitor state machine $S$ is shown in Fig.~\ref{fig:sat}.
-  Initially, all $x_i$ is set to \texttt{false}.
-  At state $s_0$, the DUT shall transmit $pkt_i$ within a unit time, transit to
-  $s_1$ and reset the clock along the transition.
-  At state $s_1$, either the DUT receives the $ack_i$ packet and
-  set $x_i$ to be \texttt{true} ($s_1 \rightarrow s_0$), or the DUT continues to
-  transmit the next packet $pkt_{i+1}$.
-  After $n$ rounds, the state machine is $s_0$ or $s_1$ depending on whether
-  $ack_{n-1}$ is received by the DUT.
-  In either case, the DUT shall evaluate $F$ and transmit $pkt_{true}$ if $F$ is
-  \texttt{true}.
-
-  \sloppy{    Consider a sniffer trace $Tr_1=\{(0, pkt_0), (2, pkt_1), (4, pkt_2),\ldots,
-  (2(n-1), pkt_{n-1}), (2n, pkt_{true})\}$.  That is, the sniffer only captures
-  all $pkt_i$ plus the final $pkt_{true}$, but none of $ack_i$.  It is easy to see
-  that $F$ is satisfiable if $S_1^+$ accepts $Tr_1$.  In particular, a successful
-  run of $S_1^+$ on $Tr_1$ would have to guess, for each $pkt_i$, whether the
-  Type-1 empty transitions should be used to infer the missing $ack_i$ packet,
-  such that $F$ is \textit{true} at the end.  Note that for $Tr_1$, no Type-2 self
-  transitions can be used since all packets in $Tr_1$ are sent from the DUT.
-  Therefore, the SAT problem of $F$ can be reduced to the VALIDATION-1 problem of
-  $S^+_1$ on sniffer trace $Tr_1$.
-  }
-
-  \sloppy{    On the other hand, consider another sniffer trace $Tr_2 =\{(0, pkt_0), (1,
-    ack_0), (2, pkt_1), (3, ack_1),\ldots, (2n-2, pkt_{n-1}), (2n-1, ack_{n-1}),\\ (2n,
-  pkt_{true}\}$. That is, the sniffer captures all $n$ pair of $pkt_i$ and $ack_i$
-  packets and the final $pkt_{true}$ packet.  Similar to $Tr_1$, $F$ is
-  satisfiable if $S^+_2$ accepts $Tr_2$. A successful transition sequence of
-  $S^+_2$ on $Tr_2$ must decide for each $ack_i$ packet, whether Type-2 self
-  transitions should be used, so that $F$ can be evaluated as true at the end.
-  Therefore, the SAT problem of $F$ can also be reduced to the VALIDATION-2
-  problem of $S^+_2$ on sniffer trace $Tr_2$.
-}
-
-Since SAT is known to be NP-complete, both the VALIDATION-1 and the
-VALIDATION-2 problem are also NP-complete.  \qed
-\end{proof}
-
-The hardness statement of the general VALIDATION problem naturally follows
-Lemma~\ref{lem:sub}.
-
-\begin{theorem}
-  VALIDATION is NP-complete.
-\end{theorem}
-
-\makeatletter{}\subsection{Searching Strategies}
-\label{subsec:search}
-
-In this section, we present an \textit{exhaustive} search algorithm of the
-accepting transition sequence of $S^+$ on sniffer trace $Tr$. It is guaranteed to
-yield an accepting sequence if there exists one, thus is exhaustive.  In the next
-sections, we present heuristics to limit the search to accepting sequences of
-$S^+$ that require relatively fewer transitions from $E^+_1 \cup E^+_2$.  Due to
-the NP-completeness of the problem, this also makes the algorithm meaningful in
-practice.
-
-\begin{figure}[t!]
-  \begin{minipage}{\textwidth}
-    \begin{algorithm}[H]
-      \caption{Exhaustive search algorithm of $S^+$ on $Tr$.}
-      \label{alg:search}
-
-      \begin{algorithmic}[1]
-        \Function{search}{S, Tr}
-          \Let{$S^+$}{\Call{augment}{S}}
-          \State\Return{\Call{extend}{[], Tr, $S^+.s_0$}}
-        \EndFunction
-        \Function{extend}{prefix, p::suffix, $s$}
-          \If{not \Call{likely}{prefix}}
-            \label{alg:search:stop}
-            {\Return{\nil}}\footnote{This check should be ignored in the exhaustive algorithm.}
-          \EndIf
-          \For{i $\in$ [0, 1, 2]}
-          \label{alg:search:order}
-          \Let{mutation}{EXTEND-i(prefix, p::suffix, $s$)}
-            \If{mutation $\ne$ \nil}
-              {\Return{mutation}}
-            \EndIf
-          \EndFor
-          \State\Return{\nil}
-        \EndFunction
-        \Function{extend-0}{prefix, p::suffix, $s$}
-          \For{$\langle s, s', p\rangle\footnote{$\langle s, s', p \rangle$ is
-              short for $\langle s, *, s', *, p, *, *\rangle$} \in E$}
-            \If{suffix = \nil}
-              {\Return{prefix@p}}
-              \label{alg:search:terminate_0}
-            \EndIf
-            \Let{mutation}{\Call{extend}{prefix@p, suffix, $s'$}}
-            \label{alg:search:append}
-            \If{mutation $\ne$ \nil}
-              {\Return{mutation}}
-            \EndIf
-          \EndFor
-          \State{\Return{\nil}}
-        \EndFunction
-        \Function{extend-1}{prefix, p::suffix, $s$}
-          \ForAll{$\langle s, s', q\rangle \in E^+_1$}
-            \If{q.time $>$ p.time}
-              \label{alg:search:type_1_limit}
-              \State{\textbf{continue}}
-            \EndIf
-            \Let{mutation}{\Call{extend}{prefix@q, p::suffix, $s'$}}
-            \label{alg:search:guess}
-            \If{mutation $\ne \nil$}
-              {\Return{mutation}}
-            \EndIf
-          \EndFor
-          \State{\Return{\nil}}
-        \EndFunction
-        \Function{extend-2}{prefix, p::suffix, $s$}
-          \ForAll{$\langle s, s, p \rangle \in E^+_2$}
-            \If{suffix = \nil}
-              {\Return{prefix}}
-              \label{alg:search:terminate_2}
-            \EndIf
-            \Let{mutation}{\Call{extend}{prefix, suffix, $s$}}
-            \label{alg:search:drop}
-            \If{mutation $\ne$ \nil}
-              {\Return{mutation}}
-            \EndIf
-          \EndFor
-          \State{\Return{\nil}}
-        \EndFunction
-      \end{algorithmic}
-    \end{algorithm}
-    \renewcommand\footnoterule{}
-  \end{minipage}
-\end{figure}
-
-The main routines of the algorithm are shown in Algorithm~\ref{alg:search}.  In
-the top level \texttt{SEARCH} routine, we first obtain the augmented state
-machine $S^+$, and then call the recursive \texttt{EXTEND} function with an empty
-prefix, the sniffer trace, and the $S^+$'s initial state. 
-In the \texttt{EXTEND} function, we try to consume the first packet in the
-remaining trace using either Type-0, Type-1 or Type-2 transition.
-Note that we always try to use Type-0 transitions before other two augmented
-transitions (line~\ref{alg:search:order}).
-This ensures the first found mutation trace will have the most number of Type-0
-transitions among all possible mutation traces.
-Intuitively, this means the search algorithm tries to utilize the sniffer's
-observation as much as possible before being forced to make assumptions.
-
-Each of the extend functions either returns the mutation trace $Tr'$, or
-\textit{\nil} if the search fails.
-In both \texttt{EXTEND-0} and
-\texttt{EXTEND-2} function, if there is a valid transition, we try to consume
-the next packet either by appending it to the prefix
-(line~\ref{alg:search:append}) or dropping it (line~\ref{alg:search:drop}).
-While in \texttt{EXTEND-1}, we guess a missing packet without consuming the next
-real packet (line~\ref{alg:search:guess}).
-Note that since only Type-0 and Type-2 consume packets, the recursion terminates
-if there is a valid Type-0 or Type-2 transition for the last packet
-(line~\ref{alg:search:terminate_0} and line~\ref{alg:search:terminate_2}).
-
-It is not hard to see that Algorithm~\ref{alg:search} terminates on any sniffer
-traces. Each node in the transition tree only has finite number of possible next
-steps, and the depth of Type-1 transitions is limited by the time available
-before the next packet (line~\ref{alg:search:type_1_limit}).
-
-\makeatletter{}\subsection{Pruning Heuristics}
-\label{subsec:heuristic}
-
-In the face of uncertainty between a possible
-protocol violation and sniffer imperfection, augmented transitions provide the
-ability to blame the latter. The exhaustive nature of
-Algorithm~\ref{alg:search} means that it always tries to blame sniffer
-imperfection whenever possible, making it reluctant to report true
-violations.
-
-Inspired by the \textit{directed model checking}~\cite{edelkamp2008survey}
-technique which is to mitigate the state explosion problem, we propose to enforce extra
-constraints on the mutation trace to restrict the search to only mutation traces
-with high likelihood. The modified \texttt{EXTEND} function checks certain
-likelihood constraints on the prefix of the mutation trace before continuing
-(line~\ref{alg:search:stop}), and stops the current search branch immediately if
-the prefix seems \textit{unlikely}.  Because of the recursive nature of the
-algorithm, other branches which may have a higher likelihood can then be
-explored.
-
-The strictness of the likelihood constraint represents a trade-off between
-precision and recall of validation. The more strict the constraints are, the
-more false positive violations will potentially be reported, hence the lower the
-precision yet higher recall. On the contrary, the more tractable the
-constraints are, the more tolerant the search is to sniffer imperfection, hence
-the more likely that it will report true violations, thus higher precision but
-lower recall.
-
-The exact forms of the constraints may depend on many factors, such as the
-nature of the protocol, properties of the sniffer, or domain knowledge.  Next,
-we propose two \textit{protocol oblivious} heuristics based on the sniffer loss
-probabilities and general protocol operations. Both heuristic contains
-parameters that can be fine tuned in practice.
-
-\subsubsection{$\mathit{NumMissing}(d, l, k)$}
-
-This heuristic states that the number of missing packets from device $d$ in any
-sub mutation traces of length $l$ shall not exceed $k$ ($k \le l$).  The sliding
-window of size $l$ serves two purposes. First, $l$ should be large enough for
-the calculated packet loss ratio to be statistically meaningful.  Second, it
-ensures that the packet losses are evenly distributed among the entire packet
-trace.
-
-The intuition behind this heuristic is that the sniffer's empirical packet loss
-probability can usually be measured before validation. Therefore, the likelihood
-that the sniffer misses more packets than prior measured loss ratio is quite
-low. The value of $l$ and $k$ can then be configured such that $k/l$ is marginally
-larger than the measured ratio.
-
-\begin{comment}
-In addition, we note that for a fixed $l$, a single threshold $k$ may not work
-for traces collected by sniffers with different packet loss probabilities.
-Intuitively, sniffers with high loss probabilities require a larger threshold.
-Therefore, we perform a \textit{stratified} search with an increasing sequence
-of thresholds $\{k_1, k_2,\ldots, k_n\}$ ($k_i < k_{i+1}$ for $1 \le i < n$).
-At round $i$, the search is restricted by constraint $NumMissing(d, l, k_i)$.
-If a mutation trace is found, the search is completed.  Otherwise, we increase
-the round number and repeat the search process.  In this way, this heuristic can
-automatically adapt to the underlying sniffer packet loss probabilities. And a
-violation is declared when no mutation trace can be found with $NumMissing(d, l,
-k_n)$.
-\end{comment}
-
-\subsubsection{$\mathit{GoBack}(k)$}
-
-This heuristic states that the search should only
-backtrack at most $k$ steps when the search gets stuck using only $E$.
-The motivation is that many protocols operate as a sequence of independent
-transactions, and the uncertainty of previous transactions often do not affect
-the next transaction.
-For instance, in 802.11 packet transmission protocol, each packet exchange,
-include the original, retransmission and acknowledgment packets, constitute a
-transaction.
-And the retransmission status of previous packets has no effect on the packets
-with subsequent sequence numbers, hence need not be explored when resolving the
-uncertainty of the packets with new sequence numbers.  Note that we do not
-require the protocol to specify an exact transaction boundary, but only need $k$
-to be sufficiently large to cover a transaction.
-
-\begin{comment}
-\subsection{Confidence of Correctness}
-
-A run of $S^+$ on sniffer packet trace $Tr$ may yield two conclusions about the
-DUT implementation: either ``definitely wrong'' when $S^+$ rejects $Tr$, or
-``possibly correct'' when $S^+$ accepts $Tr$. We now quantify the
-\textit{confidence} of the implementation correctness in the later case.
-
-Note that the augmented transitions in $S^+$ implicitly embed the assumptions
-about packet losses. In particular, Type-1 transitions assume the sniffer
-missed certain packets, while Type-2 transitions assume the sniffer
-overheard packets that were missed by the DUT. Intuitively, the portion of such
-assumptions (hence packet loss \textit{ratios}) in an accepting transition
-sequence should match with the true packet loss \textit{probabilities}.
-Otherwise, the correctness confidence, despite $S^+$ accepts $Tr$, is low.
-
-Suppose $Tr$ is the sniffer trace and $Tr'$ is its mutation that satisfies $S$.
-We can calculate the three packet loss ratios in our model shown in
-Figure~\ref{fig:prob}.
-
-\begin{align}
-  &Pr_{ds} = \frac{\parallel \{(t,p) \in Tr'\setminus Tr \land p.src = DUT\}\parallel}
-  {\parallel \{(t,p) \in Tr' \cup Tr \land p.src = DUT\}\parallel}\\
-  &Pr_{es} = \frac{\parallel \{(t,p) \in Tr'\setminus Tr \land p.src = EP\}\parallel}
-  {\parallel \{(t,p) \in Tr' \cup Tr \land p.src = EP\}\parallel}
-\end{align}where $Tr'\setminus Tr$ is the set of inferred packets that were missed by the
-sniffer (Type-1 transitions in $S^+$).
-
-Note that we can not directly calculate the packet loss ratio from Endpoint to
-DUT, since there might be packets that were missed by both the sniffer and the
-DUT. We can, however, approximate $Pr_{ed}$ using conditional packet loss
-probability.
-
-\begin{align}
-  \nonumber
-  &(1-Pr_{es})\times Pr_{ed} =\\
-  &\frac{\parallel \{(t,p) \in Tr\setminus Tr' \land p.src = EP \land p.dest = DUT\}\parallel}
-  {\parallel \{(t,p) \in Tr \cup Tr' \land p.src = EP \land p.dest = DUT\}\parallel}
-\end{align}The LHS is the probability that a packet sent from the Endpoint to the DUT was
-missed by the DUT yet captured by the sniffer. And the RHS is the ratio of
-such packets in the mutation trace.
-
-Consider the 802.11 data transmission protocol and the example trace snippet
-shown in Figure~\ref{fig:loss}, the packet loss ratios are:
-\begin{align*}
-  Pr_{ds} &= \frac{\parallel\{\it{DATA}'_0\}\parallel}{\parallel\{\it{DATA}_0, \it{DATA}'_0\}\parallel} = 0.5\\
-  Pr_{es} &= \frac{\parallel\emptyset\parallel}{\parallel\{Ack, Ack\}\parallel} = 0\\
-  Pr_{ed} &= \frac{\parallel\{Ack\}\parallel}{\parallel\{Ack, Ack\}\parallel\times (1-Pr_{es})} = 0.5
-\end{align*}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=0.4\textwidth]{loss.pdf}
-  \caption{\textbf{Example Sniffer Trace and Its Mutation Trace.}}
-  \label{fig:loss}
-\end{figure}
-
-Suppose we have the true packet loss probabilities $\overline{Pr_{ds}},
-\overline{Pr_{es}}$ and $\overline{Pr_{ed}}$, the \textit{confidence} of this
-mutation trace $Tr'$ can be defined as:
-
-\begin{align}
-  Confidence(Tr') = \prod_{i\in\{ds, es, ed\}}(1-|Pr_i-\overline{Pr_i}|)
-\end{align}
-
-By this definition, the trace confidence maximizes when all packet loss
-ratios calculated from $Tr'$ matches the true packet
-loss probabilities.
-
-Suppose we have $M$ mutation traces that satisfies $S$, then
-the overall \textit{confidence of correctness} from sniffer trace $Tr$ can be defined as:
-
-\begin{align}
-  Confidence(Tr) &= 1-\prod_{j=1}^{M}(1-Pr(Tr'_j))
-\end{align}
-
-Since $0 \le 1-Pr(Tr'_j) \le 1$ for all $j$, we have:
-
-\begin{align}
-  \nonumber
-  Confidence(Tr) &\ge 1-(1-Pr(Tr'_j)\\
-     &= Pr(Tr'_j)
-\end{align}for all $j$.
-
-This means if we can find a mutation trace $Tr'$ that satisfies $S$, then the
-overall confidence of the correctness is no less than $Pr(Tr')$.
-\end{comment}
-
-\makeatletter{}\section{Case Studies}
-\label{sec:case}
-
-We present case studies on applying our validation framework on two protocols
-implemented in the \ns{} network simulator: 802.11 data transmission and ARF
-rate control algorithm. The goal is to demonstrate how our framework can avoid
-false alarms and report true violations on incomplete sniffer traces and report
-true violations. We also report the application of our framework to a commercial
-product under development.
-
-\makeatletter{}\subsection{802.11 Data Transmission}
-\label{subsec:tx}
-
-In this section, we first show that our framework can improve validation
-precision by inferring the missing or extra packets using the augmented
-transition framework. We then demonstrate the ability of our framework to detect
-true violations by manually introducing bugs in the \ns{} implementation and
-show the precision and recall of validation results.
-
-\subsubsection{Experimental Setup}
-
-We set up two \wifi{} devices acting as the transmitter (DUT) and receiver
-respectively. Another \wifi{} device is configured in monitor mode and acts as
-the sniffer. During the experiments,we collect both the DUT packet trace
-(the ground truth) and the sniffer trace.
-
-\subsubsection{Verifying Unmodified Implementation}
-
-In the original monitor state machine shown in Fig.~\ref{fig:dot11_tx_ta}, we
-set acknowledgment timeout $T_o=334\mu$s, maximum retransmission delay
-$T_m=15$ms according to the protocol. We also adapt the state machine to include
-multiple retransmissions\footnote{The exact number of retransmissions is not
-part of the protocol, and NS-3 implementation set this to be 7.} instead of one.
-
-Let $Pr_{ds}$, $Pr_{es}$ and $Pr_{ed}$ be the packet loss probability between
-the DUT and sniffer, endpoint and sniffer, endpoint and DUT respectively.
-$Pr_{ed}$ represents the characteristics of the system being tested, while
-$Pr_{ds}$ and $Pr_{es}$ represent the sniffer's quality in capturing packets.
-
-We vary each of the three probabilities, $Pr_{ds}$, $Pr_{es}$ and $Pr_{ed}$, from
-0 to 0.5 (both inclusive) with 0.05 step.  For each loss ratio combination, we
-ran the experiment 5 times, and each run lasted
-30 seconds. In total, 6655 ($11^3\times 5$) pairs of DUT and sniffer packet
-traces were collected.
-
-To establish the ground truth of violations, we first verify the DUT packet
-traces using the \textit{original} state machine $S$.  This can be achieved by
-disabling augmented transitions in our framework.  As expected, no violation is
-detected in any DUT packet traces.
-
-We then verify the sniffer traces using the augmented state machine $S^+$.  For
-the $\mathit{GoBack}(k)$ heuristic, we set $k=7$, which is the maximum number of
-transmissions of a single packet. For the $\mathit{NumMissing}(d, l, k)$ heuristic, we
-set the sliding window size $l=100$, and $k=80$ such that no violation is
-reported. The relationship of $k$ and validation precision is studied in next
-section.
-
-Next, we present detailed analysis of the augmented transitions on the sniffer
-traces. The goal is to study for a given system packet loss probability
-$Pr_{ed}$, how the sniffer packet loss properties ($Pr_{ds}$ and $Pr_{es}$)
-affect the difference between the DUT trace and the mutation trace, which represents
-a guess of the DUT trace by the augmented state machine based on the sniffer
-trace.
-
-For all following analysis, we divide the traces into three groups according to
-$Pr_{ed}$: low ($0 \le Pr_{ed} \le 0.15$), medium ($0.20 \le Pr_{ed} \le 0.35$)
-and high ($0.40 \le Pr_{ed} \le 0.50$).
-
-The different between two packet traces can be quantified by the Jaccard distance
-metric.
-
-\begin{align}
-  Jaccard(Tr_1, Tr_2) = \frac{\left\vert Tr_1 \ominus Tr_2\right\vert}{\left\vert
-  Tr_1 \cup Tr_2\right\vert}
-\end{align}where $\ominus$ is the symmetric difference operator.  The distance is 0 if the
-two traces are identical, and is 1 when the two traces are completely different.
-The smaller the distance is, the more similar the two traces are.
-
-A na\"ive way to calculate the Jaccard distance of two traces is to use the hash
-of the $(time, packet)$ pair for set intersection and union operation.  However,
-it does not work for mutation trace which contains fabricated packets with no
-actual payload.  Therefore, we use a protocol specific canonical representation
-of packets when calculating the distance.  In particular, the string
-\texttt{r\_\it{DATA}\_i\_t} represents the $t^{th}$ transmission of a data packet
-with sequence number $i$, and $r$ represents the round of sequence numbers as it
-wraps after 4096.  And similarly \texttt{r\_Ack\_i\_t} is the corresponding
-acknowledgment packet.
-
-\begin{figure*}[t!]
-  \centering
-  \begin{subfigure}{0.33\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_10.pdf}
-        \caption{$0.05 \le Pr_{ed} \le 0.15$}
-  \end{subfigure}\hspace*{0.01\textwidth}
-  \begin{subfigure}{0.33\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_30.pdf}
-    \caption{$0.2 \le Pr_{ed} \le 0.35$}
-  \end{subfigure}\hspace*{0.01\textwidth}
-  \begin{subfigure}{0.33\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{MutationDUTJaccard3DFigure_0_50.pdf}
-    \caption{$0.4 \le Pr_{ed} \le 0.5$}
-  \end{subfigure}\hspace*{0.01\textwidth}
-  \caption{\textbf{Jaccard Distance Between Mutation and DUT Traces.} For each
-  data point, the mean of the 5 runs is used.}
-  \label{fig:mutation_dut}
-\end{figure*}
-
-Fig.~\ref{fig:mutation_dut} shows the Jaccard Distance between mutation and its
-corresponding DUT trace. We make the following observations. First, for a given
-system loss probability $Pr_{ed}$ (each sub-figure), the lower the sniffer
-packet loss probability ($Pr_{ds}$ and $Pr_{es}$), the smaller Jaccard distance
-between the DUT and mutation trace. Intuitively, this means a sniffer that
-misses less packets can enable our framework to better reconstruct the DUT
-trace.
-
-Second, we observe a \textit{protocol-specific} trend that $Pr_{ds}$ is more
-dominant than $Pr_{es}$.  This is because retransmission packets of the same
-sequence number are identical, hence when the sniffer misses multiple
-retransmission packets, our framework only needs to infer one retransmission
-packet to continue state machine execution.
-
-Finally, as the system loss probability $Pr_{ed}$ increases, the Jaccard
-distance increases more rapidly as $Pr_{ds}$ increases.  This is because the
-ratio of retransmission packet increases along with $Pr_{ed}$.
-
-We then evaluate the cost of resolving uncertainty. In particular, we use
-the Average Search Steps Per Packet (ASSPP) as a metric to quantify the search
-cost.  It is calculated by dividing the total number of search steps by the
-number of packets in the packet trace. For DUT traces, ASSPP is always 1 since
-there is no uncertainty. For sniffer traces, however, multiple search steps must
-be to conducted to resolve the potential uncertainty of each packet in the
-sniffer trace.
-
-\begin{figure*}[t!]
-  \centering
-  \begin{subfigure}{0.33\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{StepCount3DFigure_0_10.pdf}
-    \caption{$Pr_{ed} \in [0.05, 0.10, 0.15]$}
-  \end{subfigure}\hspace*{0.01\textwidth}
-  \begin{subfigure}{0.33\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{StepCount3DFigure_0_30.pdf}
-    \caption{$Pr_{ed} \in [0.2, 0.25, 0.3, 0.35]$}
-  \end{subfigure}\hspace*{0.01\textwidth}
-  \begin{subfigure}{0.33\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{StepCount3DFigure_0_50.pdf}
-    \caption{$Pr_{ed} \in [0.40, 0.45, 0.50]$}
-  \end{subfigure}
-  \caption{\textbf{Average Searching Step Per Packet.} For each data point, the mean of 5 runs
-  is used.}
-  \label{fig:cost}
-\end{figure*}
-
-Fig.~\ref{fig:cost} shows the distribution of ASSPP at different $Pr_{ed}$.
-Similar to the case in Fig.~\ref{fig:mutation_dut}, $Pr_{ds}$ plays a dominant
-role in determine the searching cost. One interesting observation for this
-particular protocol is that the
-search cost peaks when $Pr_{ds}$ is high while $Pr_{es}$ is low. In such loss
-probability combinations, sniffer misses many data packets from the DUT but
-picks up lots of \textit{dangling} $Ack$ packets from the DUT. Because the $Ack$
-packet has neither sequence numbers nor retry flag, the searching algorithm had
-a hard time resolving such uncertainty.
-
-\subsubsection{Introducing Bugs}
-
-We have demonstrated that our framework can tolerate sniffer imperfections and
-avoid raising false alarms.  The next question is, can it detect true
-violations?  To answer this question, we manually introduce several bugs in
-\ns{} implementation that concerns various aspects of 802.11 data transmission
-protocol.  More specifically, the bugs are:
-
-\begin{itemize}
-  \item \textbf{Sequence Number}: the DUT does not assign sequence number
-    correctly. For example, it may increase sequence by 2 instead of 1, or it
-    does not increase sequence number after certain packet, etc. We choose one
-    type of such bugs in each run.
-  \item \textbf{Semantic}: the DUT may retransmit even
-    after receiving $Ack$, or does not retransmit when not receiving $Ack$.
-\end{itemize}
-
-We instrument the \ns{} implementation to embed instances of bugs in each
-category. At each experiment run, we randomly decide whether and which bug to
-introduce for each category. We fix $Pr_{ds}=Pr_{es}=0.1$ and vary $Pr_{ed}$
-from 0.0 to 0.5 with 0.01 step. For each $Pr_{ed}$ value, we ran the experiment
-100 times, of which roughly 75 experiments contained bugs. In total, 5100 pairs of
-DUT and sniffer traces were collected.
-
-We use the DUT packet traces as ground truth of whether or not each experiment
-run contains bugs.
-For each $Pr_{ed}$ value, we calculate the precision and recall of violation
-detection using the sniffer traces.\begin{align}
-  \text{Precision} = \frac{\left\vert \{\text{Reported Bugs}\} \cap \{\text{True Bugs}\}\right\vert}{\left\vert
-  \{\text{Reported Bugs}\}\right\vert}\\
-  \text{Recall} = \frac{\left\vert \{\text{Reported Bugs}\} \cap \{\text{True Bugs}\}\right\vert}{\left\vert
-  \{\text{True Bugs}\}\right\vert}
-\end{align}The precision metric quantifies how \textit{useful} the validation results are ,
-while the recall metric measures how \textit{complete} the validation results
-are.
-
-\begin{figure}[t!]
-  \centering
-  \begin{subfigure}{0.48\textwidth}
-    \includegraphics[width=\textwidth]{PrecisionFigure.pdf}
-  \end{subfigure}
-  \begin{subfigure}{0.48\textwidth}
-    \includegraphics[width=\textwidth]{RecallFigure.pdf}
-  \end{subfigure}
-  \caption{\textbf{Precision and Recall of Validation Results.}}
-  \label{fig:precision}
-\end{figure}
-
-Fig.~\ref{fig:precision} shows the CDF of precision and recall of the 51
-experiments for various $k$ values. For precision, as expected, the more
-tolerant the search to sniffer losses (larger $k$), the more tolerant the
-framework is to sniffer losses, and the more precise the violation detection. In
-particular, when $k=30$, the precisions are 100\% for all $Pr_{ed}$ values.
-Second, the recall is less sensitive to the choice of $k$.  Except for the
-extreme case when $k=30$, all other thresholds can report almost all the
-violations.
-
-\makeatletter{}\subsection{ARF Rate Control Algorithm}
-\label{subsec:arf}
-
-Automatic Rate Fallback (ARF)~\cite{kamerman1997wavelan} is the first rate
-control algorithm in literature. In ARF, the sender increases the bit rate after
-$Th_1$ number of consecutive successes or $Th_2$ number of packets with at most
-one retransmission. The sender decreases bit rate after two consecutive packet
-failures or if the first packet sent after rate increase (commonly referred as
-\textit{probing} packet) fails.
-
-\sloppy{  Fig.~\ref{fig:arf_sm} shows the state machine $S$ for the packet trace
-  collected at sender (DUT), where $\it{DATA}_i^r$ denotes a data packet
-  with sequence number $i$ and bit rate $r$, $\it{DATA}_i^{r\prime}$ is a retransmission packet
-  and $Ack$ is the acknowledgment packet. The \texttt{pkg\_succ} function is
-  shown in Algorithm~\ref{alg:pkt_succ}.
-}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=0.8\textwidth]{arf_sm.pdf}
-  \caption{\textbf{Monitor State Machine for ARF Rate Control Algorithm.} Timing
-  constraints are omitted for succinctness.}
-  \label{fig:arf_sm}
-\end{figure}
-
-The \texttt{succ} variable is used to track the number of consecutive packet
-successes. It is increased after each packet success , and is reset to 0 after a
-rate increase or upon a packet failure ($s_1\rightarrow s_2$).  Similarly,
-\texttt{count} is to track the number of packets with at most one
-retransmission, and is increased after packet success, or for the first packet
-retransmission ($s_1\rightarrow s_2$). It is reset when there are two
-consecutive packet failures ($s_2\rightarrow s_3$). Finally, the \texttt{probe}
-flag is set upon rate increases to indicate the probing packet, and is cleared
-upon packet success. The variable \texttt{r} is the current bit rate, which is
-decreased if the probing packet fails ($s_1\rightarrow s_4$), or every two
-consecutive failures ($s_2\rightarrow s_3$). If \texttt{r} is not the highest
-rate, it is increased when either of the two thresholds are reached.
-
-\begin{algorithm}[t!]
-  \caption{\texttt{pkt\_succ} function}
-  \label{alg:pkt_succ}
-  \begin{algorithmic}[1]
-    \Function{pkt\_succ}{}
-    \Let{i}{(i+1)\%N}
-    \Let{succ}{succ + 1}
-    \Let{count}{count + 1}
-    \Let{probe}{false}
-    \If{r $< R$ and (succ $\ge Th_1$ or count $\ge Th_2$)}
-    \Let{r}{r+1}
-    \Let{succ}{0}
-    \Let{count}{0}
-    \Let{probe}{true}
-    \EndIf
-    \EndFunction
-  \end{algorithmic}
-\end{algorithm}
-
-In particular, the bug we found lies in the implementation's \texttt{pkt\_succ}
-function in line 6. Instead of checking \texttt{count $\ge$ Th\_2}, the
-implementation checks \texttt{count == Th\_2}. This bug also exists in the NS-3
-implementation of Adaptive ARF (AARF) algorithm~\cite{lacage2004ieee} and the
-pseudo code implementation of AARF in~\cite{lacage2004report}.
-
-Note that the \texttt{count} variable is incremented twice if a packet succeed
-after one retransmission: once in $s_1\rightarrow s_2$, once in the
-\texttt{pkt\_succ} function for the retransmission packet. Therefore, if the
-value of \texttt{count} is $Th_2-1$ and the next packet succeed after one
-retransmission, the value of \texttt{count} will be $Th_2+1$, which would fail
-the implementation's test of \texttt{count == Th\_2}.
-
-We report a bug found in NS-3 ARF~\cite{kamerman1997wavelan} implementation
-which causes the sender to get stuck at a lower rate even after enough number of
-consecutive successes. The bug was detected using sniffer traces and
-confirmed by both the DUT trace and source code inspection.
-
-\makeatletter{}\subsection{Industrial Application}
-\label{subsec:xbox}
-
-\begin{table}[t!]
-  \centering
-  \begin{tabular}{lll}
-    \toprule
-    \textbf{Protocol Aspects} & \textbf{Traces} & \textbf{Violations (\%)}\\
-    \midrule
-    Sequence Number & 3049 & 1539 (50.48\%) \\
-    Station Scheduling & 3046 & 2045 (67.14\%) \\
-    Uplink Modulation & 3127 & 8 (0.26\%) \\
-    Downlink Modulation & 3127 & 24 (0.77\%) \\
-    \bottomrule
-  \end{tabular}
-  \caption{\textbf{Validation Results on Traces from the Gaming Wireless Protocol.}}
-  \label{tab:xbox}
-\end{table}
-
-We have applied our framework for runtime verification of the wireless protocol
-implementation of a commercial product that has several millions of shipping
-devices. The product runs a proprietary wireless protocol to reduce latency and
-power consumption. Sniffer traces were collected in regular testing process, but
-were only manually inspected previously.
-
-We obtained 75 sniffer traces from the testing team for a new version of the
-protocol that is under development and testing. This team has been testing the
-implementation for a few weeks.  Each trace contained around 6 million packets
-that were captured during 1 hour and 40 minutes of protocol operation.
-
-We split the traces into \num{100000} packet segments, and applied our framework
-to validate the DUT implementation. We found that the latest implementation of
-the protocol under development had violations of the protocol specification.
-Some of the implementation bugs we found related to the sequence number
-management, station scheduling during power saving mode, and modulation rates
-adaptation.
-
-Table~\ref{tab:xbox} summarizes the validation result. Note that if we disable
-the augmented transitions, each trace will be flagged as violation because of
-the missing packets, thereby reducing the usability of the tool. We also note
-that some bugs manifest more often than others. For instance, the bugs related
-to packet sequence number and station scheduler were detected in about half the
-traces, and the bug related to the rate control algorithm was detected in only a
-few traces. This is because the previous two aspects are essential in all
-protocol operations, while the bugs related to rate control only manifest
-themselves under certain link conditions.
-
-After communicating with the testing team, we confirmed that the sequence number
-bug was already known, as it is relatively easy to detect even by manually
-examining the traces. The bug related to station scheduling was also noticed
-before, yet no quantitative results about how frequent this bug appears were
-obtained because of the lack of automatized validating tools. Finally, the bug
-related to rate control has been filed as a bug report. All reported bugs were
-fixed before the next release the product.
-
-\makeatletter{}\section{Related Work}
+\section{Related Work}
 \label{sec:related}
@@
 verification purpose.
+\end{comment}
 
@@
 sampling does not lead to spurious alarms.
-\end{comment}
-
-\makeatletter{}\section{Conclusions}
-\label{sec:conclusion}
-
-We formally define the uncertainty problem in validating wireless protocol
-implementations using sniffers. We describe a systematic augmentation of the
-protocol state machine to explicitly encode the uncertainty of sniffer traces.
-We characterize the NP-completeness of the problem and propose both an
-exhaustive search algorithm and heuristics to restrict the search to more likely
-traces. We present two case studies using \ns{} network simulator to demonstrate
-how our framework can improve validation precision and detect real bugs. We also
-report the application of our framework on a commercial product under
-development.
-
-Finally, we discuss a few challenges and future
-directions.
-
-\textbf{Verification Coverage.} Given a single sniffer trace, it is possible
-that not all the states in the state machine are visited during the verification
-process. For instance, a rate control state machine based on certain consecutive
-packet losses patterns can not be verified if no such consecutive losses appear
-in the sniffer trace. In general, given a protocol state machine, it is
-challenging to extract the packet patterns for each state to be reached and to
-alter the testing such that such patterns can be observed.
-
-\textbf{Automated State Machine Construction.} We manually constructed the
-protocol monitor state machines for the protocols studied in this paper based on
-the source code, comments and documentation. The process involves lots of labor
-effort and is time-consuming. A potential alternative is to automatically learn
-the sketch of the monitor state machines from the sniffer traces. Domain
-knowledge can then be leveraged to improve the sketch state machines.
-
-{\footnotesize
-  \bibliographystyle{abbrv}
-  \begin{thebibliography}{10}
-
-\bibitem{alur1994theory}
-R.~Alur and D.~L. Dill.
-\newblock A theory of timed automata.
-\newblock {\em Theoretical computer science}, 126(2):183--235, 1994.
-
-\bibitem{arnold2008qvm}
-M.~Arnold, M.~Vechev, and E.~Yahav.
-\newblock Qvm: an efficient runtime for detecting defects in deployed systems.
-\newblock In {\em ACM Sigplan Notices}, volume~43, pages 143--162. ACM, 2008.
-
-\bibitem{bahl2006enhancing}
-P.~Bahl, R.~Chandra, J.~Padhye, L.~Ravindranath, M.~Singh, A.~Wolman, and
-  B.~Zill.
-\newblock Enhancing the security of corporate \wifi{} networks using {DAIR}.
-\newblock In {\em Proceedings of the 4th international conference on Mobile
-  systems, applications and services}, pages 1--14. ACM, 2006.
-
-\bibitem{basin2012monitoring}
-D.~Basin, F.~Klaedtke, S.~Marinovic, and E.~Z{\u{a}}linescu.
-\newblock Monitoring compliance policies over incomplete and disagreeing logs.
-\newblock In {\em International Conference on Runtime Verification}, pages
-  151--167. Springer, 2012.
-
-\bibitem{bonakdarpour2011sampling}
-B.~Bonakdarpour, S.~Navabpour, and S.~Fischmeister.
-\newblock Sampling-based runtime verification.
-\newblock In {\em FM 2011: Formal Methods}, pages 88--102. Springer, 2011.
-
-\bibitem{cheng2006jigsaw}
-Y.-C. Cheng, J.~Bellardo, P.~Benk{\"o}, A.~C. Snoeren, G.~M. Voelker, and
-  S.~Savage.
-\newblock {\em Jigsaw: solving the puzzle of enterprise 802.11 analysis},
-  volume~36.
-\newblock ACM, 2006.
-
-\bibitem{wifried}
-M.~Ciabarra.
-\newblock {WiFried: iOS 8 WiFi Issue}.
-\newblock \url{https://goo.gl/KtRDqk}.
-
-\bibitem{das-cav15}
-A.~Das, S.~K. Lahiri, A.~Lal, and Y.~Li.
-\newblock Angelic verification: Precise verification modulo unknowns.
-\newblock In {\em Computer Aided Verification - 27th International Conference,
-  {CAV} 2015, San Francisco, CA, USA, July 18-24, 2015, Proceedings, Part {I}},
-  pages 324--342, 2015.
-
-\bibitem{surface}
-digitalmediaphile.
-\newblock Windows 10 wifi issues with surface pro 3 and surface 3.
-\newblock \url{http://goo.gl/vBqiEo}.
-
-\bibitem{edelkamp2008survey}
-S.~Edelkamp, V.~Schuppan, D.~Bo{\v{s}}na{\v{c}}ki, A.~Wijs, A.~Fehnker, and
-  H.~Aljazzar.
-\newblock Survey on directed model checking.
-\newblock In {\em International Workshop on Model Checking and Artificial
-  Intelligence}, pages 65--89. Springer, 2008.
-
-\bibitem{fei2006artemis}
-L.~Fei and S.~P. Midkiff.
-\newblock Artemis: Practical runtime monitoring of applications for execution
-  anomalies.
-\newblock In {\em ACM SIGPLAN Notices}, volume~41, pages 84--95. ACM, 2006.
-
-\bibitem{lollipop}
-Gizmodo.
-\newblock The worst bugs in android 5.0 lollipop and how to fix them.
-\newblock \url{http://goo.gl/akDcvA}.
-
-\bibitem{godefroid1997model}
-P.~Godefroid.
-\newblock Model checking for programming languages using verisoft.
-\newblock In {\em Proceedings of the 24th ACM SIGPLAN-SIGACT symposium on
-  Principles of programming languages}, pages 174--186. ACM, 1997.
-
-\bibitem{iris}
-Google.
-\newblock Google contact lens.
-\newblock \url{https://en.wikipedia.org/wiki/Google_Contact_Lens}.
-
-\bibitem{hauswirth2004low}
-M.~Hauswirth and T.~M. Chilimbi.
-\newblock Low-overhead memory leak detection using adaptive statistical
-  profiling.
-\newblock In {\em Acm Sigplan Notices}, volume~39, pages 156--164. ACM, 2004.
-
-\bibitem{kamerman1997wavelan}
-A.~Kamerman and L.~Monteban.
-\newblock Wavelan{\textregistered}-ii: a high-performance wireless lan for the
-  unlicensed band.
-\newblock {\em Bell Labs technical journal}, 2(3):118--133, 1997.
-
-\bibitem{lacage2004ieee}
-M.~Lacage, M.~H. Manshaei, and T.~Turletti.
-\newblock {IEEE 802.11 rate adaptation: a practical approach}.
-\newblock In {\em Proceedings of the 7th ACM international symposium on
-  Modeling, analysis and simulation of wireless and mobile systems}, pages
-  126--134. ACM, 2004.
-
-\bibitem{lacage2004report}
-M.~Lacage, M.~H. Manshaei, and T.~Turletti.
-\newblock {IEEE 802.11 rate adaptation: a practical approach}.
-\newblock {\em [Research Report] RR-5208}, (<inria-00070784>):25, 2004.
-
-\bibitem{lee1997passive}
-D.~Lee, A.~N. Netravali, K.~K. Sabnani, B.~Sugla, and A.~John.
-\newblock Passive testing and applications to network management.
-\newblock In {\em Network Protocols, 1997. Proceedings., 1997 International
-  Conference on}, pages 113--122. IEEE, 1997.
-
-\bibitem{mahajan2006analyzing}
-R.~Mahajan, M.~Rodrig, D.~Wetherall, and J.~Zahorjan.
-\newblock Analyzing the mac-level behavior of wireless networks in the wild.
-\newblock In {\em ACM SIGCOMM Computer Communication Review}, volume~36, pages
-  75--86. ACM, 2006.
-
-\bibitem{musuvathi2002cmc}
-M.~Musuvathi, D.~Y. Park, A.~Chou, D.~R. Engler, and D.~L. Dill.
-\newblock {CMC}: A pragmatic approach to model checking real code.
-\newblock {\em ACM SIGOPS Operating Systems Review}, 36(SI):75--88, 2002.
-
-\bibitem{mytkowicz2008observer}
-T.~Mytkowicz, P.~F. Sweeney, M.~Hauswirth, and A.~Diwan.
-\newblock Observer effect and measurement bias in performance analysis.
-\newblock 2008.
-
-\bibitem{riley2010ns}
-G.~F. Riley and T.~R. Henderson.
-\newblock The ns-3 network simulator.
-\newblock In {\em Modeling and Tools for Network Simulation}, pages 15--34.
-  Springer, 2010.
-
-\bibitem{omnipeek}
-{Savvius Inc.}
-\newblock Savvius \wifi{} adapters.
-\newblock \url{https://goo.gl/l3VXSx}.
-
-\bibitem{Shi2016}
-J.~Shi, S.~K. Lahiri, R.~Chandra, and G.~Challen.
-\newblock {\em Wireless Protocol Validation Under Uncertainty}, pages 351--367.
-\newblock Springer International Publishing, Cham, 2016.
-
-\bibitem{stoller2011runtime}
-S.~D. Stoller, E.~Bartocci, J.~Seyster, R.~Grosu, K.~Havelund, S.~A. Smolka,
-  and E.~Zadok.
-\newblock Runtime verification with state estimation.
-\newblock In {\em Runtime Verification}, pages 193--207. Springer, 2011.
-
-\bibitem{chromecast}
-Wikipedia.
-\newblock {Chromecast}.
-\newblock \url{https://en.wikipedia.org/wiki/Chromecast}.
-
-\bibitem{xbox}
-Wikipedia.
-\newblock {Xbox One} controller.
-\newblock \url{https://en.wikipedia.org/wiki/Xbox_One_Controller}.
-
-\end{thebibliography}
-
-}
-
-\end{document}

### 2017-07-09 submitted/all-in-one.tex
comments

@@
 mutated traces, each satisfying the original state machine with a specific
-likelihood. If the set is empty, the implementation definitely violates the
+likelihood. If the set of mutated traces is empty, the implementation definitely violates the
 protocol. Searching over all the mutated traces is NP-complete, but the approach
@@
 A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
-the state machine at time $t_i$. The monitor {\it rejects a trace} $Tr$ if there
-exists a prefix of $Tr$ such that all states reachable after consuming the
-prefix have no valid transitions for the next $(t, p)$ input.
-
-As an example, the monitor state machine illustrated in
-Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:
+the state machine at time $t_i$.  As an example, the monitor state machine
+illustrated in Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:
 
@@
 
+The DUT and endpoint are configured to use the IEEE 802.11g standard with both
+RTS/CTS and fragmentation disabled.  A Constant Bit Rate (CBR) UDP traffic
+(54~Mbps) is generated from the DUT to the endpoint.  The UDP packet size is
+1436 bytes, which results in a 1500 bytes \wifi{} packet.
+
+In order to control the packet loss ratios between each pair of devices, we
+developed a new propagation loss model for \ns{} called
+\texttt{MatrixRandomPropagationLossModel}.
+\sloppy{  Instead of a constant propagation
+  loss as in existing \texttt{MatrixPropagationLossModel}, the signal
+  propagation loss between a pair of nodes is determined by a binary random
+  variable of two  values: 0~dB (no loss) and 1000~dB (complete loss).
+  }
+Therefore, any packet loss
+probability can be achieved by adjusting the random variable distribution.
+The
+model supports both symmetric and asymmetric propagation losses.
+We use
+symmetric propagation loss in all our experiments.
+Finally, pcap capture is enabled in both the DUT and the sniffer devices.
+
 \subsubsection{Verifying Unmodified Implementation}
@@
 
+We report a bug found in NS-3 ARF~\cite{kamerman1997wavelan} implementation
+which causes the sender to get stuck at a lower rate even after enough number of
+consecutive successes. The bug was detected using sniffer traces and
+confirmed by both the DUT trace and source code inspection.
+
 Automatic Rate Fallback (ARF)~\cite{kamerman1997wavelan} is the first rate
@@
 
-We report a bug found in NS-3 ARF~\cite{kamerman1997wavelan} implementation
-which causes the sender to get stuck at a lower rate even after enough number of
-consecutive successes. The bug was detected using sniffer traces and
-confirmed by both the DUT trace and source code inspection.
-
 \makeatletter{}\subsection{Industrial Application}
@@
 
-We split the traces into \num{100000} packet segments, and applied our framework
-to validate the DUT implementation. We found that the latest implementation of
-the protocol under development had violations of the protocol specification.
-Some of the implementation bugs we found related to the sequence number
-management, station scheduling during power saving mode, and modulation rates
-adaptation.
-
-Table~\ref{tab:xbox} summarizes the validation result. Note that if we disable
+We first split the traces into \num{100000} packet segments, which yields \num{3127}
+traces for testing.  We then applied our framework on the traces to validate the
+DUT implementation. We found that the latest implementation of the protocol
+under development had violations of the protocol specification.  Some of the
+implementation bugs we found related to the sequence number management, station
+scheduling during power saving mode, and modulation rates adaptation.
+Table~\ref{tab:xbox} summarizes the validation results. The \textbf{Traces}
+column shows the number of traces that we have successfully validated for each
+category, and the third column shows the fraction of traces that contain at
+least one violation in that category.
+
+Note that if we disable
 the augmented transitions, each trace will be flagged as violation because of
@@
 obtained because of the lack of automatized validating tools. Finally, the bug
-related to rate control has been filed as a bug report. All reported bugs were
-fixed before the next release the product.
+related to rate control, which was unknown previously, has been filed as a bug
+report. All reported bugs were fixed before the next release the product.
 
@@
 sampling does not lead to spurious alarms.
-\end{comment}
 
@@
 
-We formally define the uncertainty problem in validating wireless protocol
+We formally define an instance of the uncertainty problem in validating wireless protocol
 implementations using sniffers. We describe a systematic augmentation of the
@@
 
+\bibitem{bornholt2014uncertain}
+J.~Bornholt, T.~Mytkowicz, and K.~S. McKinley.
+\newblock Uncertain< t>: A first-order type for uncertain data.
+\newblock {\em ACM SIGARCH Computer Architecture News}, 42(1):51--66, 2014.
+
 \bibitem{cheng2006jigsaw}
@@
 
+\bibitem{Elbaum:2014:KUT:2635868.2666608}
+S.~Elbaum and D.~S. Rosenblum.
+\newblock Known unknowns: Testing in the presence of uncertainty.
+\newblock In {\em Proceedings of the 22Nd ACM SIGSOFT International Symposium
+  on Foundations of Software Engineering}, FSE 2014, pages 833--836, New York,
+  NY, USA, 2014. ACM.
+
 \bibitem{fei2006artemis}
@@
 
+\bibitem{marino2009literace}
+D.~Marino, M.~Musuvathi, and S.~Narayanasamy.
+\newblock Literace: effective sampling for lightweight data-race detection.
+\newblock In {\em ACM Sigplan Notices}, volume~44, pages 134--143. ACM, 2009.
+
 \bibitem{musuvathi2002cmc}
@@
 
+\bibitem{sampson2014expressing}
+A.~Sampson, P.~Panchekha, T.~Mytkowicz, K.~S. McKinley, D.~Grossman, and
+  L.~Ceze.
+\newblock Expressing and verifying probabilistic assertions.
+\newblock In {\em ACM SIGPLAN Notices}, volume~49, pages 112--122. ACM, 2014.
+
 \bibitem{omnipeek}

### 2017-07-09 submitted/response.txt
add response

First, we would like to thank all reviewers for their insightful comments and
suggestions. Besides fixing typos and minor cosmetic issues, we also tried our
best to address the comments. Responses to each individual reviewers are listed
below.

To reviewer 1:

With respect to the edit distance, that's precisely the idea behind augmented
transitions. Type-1 transitions *inserts* packets to the sniffer trace while
Type-2 transitions *removes* packets from the sniffer trace. The number of
augmented transitions (Type-1 and Type-2) represents the "distance" between the
mutation trace and the original sniffer trace. The gist of the NumMissing
heuristic we proposed is exactly to limit the "distance" (hence edits) to the
sniffer trace.  In other words, the "distance" between the mutation trace and
sniffer trace represents the confidence of protocol correctness.

To reviewer 2:

With respect to the verification coverage future work, the current proposal
provides ways to analyze sniffer traces that are already being collected. While
this is useful in revisiting existing traces which are mostly likely only spot
checked, the verification is still limited to a static sets of sniffer trace
which may only cover the "normal" path of a protocol excitation. Consider the
case study of the game controller protocol, the rate control related violations
were only detected in a tiny fraction of the traces since the violation requires
non-trivial packet loss patterns to manifest itself. The idea behind our
verification coverage proposal is to proactively control the process of
experiment set up and sniffer trace collection to increase the chance of each
protocol state being visited, hence increasing the  probability of violation
detection.

To reviewer 3:

With respect to the fault model, in this paper we only consider full packet
loss, which reflects most practical sniffer's behavior. In most wireless
communications protocols, per-packet checksum is used and a packet that fails
the checksum will be dropped by the network driver.

With respect to the detailed NS-3 experiment setup, we added a paragraph in
Section 5.1.1 to explain the packet loss model we used in the experiments.
Basically, we added a new propagation loss component to NS-3 so that we can
control per-packet losses between each individual pair of devices.

### 2017-11-01 related.tex
working on related

@@
 runtime
-verification~\cite{bonakdarpour2011sampling,hauswirth2004low,arnold2008qvm,fei2006artemis,basin2012monitoring}.
-Stoller \textit{et al}~\cite{stoller2011runtime} used HMM-based state estimation
-techniques to calculate the confidence that the temporal property is satisfied
-in the presence of gaps in observation.
+verification~\cite{bonakdarpour2011sampling,hauswirth2004low,arnold2008qvm,fei2006artemis,basin2012monitoring,sistla2011runtime}.
+In particular, Stoller \textit{et al}~\cite{stoller2011runtime} used HMM-based
+state estimation techniques to calculate the confidence that the temporal
+property is satisfied in the presence of gaps in observation. HMM-based
+approaches proposed in~\cite{stoller2011runtime} suffer from overhead of both
+running time and memory consumption. Optimizations 
+such as approximate precomputed tables~\cite{bartocci2012adaptive} and particle
+filtering~\cite{kalajdzic2013runtime} were later proposed.
 
-While it seems possible to adapt the
-method in~\cite{stoller2011runtime} to our problem, we note several advantages
-of our approach. First, the
-automatically augmented state machine precisely encodes the protocol
-specification and the uncertainty. This is intuitive to design and natural for
+In our problem, there is not only gaps in the observation (packets missed by the
+sniffer), but also uncertainty regarding observed events, which lies in whether
+a packet in the sniffer trace was received by its destination device. This
+unique challenge makes it infeasible to directly apply the HMM-based approaches
+proposed in~\cite{stoller2011runtime}.
+Besides, there are several advantages of our proposed augmented transition
+approach. First, the automatically augmented state machine precisely encodes the
+protocol specification and the uncertainty. This is intuitive to design and
+natural for
 reporting the evidence for a trace being successful. We do not require a user
@@
 exhaustive search.
+
+\textbf{Edit Distance.} The closet work to ours is the Edit distance based approach
+to monitor STL properties of real time systems~\cite{jakvsic2016quantitative}.
 

### 2017-11-14 related.tex
related

@@
 
-In our problem, there is not only gaps in the observation (packets missed by the
+In our problem, there are not only gaps in the observation (packets missed by the
 sniffer), but also uncertainty regarding observed events, which lies in whether
@@
 
-\textbf{Edit Distance.} The closet work to ours is the Edit distance based approach
-to monitor STL properties of real time systems~\cite{jakvsic2016quantitative}.
+\textbf{Edit Distance.} The closet work to ours
+is~\cite{jakvsic2016quantitative}, which uses the weighted edit distance to
+measure the robustness between digitized Cyber-Physical System (CPS) signals and
+STL specifications. The automaton of verifying STL properties is first
+translated into weighted edit automaton, which explicitly handles substitution,
+insertion and deletion by augmenting the original automaton with transitions and
+associating to them the appropriate weight function. An observed signal sequence
+is then input into the weighted edit automaton to compute the weighted edit
+distance between the observation and the specification.
+
+We share the same spirit with~\cite{jakvsic2016quantitative} in that we also
+explicitly handle the sniffer trace uncertainty using the augmented transitions.
+However, there is no need for substitution in our application scenario, as the
+sniffer can only either receive or miss a packet, but can not mis-interpret
+packets. In addition, ~\cite{jakvsic2016quantitative} only considers real value
+observations, and relies on this restriction to define the edit distance for
+insertion transitions. The approach can not be directly applied in our case
+since we need to deal with compound observations (packets).
 

### 2017-11-14 framework/heuristic.tex
all-in-one
work

@@
 The motivation is that many protocols operate as a sequence of independent
-transactions, and the uncertainty of previous transactions often do not affect
+transactions, and the uncertainty of previous transactions often does not affect
 the next transaction.
 For instance, in 802.11 packet transmission protocol, each packet exchange,
-include the original, retransmission and acknowledgment packets, constitute a
+include the original, retransmission and acknowledgment packets, constitutes a
 transaction.
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{./figures/loss.pdf}
+  \includegraphics[width=0.4\textwidth]{loss.pdf}
   \caption{\textbf{Example Sniffer Trace and Its Mutation Trace.}}

### 2017-11-14 introduction.tex
comments
work

@@
 
-We present a new technique than enables validation of protocol implementations using
+We present a new technique that enables validation of protocol implementations using
 wireless sniffers. Given a monitor state machine representing the protocol being
